### 2015-09-09 paper.tex
starting pieces

\documentclass{sigchi}

% Use this command to override the default ACM copyright statement
% (e.g. for preprints).  Consult the conference website for the
% camera-ready copyright statement.

%% EXAMPLE BEGIN -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)
% \toappear{Permission to make digital or hard copies of all or part of this work for personal or classroom use is      granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than ACM must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org. \\
% {\emph{CHI'14}}, April 26--May 1, 2014, Toronto, Canada. \\
% Copyright \copyright~2014 ACM ISBN/14/04...\$15.00. \\
% DOI string from ACM form confirmation}
%% EXAMPLE END -- HOW TO OVERRIDE THE DEFAULT COPYRIGHT STRIP -- (July 22, 2013 - Paul Baumann)

% Arabic page numbers for submission.  Remove this line to eliminate
% page numbers for the camera ready copy

%\pagenumbering{arabic}

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}

% llt: Define a global style for URLs, rather that the default one
\makeatletter
\def\url@leostyle{%
  \@ifundefined{selectfont}{\def\UrlFont{\sf}}{\def\UrlFont{\small\bf\ttfamily}}}
\makeatother
\urlstyle{leo}

% To make various LaTeX processors do the right thing with page size.
\def\pprw{8.5in}
\def\pprh{11in}
\special{papersize=\pprw,\pprh}
\setlength{\paperwidth}{\pprw}
\setlength{\paperheight}{\pprh}
\setlength{\pdfpagewidth}{\pprw}
\setlength{\pdfpageheight}{\pprh}

% Make sure hyperref comes last of your loaded packages, to give it a
% fighting chance of not being over-written, since its job is to
% redefine many LaTeX commands.
\definecolor{linkColor}{RGB}{6,125,233}
\hypersetup{%
  pdftitle={SIGCHI Conference Proceedings Format},
  pdfauthor={LaTeX},
  pdfkeywords={SIGCHI, proceedings, archival format},
  bookmarksnumbered,
  pdfstartview={FitH},
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=black,
  urlcolor=linkColor,
  breaklinks=true,
}

% create a shortcut to typeset table headings
% \newcommand\tabhead[1]{\small\textbf{#1}}

% End of preamble. Here it comes the document.
\begin{document}

\title{SIGCHI Conference Proceedings Format}

\numberofauthors{3}
\author{%
  \alignauthor{1st Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{2nd Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
  \alignauthor{3rd Author Name\\
    \affaddr{Affiliation}\\
    \affaddr{City, Country}\\
    \email{e-mail address}}\\
}

\maketitle

\begin{abstract}
  UPDATED---\today. This sample paper describes the
  formatting requirements for SIGCHI conference proceedings, and
  offers recommendations on writing for the worldwide SIGCHI
  readership. Please review this document even if you have submitted
  to SIGCHI conferences before, as some format details have changed
  relative to previous years. Abstracts should be about 150 words and
  are required.
\end{abstract}

\keywords{Authors' choice; of terms; separated; by semi\-colons;
  commas, within terms only; this section is required.}

\category{H.5.m.}{Information Interfaces and Presentation
  (e.g. HCI)}{Miscellaneous} \category{See
  \url{http://acm.org/about/class/1998/} for the full list of ACM
  classifiers. This section is required.}{}{}

\section{Introduction}

Developing good user interfaces for web or mobile applications is a
complex and expensive endeavor. One reason is the combination of
devices, interaction modalities and workflows that need to be
supported.

Adopting Usage Centered Development practices is effective, as is
following established design principles~\cite{constantine99}. Early
prototyping~\cite{buxton07} and exploration of the 5-dimensional prototyping
space~\cite{mccurdy06} is one of the most effective techniques,
especially when paired with usability investigations based on user
testing or heuristic evaluations.

However, it requires development of prototypes that are usually
developed with certain tasks in mind, and therefore are quite
restricted in terms of depth and breadth of supported use
cases. Furthermore, usability results are always surrounded by a cloud
of uncertainty, due to subjectivity introduced by participants and
facilitators or by other contingency factors involved in the
analysis. Thus, although an effort needs to be expended to develop and
use prototypes, less than optimal results are obtained.

A designer, while conceiving and developing a solution, might benefit
from answers to seemingly simple questions that should not require a
significant investment of time and effort. Given one or more potential
solutions and some usage scenarios, interesting questions could
include: ``What are the optimal execution paths?'' or ``How many such
paths exist?'', or ``Are usage mistakes recoverable?''. At the moment,
even these straightforward questions are quite complex to answer. In
fact, they require inspection of prototypes, manual tracking of which
screens and widgets are used at which stage, and exhaustive searches.

This should not be the case, however. These answers provide important
insights to a designer, and support decisions related to benchmarking
different solutions, to identification of optimal or mistaken paths,
to assessment of suitability of a design with respect to scenarios.

In this paper we present an approach that allows designer to quickly
be able to get these kinds of answers, and present it on a case study
comparing four different web mail front ends. We have chosen this
domain because it is very well understood by readers and yet those
trivial questions lead to non trivial results.

The approach is based on UML statechart models of the user interfaces
which are automatically processed to produce \emph{interaction
  graphs}. These are then used to unfold \emph{execution traces} that
are dependent on the specific usage scenarios being considered in the
analysis. On traces several graph-theoretic computations can be
performed to produce a dhasboard of different results that provide the
answers. Except for development of models and specification of the
desired scenarios, the other steps are totally automatic; models of a
design could be developed in a matter of a couple of hours.

The contribution consists of the development of a tool that transforms
models and scenario specifications into execution traces, and the
definition of metrics that provide concise, precise and objective
measures of a design.

% REFERENCES FORMAT
% References must be the same font size as other body text.
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{gb,database}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2015-09-09 paper.tex
initial draft of abstract and intro

@@
 
-\title{SIGCHI Conference Proceedings Format}
+\title{To be decided\\
+Measuring Interaction before developing prototypes\\
+Measuring Interaction Graphs}
 
@@
 \begin{abstract}
-  UPDATED---\today. This sample paper describes the
-  formatting requirements for SIGCHI conference proceedings, and
-  offers recommendations on writing for the worldwide SIGCHI
-  readership. Please review this document even if you have submitted
-  to SIGCHI conferences before, as some format details have changed
-  relative to previous years. Abstracts should be about 150 words and
-  are required.
+  Early development of prototypes and testing them with users is a good
+  practice for user interfaces development. However, prototypes have
+  to cover specific usage scenarios, and because they are limited
+  in focus, the whole picture of the user interface is easily lost. Even
+  simple questions dealing with numerosity and length of
+  optimal execution paths or impact of possible user errors can be
+  answered only for the specific scenarios being analyzed.
+
+  We discuss a tool that transforms models of the behavior of a user
+  interface into a graph. This is then used to specify usage
+  scenarios, and to generate possible execution traces. Metrics based
+  on number and length of different possible execution paths, with or
+  without possible mistakes, can be easily computed. We apply the tool
+  and metrics to well known examples of web mail apps, and show that
+  important conclusions can be drawn even before the first prototype
+  is built.
+
 \end{abstract}
@@
 following established design principles~\cite{constantine99}. Early
-prototyping~\cite{buxton07} and exploration of the 5-dimensional prototyping
-space~\cite{mccurdy06} is one of the most effective techniques,
-especially when paired with usability investigations based on user
-testing or heuristic evaluations.
+prototyping~\cite{buxton07} to explore part of the 5-dimensional
+prototyping space~\cite{mccurdy06} is one of the most effective
+techniques, especially when paired with usability investigations based
+on user testing or heuristic evaluations.
 
-However, it requires development of prototypes that are usually
+However, it still requires development of prototypes that are usually
 developed with certain tasks in mind, and therefore are quite

### 2015-09-10 paper.tex
Simon's initial Changes and comments to Abstract and Introduction.

@@
 \usepackage{todonotes}
+\usepackage{soul}
+\usepackage{paralist}
+
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage[super]{nth}
+\usepackage{amsmath}
 
@@
 \begin{abstract}
-  Early development of prototypes and testing them with users is a good
-  practice for user interfaces development. However, prototypes have
+  Early development and testing of prototypes is good
+  practice for user interface development. However, prototypes have
   to cover specific usage scenarios, and because they are limited
@@
   optimal execution paths or impact of possible user errors can be
-  answered only for the specific scenarios being analyzed.
+  answered only for the specific scenarios being analysed.
 
-  We discuss a tool that transforms models of the behavior of a user
+  We discuss a tool that transforms models of the behaviour of a user
   interface into a graph. This is then used to specify usage
@@
   without possible mistakes, can be easily computed. We apply the tool
-  and metrics to well known examples of web mail apps, and show that
+  and metrics to well known examples of web mail apps, \hl{and show that
   important conclusions can be drawn even before the first prototype
-  is built.
+  is built. --- \textbf{replace with a real example statement}}.
 
@@
 
-\keywords{Authors' choice; of terms; separated; by semi\-colons;
-  commas, within terms only; this section is required.}
+\keywords{Experimental; Evaluation; Statecharts: UML; UML-IDEA; Testing.}
 
-\category{H.5.m.}{Information Interfaces and Presentation
-  (e.g. HCI)}{Miscellaneous} \category{See
-  \url{http://acm.org/about/class/1998/} for the full list of ACM
-  classifiers. This section is required.}{}{}
+\category{H.5.1}{Information interfaces and presentation (e.g., HCI)}{Multimedia Information Systems}. 
+\category{H.5.2}{Information interfaces and presentation (e.g.,
+  HCI)}{User Interfaces}.
 
 \section{Introduction}
+We present an approach that allows a designer to quickly 
+\begin{inparaenum}[(a)]
+\item understand how prone an application is to user navigation errors;  
+\item understand how recoverable the application is from those errors;
+\item support task analysis and task / scenario design;
+\item support interaction design (IxD) in terms of consistency, error-proneness, and recovery premature commitment;
+\end{inparaenum}
+and further, allows different designs to be objectively compared to support designers in evaluating interaction sequences. To explain our work we present a case study
+comparing four different web mail front ends. We have chosen this
+domain because it is very well understood by readers and yet those
+trivial questions lead to non trivial results.
+
+The approach is based on UML state-chart models of the user interfaces
+which are automatically processed to produce \emph{interaction graphs}. These are then used to unfold \emph{execution traces} that
+are dependent on the specific usage scenarios being considered in the
+analysis. On traces several graph-theoretic computations can be
+performed to produce a dashboard of different results that provide the
+answers. Except for development of models and specification of the
+desired scenarios, the other steps are totally automatic; models of a
+design could be developed in a matter of a couple of hours.
 
 Developing good user interfaces for web or mobile applications is a
-complex and expensive endeavor. One reason is the combination of
+complex and expensive endeavour. One reason is the combination of
 devices, interaction modalities and workflows that need to be
@@
 
-Adopting Usage Centered Development practices is effective, as is
+Adopting Usage Centred Development practices is effective, as is
 following established design principles~\cite{constantine99}. Early
-prototyping~\cite{buxton07} to explore part of the 5-dimensional
+prototyping~\cite{buxton07} to explore part of the five-dimensional
 prototyping space~\cite{mccurdy06} is one of the most effective
@@
 solutions and some usage scenarios, interesting questions could
-include: ``What are the optimal execution paths?'' or ``How many such
-paths exist?'', or ``Are usage mistakes recoverable?''. At the moment,
+include: \hl{``What are the optimal execution paths?'' or ``How many such
+paths exist?'', or ``Are usage mistakes recoverable?'' -- \textbf{these seem to be software `executionist' and not so much user/designer is -- maybe we can say something more designer and save these to the technical bit of the paper?}}.
+At the moment,
 even these straightforward questions are quite complex to answer. In
@@
 to assessment of suitability of a design with respect to scenarios.
-
-In this paper we present an approach that allows designer to quickly
-be able to get these kinds of answers, and present it on a case study
-comparing four different web mail front ends. We have chosen this
-domain because it is very well understood by readers and yet those
-trivial questions lead to non trivial results.
-
-The approach is based on UML statechart models of the user interfaces
-which are automatically processed to produce \emph{interaction
-  graphs}. These are then used to unfold \emph{execution traces} that
-are dependent on the specific usage scenarios being considered in the
-analysis. On traces several graph-theoretic computations can be
-performed to produce a dhasboard of different results that provide the
-answers. Except for development of models and specification of the
-desired scenarios, the other steps are totally automatic; models of a
-design could be developed in a matter of a couple of hours.
 

### 2015-09-11 paper.tex
revised abstract and intro; added figures.

@@
 
-\title{To be decided\\
+\title{Measuring Interaction Design\\
 Measuring Interaction before developing prototypes\\
@@
 
-\numberofauthors{3}
+\numberofauthors{2}
 \author{%
@@
     \email{e-mail address}}\\
-  \alignauthor{3rd Author Name\\
-    \affaddr{Affiliation}\\
-    \affaddr{City, Country}\\
-    \email{e-mail address}}\\
 }
@@
 \begin{abstract}
-  Early development and testing of prototypes is good
-  practice for user interface development. However, prototypes have
-  to cover specific usage scenarios, and because they are limited
-  in focus, the whole picture of the user interface is easily lost. Even
-  simple questions dealing with numerosity and length of
-  optimal execution paths or impact of possible user errors can be
-  answered only for the specific scenarios being analysed.
+  Early development of prototypes is good practice for user interface
+  development. However, they have to cover specific usage scenarios,
+  and because they are limited in focus, the whole picture of the user
+  interface is easily lost. Even simple questions dealing with
+  numerosity and length of execution paths or impact of possible user
+  errors can be answered only for the specific scenarios being
+  analysed.
 
-  We discuss a tool that transforms models of the behaviour of a user
-  interface into a graph. This is then used to specify usage
-  scenarios, and to generate possible execution traces. Metrics based
-  on number and length of different possible execution paths, with or
-  without possible mistakes, can be easily computed. We apply the tool
-  and metrics to well known examples of web mail apps, \hl{and show that
-  important conclusions can be drawn even before the first prototype
-  is built. --- \textbf{replace with a real example statement}}.
+  We discuss a tool that transforms models of a user interface into a
+  graph. This is then used to specify usage scenarios, and to generate
+  possible execution traces. Metrics based on different possible
+  execution paths, with or without possible mistakes, can be easily
+  computed. When applying these metrics to Gmail and Roundcube in a
+  typical scenario, we learn for example that Gmail has 7 times more
+  optimal paths, it has 4x more paths include at most one possible
+  user error, that it has 20\% fewer steps.
 
@@
 \section{Introduction}
+
+\begin{quote}
+  Subcommittee: Technology, Systems, and Engineering
+
+People we should pay particular attention to:
+Caroline Appert
+Conversy, Stéphane
+Nebeling, Michael
+
+Keyword we need to pay attention to:
+This subcommittee will focus on technology, systems and engineering
+contributions that enable, improve, or advance interaction. This will
+include software and hardware technologies and systems that enable and
+demonstrate novel interactive capabilities, as well as languages,
+methods and tools for construction and engineering of interactive
+systems. Engineering contributions should clearly demonstrate how they
+address interactive systems concerns such as, for example,
+scalability, reliability, interoperability, testing, and
+performance. Systems and technology contributions will be judged by
+their technical innovation and/or ability to connect, simplify or
+enrich interactions, for example in intelligent interfaces and
+mobile/ubiquitous computing.
+
+\end{quote}
+
 We present an approach that allows a designer to quickly 
 \begin{inparaenum}[(a)]
-\item understand how prone an application is to user navigation errors;  
+\item understand how supportive an application is with respect to  user efficiency;
+\item understand how prone the application is to user navigation
+  errors;
 \item understand how recoverable the application is from those errors;
 \item support task analysis and task / scenario design;
-\item support interaction design (IxD) in terms of consistency, error-proneness, and recovery premature commitment;
-\end{inparaenum}
-and further, allows different designs to be objectively compared to support designers in evaluating interaction sequences. To explain our work we present a case study
-comparing four different web mail front ends. We have chosen this
-domain because it is very well understood by readers and yet those
-trivial questions lead to non trivial results.
+%\item support interaction design (IxD) in terms of consistency, error-proneness, and recovery premature commitment;
+\end{inparaenum}.  
+%
+Our approach further allows different designs to be
+objectively compared to support designers in evaluating interaction
+sequences. To explain our work we present a case study comparing four
+different web mail front ends. We have chosen this domain because it
+is very well understood by readers and yet those trivial questions
+lead to non trivial results.
 
 The approach is based on UML state-chart models of the user interfaces
-which are automatically processed to produce \emph{interaction graphs}. These are then used to unfold \emph{execution traces} that
+which are automatically processed to produce \emph{interaction
+  graphs}. These are then used to unfold \emph{execution traces} that
 are dependent on the specific usage scenarios being considered in the
@@
 from answers to seemingly simple questions that should not require a
-significant investment of time and effort. Given one or more potential
+significant investment of time and effort. For example, given one or more potential
 solutions and some usage scenarios, interesting questions could
-include: \hl{``What are the optimal execution paths?'' or ``How many such
-paths exist?'', or ``Are usage mistakes recoverable?'' -- \textbf{these seem to be software `executionist' and not so much user/designer is -- maybe we can say something more designer and save these to the technical bit of the paper?}}.
-At the moment,
-even these straightforward questions are quite complex to answer. In
-fact, they require inspection of prototypes, manual tracking of which
-screens and widgets are used at which stage, and exhaustive searches.
+include: ``How many different ways can be followed to carry out the
+scenario?'', ``Which are the shortest ones?'', ``If a user makes a
+mistake, would he or she be able to recover?'', ``How many steps
+would the recovery require?''. For example, when designing and
+evaluating embedded user interfaces (such as when dealing with plane's
+cockpits~\cite{conversy07}), other relevant questions might include
+``How would the above properties change if we add a certain a
+widget?'', or ``... if we replace a widget with another?''.  At the
+moment, even these straightforward questions are quite complex to
+answer. In fact, they require inspection of prototypes, manual
+tracking of which screens and widgets are used at which stage, and
+exhaustive searches.
 
@@
 
-The contribution consists of the development of a tool that transforms
+Our contribution consists of the development of a tool that transforms
 models and scenario specifications into execution traces, and the
 definition of metrics that provide concise, precise and objective
-measures of a design.
+measures of a design. The case study we illustrate shows that among
+four web mail applications, and with respect to a  typical usage
+scenario, Gmail has the largest number of shortest sequences of steps
+(even when users are supposed to make 1 or 2 mistakes), but when users
+make more than 2 mistakes the number drops significantly; on average,
+best and worst cases, Gmail features also shorter paths, requiring 20\%
+fewer steps; however, the probability that a user hits an optimal path
+with Gmail is almost half of that of another application. All these
+values suggest that Gmail offers more  efficient options to
+accomplish tasks included in the scenario, but that it might be more
+difficult for novice users to exploit the most efficient
+methods. Further inspections show that some differences are due to the
+slightly different interaction structures adopted for uploading messages.
+
+\section{Background}
+\label{sec:background}
+
+(right now these are random notes)
+
+SwingState: work done by 
+Caroline Appert\cite{appert2008swingstates}
+
+in java, FSA used as a conceptual framework to write the code of
+widgets so that events and event handlers in the ui can more easily be
+conceived, developed and verified.
+
+They say: StateCharts were used in the StateMaster User Interface
+Management System [40], and a variant of them specifically tailored to
+designing user interfaces was used in the more recent HsmTk toolkit
+[18]. StateCharts however are significantly more complicated and hard
+to learn than plain state machines, and our experience is that user
+interface designers and developers have difficulties exploiting their
+power. Other approaches include Petri Nets [41], which have also been
+used to specify user interfaces, for example in the PetShop system
+[42]. Here too, the learning curve is steep, making the adoption of
+such a model by developers difficult.
+
+With SwingStates, any number of state machines can run simulateneously. A state machine
+can be active, i.e., handling the events it receives, or inactive,
+i.e., ignoring events
+
+We adopt a similar apporach, but using UML state machine (which are
+more powerful than FSA) for conceiving, guiding development, analysis
+and verification of the behavior of the entire app.
+
+Nebeling says:\cite{nebeling2011metrics}
+
+Despite the fact that screen sizes and average screen resolutions have
+dramatically increased over the past few years, little attention has
+been paid to the design of web sites for large, high-resolution
+displays that are now becoming increasingly used both in enterprise
+and consumer spaces. We present a study of how the visual area of the
+browser window is currently utilised by news web sites at different
+widescreen resolutions. The analysis includes measurements of space
+taken up by the article content, embedded ads and the remaining
+components as they appear in the viewport of the web browser. The
+results show that the spatial distribution of page elements does not
+scale well with larger viewing sizes, which leads to an increasing
+amount of unused screen real estate and unnecessary scrolling. We
+derive a number of device-sensitive metrics to measure the quality of
+web page layout in different viewing contexts, which can guide the
+design of flexible layout templates that scale effectively on large
+screens.
+
+Stephane Conversy\cite{conversy07} says: (friend of nicolas roussel in
+the committee)
+
+ARINC 661 provides precise information for communication protocol
+between application (called User Applications) and user interface
+components (called widgets) as well as precise information about the
+widgets themselves. However, in ARINC 661, no information is given
+about the behaviour of these widgets and about the behaviour of an
+application made up of a set of such widgets.
+
+The purpose of ARINC 661 specification (ARINC 661, 2002) is to define
+interfaces to a Cockpit Display System (CDS) used in interactive
+cockpits that are now under
+
+application of a formal description technique to the various elements
+of ARINC 661 specification within an industrial project. This formal
+description technique called Interactive Cooperative Objects defines
+in a precise and non-ambiguous way all the elements of ARINC 661
+specification. The application of the formal description techniques is
+shown on an interactive application called MPIA (Multi Purpose
+Interactive Application). Within this application, we present how ICO
+are used for describing interactive widgets, User Applications and
+User Interface servers (in charge of interaction techniques). The
+emphasis is put on the model-based management of the feel of the
+applications allowing rapid prototyping of the external presentation
+and the interaction techniques.
+
+The Interactive Cooperative Objects (ICOs) formalism is a formal
+description technique dedicated to the specification of interactive
+systems [4, 11]. It uses concepts borrowed from the object-oriented
+approach to describe the structural or static aspects of systems, and
+uses high-level Petri nets [8] to describe their dynamic or
+behavioural aspects.
+
+The Interactive Cooperative Objects (ICOs) formalism is a formal
+description technique dedicated to the specification of interactive
+systems [4, 11]. It uses concepts borrowed from the object-oriented
+approach to describe the structural or static aspects of systems, and
+uses high-level Petri nets [8] to describe their dynamic or
+behavioural aspects.
+
+Voida\cite{voida} says:
+We describe an alternative model for organizing the
+desktop interface-activity-based computing-and
+identify a series of high-level system requirements for
+interfaces that use activity as their primary organizing
+principle.
+
+We could mention this work when we introduce scenarios, and say that
+people inventing these new metaphors could take advantage of our
+metrics if they model 2+ apps.
+
+Eg. they say: Requirement 2. Activity-based systems should provide
+lightweight mechanisms to create, change, and alter
+activities, since heavyweight interaction techniques are
+likely to deter adoption and use.
+
+eg Requirement 7. Because information sharing is a
+“common case” in knowledge work, lightweight sharing
+capabilities should be integrated directly as a first-class
+interaction technique.
+
+\section{Generation of execution traces}
+\label{sec:traces}
+
+\begin{quote}
+  here we describe how we transform state machines into graphs, how we
+  specify scenarios, how the system computes traces, how metrics are
+  computed. 
+\end{quote}
+
+\section{A case study}
+\label{sec:casestudy}
+
+\subsection{Models and scenarios}
+\label{sec:models-scenarios}
+
+\subsection{Results}
+\label{sec:results}
+
+\clearpage
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/num-paths.png}
+  \caption{Number of geodesic paths split by detour order.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/averagecase.png}
+  \caption{Average length of  geodesic paths split by detour order.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/bestcase.png}
+  \caption{Minimum length of geodesic paths split by detour order.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/worstcase.png}
+  \caption{Maximum length of geodesic paths split by detour order.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/prob-opt-path.png}
+  \caption{Frequency of an optimal path.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/pr.png}
+  \caption{Probability that a random walk visit detour 0, 1, 2 or 3 states.}
+\end{figure}
+
+\section{Discussion}
+\label{sec:discussion}
+
+\section{Conclusion}
+\label{sec:conclusion}
 

### 2015-09-11 paper.tex
details

@@
 \item understand how recoverable the application is from those errors;
-\item support task analysis and task / scenario design;
+\item support task analysis and task / scenario design.
 %\item support interaction design (IxD) in terms of consistency, error-proneness, and recovery premature commitment;
-\end{inparaenum}.  
+\end{inparaenum}
 %
-Our approach further allows different designs to be
-objectively compared to support designers in evaluating interaction
-sequences. To explain our work we present a case study comparing four
-different web mail front ends. We have chosen this domain because it
-is very well understood by readers and yet those trivial questions
-lead to non trivial results.
+Our approach further allows different designs to be objectively
+compared to support designers in evaluating interaction sequences. To
+explain our work we present (\S~\nameref{sec:casestudy}) a case study
+comparing four different web mail front ends. We have chosen this
+domain because it is very well understood by readers and yet those
+trivial questions lead to non trivial results.
 
@@
 are dependent on the specific usage scenarios being considered in the
-analysis. On traces several graph-theoretic computations can be
-performed to produce a dashboard of different results that provide the
-answers. Except for development of models and specification of the
-desired scenarios, the other steps are totally automatic; models of a
-design could be developed in a matter of a couple of hours.
+analysis (\S~\nameref{sec:traces}). On traces several graph-theoretic
+computations can be performed to produce a dashboard of different
+results that provide the answers. Except for development of models and
+specification of the desired scenarios, the other steps are totally
+automatic; models of a design could be developed in a matter of a
+couple of hours.
 

### 2015-09-19 paper.tex
completed draft of examples

--- 2015-09-11 paper.tex
@@
 
-%\pagenumbering{arabic}
+\pagenumbering{arabic}
 
@@
 \usepackage{paralist}
+\usepackage{listings}
 
@@
 \begin{abstract}
-  Early development of prototypes is good practice for user interface
-  development. However, they have to cover specific usage scenarios,
-  and because they are limited in focus, the whole picture of the user
-  interface is easily lost. Even simple questions dealing with
-  numerosity and length of execution paths or impact of possible user
-  errors can be answered only for the specific scenarios being
-  analysed.
+  Early development of prototypes of user interfaces is good
+  practice. However, they have to cover specific usage scenarios, and
+  because they are limited in focus, the whole picture of the user
+  interface is easily lost. Even simple questions dealing with number
+  and length of interaction paths or impact of possible user errors
+  can be answered only for the specific scenarios being analysed.
 
-  We discuss a tool that transforms models of a user interface into a
+  We present a tool that transforms models of a user interface into a
   graph. This is then used to specify usage scenarios, and to generate
-  possible execution traces. Metrics based on different possible
-  execution paths, with or without possible mistakes, can be easily
-  computed. When applying these metrics to Gmail and Roundcube in a
-  typical scenario, we learn for example that Gmail has 7 times more
+  possible interaction paths. Metrics based on possible paths, with or
+  without possible mistakes, can be easily computed. When applying
+  these metrics get precise and objective values that can be used to
+  compare designs or to identify weakness. When applying them to mail
+  applications, we learn for example that Gmail has \hl{7 times more
   optimal paths, it has 4x more paths include at most one possible
-  user error, that it has 20\% fewer steps.
+  user error, that it has 20\% fewer steps.}
 
@@
 demonstrate novel interactive capabilities, as well as languages,
-methods and tools for construction and engineering of interactive
+methods and tools for \hl{CONSTRUCTION AND ENGINEERING} of interactive
 systems. Engineering contributions should clearly demonstrate how they
@@
 performance. Systems and technology contributions will be judged by
-their technical innovation and/or ability to connect, simplify or
+their technical innovation and/or ability to connect, SIMPLIFY or
 enrich interactions, for example in intelligent interfaces and
@@
 
-We present an approach that allows a designer to quickly 
+We present an approach that allows a designer to assess interaction
+design (IxD) qualities such as efficiency, consistency,
+error-proneness and recovery from premature committment. Key
+importance is given to the ability to quickly
 \begin{inparaenum}[(a)]
-\item understand how supportive an application is with respect to  user efficiency;
-\item understand how prone the application is to user navigation
+\item understand how supportive a user interface (UI) is with respect
+  to user efficiency;
+\item understand how prone the UI is to user navigation
   errors;
-\item understand how recoverable the application is from those errors;
-\item support task analysis and task / scenario design;
-%\item support interaction design (IxD) in terms of consistency, error-proneness, and recovery premature commitment;
-\end{inparaenum}.  
+\item understand how recoverable the UI is from those errors; and,
+\item support task analysis and task/scenario design.
+\end{inparaenum}
 %
-Our approach further allows different designs to be
-objectively compared to support designers in evaluating interaction
-sequences. To explain our work we present a case study comparing four
-different web mail front ends. We have chosen this domain because it
-is very well understood by readers and yet those trivial questions
-lead to non trivial results.
+This allows different designs to be objectively compared to support
+designers in the construction and engineering of interactive
+systems. To explain our work we present a comparison of four different
+web mail front ends. We have chosen this domain because it is very
+well known and yet several trivial questions lead to non trivial
+results. We applied the same techniques also in other domains, such as
+HVAC and other embedded UIs.
 
-The approach is based on UML state-chart models of the user interfaces
-which are automatically processed to produce \emph{interaction
-  graphs}. These are then used to unfold \emph{execution traces} that
-are dependent on the specific usage scenarios being considered in the
-analysis. On traces several graph-theoretic computations can be
-performed to produce a dashboard of different results that provide the
+The approach is based on UML statechart models of UIs which are
+automatically processed to produce \emph{interaction graphs}. These
+are then used to specify interaction scenarios and unfold interaction
+paths (called \emph{execution traces}) determined by the specific
+scenarios being considered. Traces are fed to several graph-theoretic
+computations which produce a dashboard with different results with the
 answers. Except for development of models and specification of the
-desired scenarios, the other steps are totally automatic; models of a
-design could be developed in a matter of a couple of hours.
+desired scenarios, which have to be done manually, the other steps are
+totally automatic; models of a design (such as the ones shown in the
+paper) could be developed in a matter of a couple of hours.
 
-Developing good user interfaces for web or mobile applications is a
+Developing good UIs for web or mobile applications is a
 complex and expensive endeavour. One reason is the combination of
@@
 supported.
+%
+Adopting Usage Centred Development practices is effective, as is
+following established design principles~\cite{constantine99}. In
+particular, one of the most effective technique is early
+prototyping~\cite{buxton07} to explore part of the five-dimensional
+prototyping space~\cite{mccurdy06}. It is particularly effective when
+paired with usability investigations based on user testing or
+heuristic evaluations~\cite{rubin08,preece02}.
 
-Adopting Usage Centred Development practices is effective, as is
-following established design principles~\cite{constantine99}. Early
-prototyping~\cite{buxton07} to explore part of the five-dimensional
-prototyping space~\cite{mccurdy06} is one of the most effective
-techniques, especially when paired with usability investigations based
-on user testing or heuristic evaluations.
+However, it requires prototypes which are usually developed with
+certain tasks in mind, and therefore are quite restricted in terms of
+depth, breadth, dynamics and data. Furthermore, in addition to the
+possible bias introduced by prototypes, usability results are always
+surrounded by a cloud of uncertainty, due to subjectivity introduced
+by participants and facilitators or by other contingency factors
+involved in the analysis. Thus, although a  significant effort
+needs to be expended to develop and use prototypes, less than optimal
+results are obtained.
 
-However, it still requires development of prototypes that are usually
-developed with certain tasks in mind, and therefore are quite
-restricted in terms of depth and breadth of supported use
-cases. Furthermore, usability results are always surrounded by a cloud
-of uncertainty, due to subjectivity introduced by participants and
-facilitators or by other contingency factors involved in the
-analysis. Thus, although an effort needs to be expended to develop and
-use prototypes, less than optimal results are obtained.
-
-A designer, while conceiving and developing a solution, might benefit
-from answers to seemingly simple questions that should not require a
-significant investment of time and effort. For example, given one or more potential
-solutions and some usage scenarios, interesting questions could
-include: ``How many different ways can be followed to carry out the
-scenario?'', ``Which are the shortest ones?'', ``If a user makes a
-mistake, would he or she be able to recover?'', ``How many steps
-would the recovery require?''. For example, when designing and
-evaluating embedded user interfaces (such as when dealing with plane's
-cockpits~\cite{conversy07}), other relevant questions might include
-``How would the above properties change if we add a certain a
-widget?'', or ``... if we replace a widget with another?''.  At the
-moment, even these straightforward questions are quite complex to
-answer. In fact, they require inspection of prototypes, manual
-tracking of which screens and widgets are used at which stage, and
-exhaustive searches.
+Even worse, there might be questions whose answer cannot be easily
+found. For example, given one or more potential IxD solutions and some
+usage scenarios, interesting questions could include: ``How many
+different ways can be followed by the user to carry out the scenario?'', ``Which
+are the shortest ones?'', ``If a user makes a mistake, would he or she
+be able to recover?'', ``How many steps would the recovery
+require?''. When designing and evaluating embedded UIs (such as when
+dealing with plane's cockpits~\cite{conversy07}), other relevant
+questions might include ``How would the above properties change if we
+add a certain a widget?'', or ``... if we replace a widget with
+another?''.  At the moment, even these straightforward questions are
+quite complicated to answer. In fact, they require development of
+prototypes, inspecting them, manual tracking of which screens and
+widgets are used at which stage, and a systematic manual analysis.
 
 This should not be the case, however. These answers provide important
-insights to a designer, and support decisions related to benchmarking
-different solutions, to identification of optimal or mistaken paths,
-to assessment of suitability of a design with respect to scenarios.
+insights to a designer, and potentially support assements of potential user
+flexibility, user efficiency, error proneness, ability to recover,
+compactness and consistency of a design.
 
 Our contribution consists of the development of a tool that transforms
-models and scenario specifications into execution traces, and the
+models and scenario specifications into interaction paths, and the
 definition of metrics that provide concise, precise and objective
-measures of a design. The case study we illustrate shows that among
-four web mail applications, and with respect to a  typical usage
-scenario, Gmail has the largest number of shortest sequences of steps
-(even when users are supposed to make 1 or 2 mistakes), but when users
-make more than 2 mistakes the number drops significantly; on average,
-best and worst cases, Gmail features also shorter paths, requiring 20\%
-fewer steps; however, the probability that a user hits an optimal path
-with Gmail is almost half of that of another application. All these
-values suggest that Gmail offers more  efficient options to
-accomplish tasks included in the scenario, but that it might be more
-difficult for novice users to exploit the most efficient
-methods. Further inspections show that some differences are due to the
-slightly different interaction structures adopted for uploading messages.
+measures of a design. The examples included in the paper show that
+among four web mail applications, and with respect to a typical usage
+scenario, Gmail is the most efficient and flexible UI, with the best
+ability to recover from certain mistakes. In fact, Gmail has the
+largest number of shortest sequences of steps (even when users are
+supposed to make 1 or 2 mistakes), but when users make more than 2
+mistakes the number of paths drops significantly; on average, best and worst
+cases, Gmail features also shorter paths, requiring \hl{CHECK 20\% fewer steps};
+however, the probability that a user hits an optimal path with Gmail
+is almost half of that of another application, and the probability
+that a random walker hits a state that is not due to a mistake is
+\hl{SAY SOMETHING}.
+%
+These values suggest that Gmail offers more ways to accomplish tasks
+included in the scenario, that comparably more of these ways do not
+involve extra steps, that they require fewer steps, and that it might
+be more difficult for novice users to exploit the most efficient
+ways. Further inspections show that some differences are due to the
+slightly different interaction structures adopted for uploading
+messages. If Gmail didn't exist yet, its designers could obtain these
+answers well before performing usability studies.
 
@@
 
-\section{Generation of execution traces}
+\section{Generation of interaction paths}
 \label{sec:traces}
@@
 
-\section{A case study}
+Notice that a scenario specifies only the desired occurrences of user
+actions, not the necessary ones. For example, the model specifies that
+in order to view a conversation while reading another one, one has to
+\texttt{goBack} to the inbox first. It is the task of MIGTOOL to
+unfold cyles in the graph and compute all the  paths that connect the
+desired user actions.
+
+Actions specified in a scenario are called \texttt{bridges}, because
+they represent one or more edges in the graph that have to be followed
+when carrying out the
+
+\section{Comparing applications}
 \label{sec:casestudy}
+
+In this section we describe some examples based on well known web mail
+clients, namely Gmail, Horde, SquirrelMail and Roundcube. We chose
+these examples because they are very well known, and therefore are
+easy to describe and understand. And yet, despite email being a very
+well understood domain, the kind of questions that can be posed and
+the answers that are found provide an insight on some of the usability
+properties of these applications.
 
@@
 
-\subsection{Results}
-\label{sec:results}
+Models of the four applications have been manually developed using UML
+state machines. In order to support a fair comparison, all four models
+cover the same set of use cases: listing the content of the inbox,
+reading a message or conversation, replying to a message, composing
+and sending a new message.
 
-\clearpage
+Figure~\ref{fig:model} shows part of the model of Gmail. At some
+point, the user might be viewing all the conversations of the inbox
+(state \texttt{viewingConversations}); available actions include
+moving to the next or previuous block of conversations, refreshing the
+list, or opening a specific conversation (transition
+\texttt{open(conversation)}). This transition (which is assumed to
+occur when the user clicks on one of the visible conversations) leads
+to the state \texttt{viewingAConversation}, where the behavior of the
+system is defined by a more detailed state machine. By default the
+user is viewing a conversation, but by performing the \texttt{reply}
+action the UI moves to a state called \texttt{replying}, where the
+body of the message can be typed, the subject can be changed, or
+another addressee can be added.
+
+Notice that this behavior ``happens'' in one of the two concurrent
+regions specified by this model. In parallel to this, the user can
+either be reading messages (state \texttt{reading}) or may be
+composing a new message (state \texttt{composing}). In the latter case
+the user can independently add recipients, attachments, write the body
+or subject of the message; and send or cancel it.
+
+\begin{figure*}[tbh]
+  \centering
+  \includegraphics[width=\linewidth]{figures/gmail_2.png}
+  \caption{Part of the Gmail model.}\label{fig:model}
+\end{figure*}
+
+The model that we show here is part of what we used in the examples
+reported below, and that is only a part of what the real Gmail
+application supports. The actual model that we used consists of 
+24 states, 12 pseudo states, 12 regions, and 61 transitions.
+
+The flattening process, which takes a fraction of a second on a
+low-cost PC, produces a directed multigraph with 47 vertices and 634
+edges; when simplified by collapsing multiple edges between any pair of
+vertices, the graph includes 312 edges. Each vertex represents one of
+the possible combinations of simple states in any of the regions that
+can be active at the same time.
+
+Similar models and corresponding graphs were produced for the other
+three applications.
+
+\subsection{Analysis of interaction designs}
+\label{sec:analysis}
+
+Inspection of the interaction graph is not particularly useful because
+even for small graphs like the one  obtained from our Gmail model
+no particular structure is evident. Figure~\ref{fig:gom} shows a plot
+using a circular layout of the 47 states: because of the large
+number of cycles that exist among states, edges form an intricated web
+of possible action sequences. This in practice greatly reduces usefulness of
+typical network analysis metrics, such as \emph{betweenness,
+  eccentricity, page-rank, eigenvalue} centrality measures.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{figures/num-paths.png}
-  \caption{Number of geodesic paths split by detour order.}
+  \includegraphics[width=\linewidth]{figures/gom.png}
+  \caption{Graph obtained from the Gmail model.}
+  \label{fig:gom}
 \end{figure}
 
-\begin{figure}
+To be able to obtain results that bear upon usability, we process
+further the graph, first by specifying an interaction scenario using
+vertices and edges of the graph, and secondly by computing possible
+execution paths.
+
+Specification of a scenario consists of formulating it using the
+language provided by the graph, i.e. deciding which is the initial
+state and which user actions need to be considered. For example, the
+following fragment of R code specifies that the scenario should
+include, in the given order, two occurrences of
+\texttt{open(conversation)}, followed by \texttt{reply}, followed by
+\texttt{typeBody}, etc.: \lstset{language=R}
+\begin{lstlisting}
+edgesWithLabel(gmail,"open(conversation)"),
+edgesWithLabel(gmail,"open(conversation)"), 
+edgesWithLabel(gmail,"reply"),
+edgesWithLabel(gmail,"typeBody"),
+edgesWithLabel(gmail,"send"),
+edgesWithLabel(gmail,"compose"),
+edgesWithLabel(gmail,"addRecipient"),
+edgesWithLabel(gmail,"open(files)"),
+edgesWithLabel(gmail,"writeBody"),
+edgesWithLabel(gmail,"writeSubject"),
+edgesWithLabel(gmail,"send")
+\end{lstlisting}
+In plain language such a scenario means opening one and then a second conversation,
+replying to the last message of the second one by typing the body of
+the response, sending it, and then composing a new message by adding a
+recipient, an attachment, typing the body and subject, and finally
+sending it.
+
+The execution traces for such a Gmail scenario, with a detour limit of
+3, consist of a graph with 474 vertices and 1753 edges. These traces
+entail 12 possible geodesic paths with order 0, 82 with order 1, 30
+with order 2 and 33 with order 3. Figure~\ref{fig:numpaths} shows the
+number of paths obtained for the four applications, split by detour
+order.
+
+\begin{figure}[tbh]
   \centering
-  \includegraphics[width=\linewidth]{figures/averagecase.png}
-  \caption{Average length of  geodesic paths split by detour order.}
+  \includegraphics[width=\linewidth]{figures/gpd.png}
+  \caption{Number of geodesic paths split by detour order.}
+  \label{fig:numpaths}
 \end{figure}
 
-\begin{figure}
+% tot.hom
+% ##     N min max        M       rel  N
+% ## d0  4  15  18 15.75000 0.2539683  4
+% ## d1 32  16  22 17.94531 1.7831955 36
+% ## d2 18  18  24 19.25000 0.9350649 54
+% ## d3 29  19  24 19.95690 1.4531317 83
+%  tot.gom
+% ##     N min max        M       rel   N
+% ## d0 12  13  17 15.08333 0.7955801  12
+% ## d1 82  14  24 19.11179 4.2905456  94
+% ## d2 30  16  25 20.57778 1.4578834 124
+% ## d3 33  17  26 21.51515 1.5338028 157
+%  tot.sqm
+% ##     N min max        M       rel   N
+% ## d0  4  15  19 17.00000 0.2352941   4
+% ## d1 44  16  24 19.77273 2.2252874  48
+% ## d2 36  18  26 20.88889 1.7234043  84
+% ## d3 56  19  25 21.64286 2.5874587 140
+%  tot.rcm
+% ##     N min max        M       rel  N
+% ## d0  2  15  18 16.50000 0.1212121  2
+% ## d1 27  16  24 19.07407 1.4155340 29
+% ## d2 22  18  25 20.18182 1.0900901 51
+% ## d3 31  19  24 21.04839 1.4727969 82
+
+Gmail has the highest number of order 0 and 1 paths, and the highest
+difference between the number of order 0 and 1, and between order 1
+and 2 (at least 3 times as many order 0 paths than any of the other
+applications, and at least twice as many order 1 paths as the any of
+the other ones).  Gmail offers 3 times as many error-free alternative
+paths to accomplish the scenario, which indicates that users might
+more easily follow one of those paths. However, Gmail offers also
+almost twice as many order 1 paths, which means that users could be
+more easily induced into an erroneous path than when using another
+system. Because the number of order 2 or 3 paths decreases, Gmail
+reduces therefore the ``error proneness'' of this UI, for executions
+that involve 2 or 3 errors. Notice that Roundcube has the smallest
+number of order 0 paths (2 of them), which means that users are not
+given much flexibility and freedom in carrying out correctly the
+scenario.
+For none of the systems a detour leads to dead-ends preventing the
+user to complete the scenario.
+
+Figure~\ref{fig:bestcase} shows the length of paths in the best case,
+i.e., when users would always choose the shortest route. Gmail offers
+the shortest paths across the four detour orders (for order 0 the
+length is 13 steps; for order 3 the length is 17 steps; the other
+applications are remarkably similar among them).  A plausible
+interpretation is therefore that Gmail not only offers many more
+error-free paths, but also gives the shortest ones. Users are given
+more flexibility and more efficiency.
+
+\begin{figure}[htb]
   \centering
-  \includegraphics[width=\linewidth]{figures/bestcase.png}
+  \includegraphics[width=\linewidth]{figures/min-paths.png}
   \caption{Minimum length of  geodesic paths split by detour order.}
+  \label{fig:bestcase}
 \end{figure}
 
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/worstcase.png}
-  \caption{Maximum length of geodesic paths split by detour order.}
-\end{figure}
+%Figure~\ref{fig:averagecase}
+
+% \begin{figure}[htb]
+%   \centering
+%   \includegraphics[width=\linewidth]{figures/mean-paths.png}
+%   \caption{Average length of  geodesic paths split by detour order.}
+%   \label{fig:averagecase}
+% \end{figure}
 
@@
   \caption{Frequency of an optimal path.}
+  \label{fig:freqopt}
 \end{figure}
+
+To combining these two results, we can easily compute the frequency of
+paths with different length. Figure~\ref{fig:freqopt} shows, for each
+application, the frequency of order 0 paths, the frequency of paths
+with length less or equal to 15 (the minimum length across the four
+applications), and the frequency of optimal paths (the shortest
+ones). These values show that Gmail users have the lowest probability
+to hit an order 0 path (because of the relative large number of order
+1 paths made available by Gmail), have the highest probability to hit
+a path with length 15 or less, have the lowest probability of hitting
+the shortest paths.  Thus, flexibility and efficiency that can be
+exploited with Gmail is counterbalanced by the required knowledge and
+capability of chosing an optimal path. In other words, Gmail offers
+many detours of order 1 which increase flexibility for some users and
+might decrease effectiveness for other ones.
+
+% >   d=0.050 # 20% chances that a user makes a generic mistake
+% >   dprh=detour.page.rank(sc.hom,3,d);dprh
+%          0          1          2          3 
+% 0.73455373 0.11735984 0.05684419 0.09124223 
+% >   dprg=detour.page.rank(sc.gom,3,d);dprg
+%          0          1          2          3 
+% 0.69210344 0.17697177 0.06228221 0.06864257 
+% >   dprs=detour.page.rank(sc.sqm,3,d);dprs
+%          0          1          2          3 
+% 0.79062415 0.07122957 0.05364190 0.08450438 
+% >   dprr=detour.page.rank(sc.rcm,3,d);dprr
+%          0          1          2          3 
+% 0.76538478 0.08355996 0.06247253 0.08858273 %
+
+Another probabilistic analysis can be performed using page rank, which
+can be used to compute the probability that a random walk in a graph
+visits a certain vertex. We computed the page rank (with a damping
+value of 5\% - meaning that the randowm walker with probability 5\%
+jumps to an arbitrary state and probability 95\% chooses one of the
+actions available in the current state) for each vertex in the graph,
+and then added the page rank value for vertices with different detour order.
+Figure~\ref{fig:pr} shows the resulting probabilities.
 
@@
   \includegraphics[width=\linewidth]{figures/pr.png}
-  \caption{Probability that a random walk visit detour 0, 1, 2 or 3 states.}
+  \caption{Probability that a random walk visits detour 0, 1, 2 or 3 states.}
+  \label{fig:pr}
 \end{figure}
+
+With Gmail the probability of visiting an order 0 state is close to
+70\%, the lowest among the four applications. But when it comes to
+visiting an order 0 or 1 state, the probability increases to 87\%,
+which is the highest. Thus, to compare Gmail with SquirrelMail, a
+completely random usage of SquirrelMail has 10\% more probability of
+hitting an error-free state than Gmail. That advantage is reduced when
+considering order 0 and 1 paths, because with SquirrelMail the
+probability is 85\% and Gmail it is 87\%.
+
+Manual inspection of the shortest paths indicates that one reason of
+the higher potential efficiency offered by Gmail is due to the
+fact that users can start composing a new message while reading a
+conversation, whereas in other applications an explicit ``closing'' of
+the reading activity has to be performed. Another reason is in the
+more streamlined process to attach a file: in Gmail one needs to
+select the file(s) and they are automatically uploaded, whereas in
+other applications one has to explicitly perform the uploading step
+after selecting them.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figures/ad.png}
+  \caption{Action density.}
+  \label{fig:ad}
+\end{figure}
+
+Figure~\ref{fig:ad} shows the \emph{action density} of the four
+applications, in terms of average number of actions per state involved
+in traces, and average number of \emph{unique} actions per state. The
+former is an overall measure of the number of options that are made
+available by a UI, the latter can be used to analyse how many
+\emph{new} options the user is presented with in any state. For our
+examples, the values are all in the range between 5.9 actions/state in
+the case of Gmail and 7.1 for Horde, and 1.3 unique actions for
+SquirrelMail and 2 for Horde. This suggests that Gmail features a more
+compact design (fewer actions to do the same things), and SquirrelMail
+is even more compact when it comes to the different types of actions;
+thus it could be easier to learn.
+
+\subsection{Comparing scenarios}
+
+The same kind of analysis can be carried out to assess how suitable a
+design is for a set of different scenarios. For example, a designer
+might be interested in understanding what are some quantitative
+differences in replying to a message as opposed to composing a new
+one.
+
+For Horde, it turns out that composing a new message entails many more
+order 0 paths (154 vs 86), and a comparable number of higher order
+paths. The path length in the best case is the same across the two
+scenarios, but in the average case composing has a length of 7.75
+steps compared to 8.5 for reply. Probability of hitting an optimal
+path is 4 times higher for compose.
+
+This means that users will have a twice as much large choice of
+correct paths when composing a new message rather than ``simply''
+replying to a read one. On average, when composing, users could be
+9\% more efficient, and they are 4 times more likely to do the right
+things. Thus, Horde is more suitable for composing new messages than
+it is for replying to existing ones.
+
+\subsection{Adding features}
+
+Execution traces can be used also to assess what is the effect of
+adding a widget or feature to an existing UI.  For example, we studied
+the cruise control features of cars. One of the examples is system S,
+where the driver can engage the system, and once it is engaged, speed
+can be increased or decreased with small or large steps. Of course the
+system can also be disengaged (by pressing the brake pedal, for
+example).
+System A is more elaborate, as it includes also a memory function:
+when it is disengaged it remembers the current speed, which can be
+recalled later on.  There are two ways to re-engage it: one
+by setting a new speed, and one by recalling the previous one. In
+addition,  if the car drives for more than 5 minutes at a higher
+speed than the set one, system A automatically disengages.
+
+Thus, one possible design question is ``What are the effects of adding
+these functionalities?'' in a typical driving scenario where a speed
+is selected, then the system is disengaged, and then the same speed
+needs to be set?
+
+System S (where we assumed that re-setting the speed is done manually
+by the driver with 4 actions on ``up'' and ``down'' to get
+approximately get the same speed) leads to 1 order 0 path requiring 7
+steps; there are 8 order 1 paths with average length  9.4. The
+probability of hitting the optimal path is 0.11; action density is
+1.75, and unique action density is 1.25.
+
+On the other hand, system A has 2 order 0 paths (average length 4.5),
+and 7 order 1 paths (average length 6.8). 
+The optimal path probability is 0.06; action density is
+2.2, and unique action density is 1.2.
+
+In both cases the reasons for detours deal with the possibility of
+disengaging the system at the wrong moment.
+
+Therefore we can conclude that:
+\begin{inparaenum}
+\item system A makes users more efficient;
+\item with A there are two possible ways to achieve the scenario, thus
+  more flexibility is given;
+\item with A the probability of doing the right thing is almost half
+  of system S: it might be more difficult to do the right thing
+  because more possibilities are offered;
+\item system A features a more compact design, with fewer action types
+  to be performed at each moment, being therefore potentially easier
+  to learn and remember.
+\end{inparaenum}
+
+\hl{present them in terms of :
+consistency,
+error proneness, efficiency, premature committment, flexibility}
+
+\clearpage
+
+% \begin{table*}[ht]
+% \centering
+% \begin{tabular}{rllll}
+%   Step &Horde & Gmail & Squirrel & Roundcube\\
+%   \hline
+%   1 & openMessage(msg) & open(conversation) & openMessage(msg) & openMessage(msg) \\ 
+%   2 & exitMessage & goBack & exitMessage &  exitMessage \\ 
+%   3 & openMessage(msg) & open(conversation) & openMessage(msg) & openMessage(msg) \\ 
+%   4 & reply & reply & reply & reply \\ 
+%   5 &  typeBody & typeBody & typeBody                       &  typeBody \\ 
+%   6 &  send &  send & send                                  & send \\ 
+%   7 &  exitMessage &  & exitMessage                  &  exitMessage \\ 
+%   8 & compose &compose & compose                        & compose \\ 
+%   9 & addReceiver(receiver) &  addReceiver(receiver)  & addReceiver(receiver)   & addReceiver(receiver) \\ 
+%   10 & browseFiles          & attachFiles   & browseFiles  & attachAFile \\ 
+%   11 & selectAttachments(file) & open(files)  & selectAttachments(file)          &  selectAttachments(files) \\ 
+%   12 & update                &  &add   &  upload \\ 
+%   13 &  typeSubject          &  typeSubject          &  typeSubject & typeSubject \\ 
+%   14 & typeBody &         typeBody                    & typeBody     & typeBody \\ 
+%   15 &  send &           send                    &  send        &  send \\ 
+%    \hline
+% \end{tabular}
+% \end{table*}
 
@@
 \label{sec:discussion}
+
+\hl{dire} di SEPARATING concerns, ad es M-V-C, per vedere le
+  implicazioni di ciascuna e non soffrire di ``saturation'', vedi adams
+
+dire che le ns analisi sono prive di assunzioni su conoscenze
+abilita'; le probability sono puramente sulla base delle frequenze.
+
+dire di cogtools: noi non prevediamo i tempi (richiede modeling
+psicologico e dei specifici widgets). pero' forniamo un'analisi che
+puo' guidare chi applica cogtool.
+
+dire di horrocks \cite{horrocks99}
+
+dire di agile modeling \cite{ambler02}
+
+issues dealing with complexity of searching paths
 

### 2015-09-19 paper.tex
introduction should be ok

@@
 \usepackage{todonotes}
+\newcommand{\mytodo}[1]{\todo[inline]{#1}}
 \usepackage{soul}
@@
 
-\title{Measuring Interaction Design\\
-Measuring Interaction before developing prototypes\\
-Measuring Interaction Graphs}
+\title{Measuring Interaction Design}
 
@@
 \begin{abstract}
-  Early development of prototypes of user interfaces is good
-  practice. However, they have to cover specific usage scenarios, and
-  because they are limited in focus, the whole picture of the user
-  interface is easily lost. Even simple questions dealing with number
-  and length of interaction paths or impact of possible user errors
-  can be answered only for the specific scenarios being analysed.
+  Early prototyping of user interfaces is an established good
+  practice. However, because prototypes cover only some usage
+  scenarios, the whole picture of the user interface is easily
+  lost.  Questions dealing with number of required steps,
+  possible interaction paths or impact of possible user errors can be
+  answered only for the specific scenarios being analysed and after
+  tedious manual inspection.
 
   We present a tool that transforms models of a user interface into a
-  graph. This is then used to specify usage scenarios, and to generate
-  possible interaction paths. Metrics based on possible paths, with or
-  without possible mistakes, can be easily computed. When applying
-  these metrics get precise and objective values that can be used to
-  compare designs or to identify weakness. When applying them to mail
-  applications, we learn for example that Gmail has \hl{7 times more
-  optimal paths, it has 4x more paths include at most one possible
-  user error, that it has 20\% fewer steps.}
+  graph, upon which usage scenarios can be easily specified, and used
+  to compute possible interaction paths. Metrics based on possible
+  paths, with or without possible user errors, can be easily computed.
+  For example, when applying them to mail applications, we learn that
+  Gmail has 3 times more shortest routes, has 2x more routes that
+  include one user error, has routes with 13\% fewer steps, but has
+  the smallest probability to hit an optimal route.
 
@@
 \begin{quote}
-  Subcommittee: Technology, Systems, and Engineering
+\mytodo{ Subcommittee: Technology, Systems, and Engineering
 
@@
 mobile/ubiquitous computing.
+}
 
@@
 We present an approach that allows a designer to assess interaction
-design (IxD) qualities such as efficiency, consistency,
+design (IxD) qualities such as efficiency,
 error-proneness and recovery from premature committment. Key
@@
 \item understand how recoverable the UI is from those errors; and,
-\item support task analysis and task/scenario design.
+\item perform objective quantitative comparison of different designs
+  to support construction and engineering of interactive systems.
 \end{inparaenum}
-%
-This allows different designs to be objectively compared to support
-designers in the construction and engineering of interactive
-systems. To explain our work we present a comparison of four different
-web mail front ends. We have chosen this domain because it is very
-well known and yet several trivial questions lead to non trivial
-results. We applied the same techniques also in other domains, such as
-HVAC and other embedded UIs.
 
-The approach is based on UML statechart models of UIs which are
-automatically processed to produce \emph{interaction graphs}. These
-are then used to specify interaction scenarios and unfold interaction
-paths (called \emph{execution traces}) determined by the specific
-scenarios being considered. Traces are fed to several graph-theoretic
-computations which produce a dashboard with different results with the
-answers. Except for development of models and specification of the
-desired scenarios, which have to be done manually, the other steps are
-totally automatic; models of a design (such as the ones shown in the
-paper) could be developed in a matter of a couple of hours.
+To explain our work we present a comparison of four web mail front
+ends. We have chosen this domain because it is easy to understand and
+yet several trivial questions are not easily answered. We applied
+the same techniques also in other domains, such as HVAC and other
+embedded UIs.
 
@@
 %
-Adopting Usage Centred Development practices is effective, as is
+Adopting Usage Centered Development (UCD) practices is effective, as is
 following established design principles~\cite{constantine99}. In
 particular, one of the most effective technique is early
-prototyping~\cite{buxton07} to explore part of the five-dimensional
-prototyping space~\cite{mccurdy06}. It is particularly effective when
+prototyping to explore part of the five-dimensional fidelity
+prototyping space~\cite{mccurdy06,buxton07}. It is particularly effective when
 paired with usability investigations based on user testing or
-heuristic evaluations~\cite{rubin08,preece02}.
+heuristic evaluations~\cite{preece02,rubin08}.
 
-However, it requires prototypes which are usually developed with
+However, UCD requires prototypes which are usually developed with
 certain tasks in mind, and therefore are quite restricted in terms of
@@
 involved in the analysis. Thus, although a  significant effort
-needs to be expended to develop and use prototypes, less than optimal
+needs to be directed to develop and use prototypes, less than optimal
 results are obtained.
 
-Even worse, there might be questions whose answer cannot be easily
-found. For example, given one or more potential IxD solutions and some
-usage scenarios, interesting questions could include: ``How many
-different ways can be followed by the user to carry out the scenario?'', ``Which
-are the shortest ones?'', ``If a user makes a mistake, would he or she
-be able to recover?'', ``How many steps would the recovery
-require?''. When designing and evaluating embedded UIs (such as when
-dealing with plane's cockpits~\cite{conversy07}), other relevant
-questions might include ``How would the above properties change if we
-add a certain a widget?'', or ``... if we replace a widget with
-another?''.  At the moment, even these straightforward questions are
-quite complicated to answer. In fact, they require development of
-prototypes, inspecting them, manual tracking of which screens and
-widgets are used at which stage, and a systematic manual analysis.
+Even worse, there might be questions that cannot be cannot be easily
+answered. Given one or more potential designs and some usage
+scenarios, interesting questions could include: ``How many different
+routes can be followed by the user to carry out the scenario?'',
+``Which are the shortest ones?'', ``If a user makes a mistake, would
+he or she be able to recover?'', ``How many steps would the recovery
+require?''. In a different setting, when designing and evaluating
+embedded UIs (such as when dealing with plane's
+cockpits~\cite{conversy07}), other relevant questions might include
+``How would the above properties change if we add a certain a
+widget?'', or ``... if we replace a widget with another?''.  At the
+moment, these straightforward questions are quite complicated to
+answer. In fact, they require development of prototypes, inspecting
+them, manual tracking of which screens and widgets are used at which
+stage, and a systematic manual analysis.
 
-This should not be the case, however. These answers provide important
-insights to a designer, and potentially support assements of potential user
-flexibility, user efficiency, error proneness, ability to recover,
-compactness and consistency of a design.
+This should not be the case, however. Answers to these questions could
+be automatically computed. In this way a system could provide
+important insights to a designer, and  support assements of
+potential user flexibility, user efficiency, error proneness, ability
+to recover, compactness and consistency of a design.
 
-Our contribution consists of the development of a tool that transforms
-models and scenario specifications into interaction paths, and the
-definition of metrics that provide concise, precise and objective
-measures of a design. The examples included in the paper show that
-among four web mail applications, and with respect to a typical usage
-scenario, Gmail is the most efficient and flexible UI, with the best
-ability to recover from certain mistakes. In fact, Gmail has the
-largest number of shortest sequences of steps (even when users are
-supposed to make 1 or 2 mistakes), but when users make more than 2
-mistakes the number of paths drops significantly; on average, best and worst
-cases, Gmail features also shorter paths, requiring \hl{CHECK 20\% fewer steps};
-however, the probability that a user hits an optimal path with Gmail
-is almost half of that of another application, and the probability
-that a random walker hits a state that is not due to a mistake is
-\hl{SAY SOMETHING}.
+Our approach is based on UML statechart models of UIs which are
+automatically processed to produce \emph{interaction graphs}. These
+graphs are then used to specify interaction scenarios and to unfold
+the possible interaction paths (called \emph{execution traces})
+determined by the specific scenarios being considered. Traces are fed
+to several graph-theoretic computations which produce a dashboard with
+different results. Except for development of models and specification
+of the desired scenarios, which have to be done manually, the other
+steps are totally automatic; models of a design (such as the ones
+shown below) can be developed in a matter of a couple of hours.
+
+Our contribution consists of
+\begin{inparaenum}[(a)]
+\item the idea of using a model to specify interaction scenarios,
+\item the development of a tool (MIG-TOOL, Measuring Interaction
+  Graphs Tool) that transforms models and scenario specifications into
+  interaction paths, and
+\item the definition of metrics that provide concise, precise and
+  objective measures of a design.
+\end{inparaenum}
+The examples included in the paper show that among four web mail
+applications, and with respect to a typical usage scenario, Gmail is
+the most efficient and flexible UI, with the best ability to recover
+from user errors, but only for users that possess a certain level of
+proficiency. In fact, Gmail has the largest number of shortest paths
+(when users are supposed to make 0 or 1 error); when users make 2 or
+more errors the number of paths drops significantly, reducing the
+error-proneness of the UI; on the best case, Gmail features also the
+shortest paths, requiring 13\% fewer steps; however, the probability
+that a user hits an optimal path with Gmail is 10 times smaller than
+the best of the other applications, and the probability that a random
+walker hits a state that is not due to an error is 10\% smaller than
+the best of the other applications.
 %
@@
 slightly different interaction structures adopted for uploading
-messages. If Gmail didn't exist yet, its designers could obtain these
-answers well before performing usability studies.
+messages. If Gmail didn't exist yet, by using MIGTOOL its designers
+could obtain these answers well before developing prototypes and
+performing usability studies.
 
@@
 when carrying out the
+
+issues dealing with complexity of searching paths
 
@@
 applications, and at least twice as many order 1 paths as the any of
-the other ones).  Gmail offers 3 times as many error-free alternative
+the other ones). 
+ Gmail offers 3 times as many error-free alternative
 paths to accomplish the scenario, which indicates that users might
-more easily follow one of those paths. However, Gmail offers also
+more easily follow one of those paths. 
+\mytodo{result}
+However, Gmail offers also
 almost twice as many order 1 paths, which means that users could be
 more easily induced into an erroneous path than when using another
-system. Because the number of order 2 or 3 paths decreases, Gmail
+system.  
+\mytodo{result}
+Because the number of order 2 or 3 paths decreases, Gmail
 reduces therefore the ``error proneness'' of this UI, for executions
@@
 scenario. 
+\mytodo{result}
 For none of the systems a detour leads to dead-ends preventing the
@@
 the shortest paths across the four detour orders (for order 0 the
-length is 13 steps; for order 3 the length is 17 steps; the other
+length is 13 steps, saving more than 13\% steps compared to other
+applications; for order 3 the length is 17 steps; the other
 applications are remarkably similar among them).  A plausible
@@
 error-free paths, but also gives the shortest ones. Users are given
-more flexibility and more efficiency.
+more flexibility and more efficiency.  \mytodo{result} Because also
+paths with order 1 or more are the shortest ones among the four
+applications, Gmail also makes users more efficient in recovering from
+errors.  \mytodo{result}
 
@@
 
-To combining these two results, we can easily compute the frequency of
+To combine these two results, we can easily compute the frequency of
 paths with different length. Figure~\ref{fig:freqopt} shows, for each
@@
 a path with length 15 or less, have the lowest probability of hitting
-the shortest paths.  Thus, flexibility and efficiency that can be
-exploited with Gmail is counterbalanced by the required knowledge and
-capability of chosing an optimal path. In other words, Gmail offers
+the shortest paths (10 times smaller than the best of the other
+applications).  Thus, the flexibility and efficiency that can be
+exploited with Gmail are counterbalanced by the required knowledge and
+capability of chosing an optimal path. In particular, Gmail offers
 many detours of order 1 which increase flexibility for some users and
-might decrease effectiveness for other ones.
+might decrease effectiveness for less knowledgeable ones.
+\mytodo{result}
 
-% >   d=0.050 # 20% chances that a user makes a generic mistake
+% >   d=0.050 # 5% chances that a user makes a generic mistake
 % >   dprh=detour.page.rank(sc.hom,3,d);dprh
@@
 can be used to compute the probability that a random walk in a graph
-visits a certain vertex. We computed the page rank (with a damping
-value of 5\% - meaning that the randowm walker with probability 5\%
+visits a certain vertex~\cite{page1999pagerank}. We computed the page rank (with a damping
+value of 5\% - meaning that the random walker with probability 5\%
 jumps to an arbitrary state and probability 95\% chooses one of the
@@
 considering order 0 and 1 paths, because with SquirrelMail the
-probability is 85\% and Gmail it is 87\%.
+probability is 85\% and Gmail it is 87\%.  This means that somebody
+with no knowledge on how to use an email front end, when using Gmail
+would have 10\% fewer chances of carrying out the scenario without
+making any error, as opposed to when using SquirrelMail: SquirrelMail
+provides more guidance. Across the four UIs, the probability of making
+at most 1 error is approximately the same\footnote{These results are
+  not very sensitive to the value chosen for the damping factor:
+  differences across the four applications are very similar when $d$
+  ranges in $[0.05,0.20]$.}.
+
+\mytodo{result}
 
@@
 after selecting them.
+
+\mytodo{result}
 
@@
 thus it could be easier to learn.
+
+\mytodo{result}
 
@@
 it is for replying to existing ones.
+\todo[inline]{result}
 
@@
 is selected, then the system is disengaged, and then the same speed
-needs to be set?
+needs to be set.
 
 System S (where we assumed that re-setting the speed is done manually
-by the driver with 4 actions on ``up'' and ``down'' to get
+by the driver with 4 actions on ``up'' and ``down'' to
 approximately get the same speed) leads to 1 order 0 path requiring 7
@@
 \end{inparaenum}
-
-\hl{present them in terms of :
-consistency,
-error proneness, efficiency, premature committment, flexibility}
-
-\clearpage
+\todo[inline]{result}
 
@@
 
+feasibility:
+
+sensitivity with respect to models
+
+dire di horrocks \cite{horrocks99}
+
+dire di agile modeling \cite{ambler02}
+
 \hl{dire} di SEPARATING concerns, ad es M-V-C, per vedere le
   implicazioni di ciascuna e non soffrire di ``saturation'', vedi adams
+
+valdiity of results:
 
@@
 
-dire di horrocks \cite{horrocks99}
-
-dire di agile modeling \cite{ambler02}
-
-issues dealing with complexity of searching paths
+MIGTOOL is part of a suite of model-based tools for designing,
+analysing and testing user interfaces developed by the company
+ANONYMIZED. Development of a web-based user interface of MIGTOOL is
+underway; the company plans to make it freely available for research
+purposes.
 

### 2015-09-24 paper.tex
first version of background section completed.

@@
 \newcommand{\mytodo}[1]{\todo[inline]{#1}}
+\newcommand{\ie}{\emph{i.e.}}
+\newcommand{\eg}{\emph{e.g.}}
+\newcommand{\migtool}{MIG-Tool\ }
 \usepackage{soul}
@@
 
-Even worse, there might be questions that cannot be cannot be easily
+Even worse, there might be questions that cannot be easily
 answered. Given one or more potential designs and some usage
@@
 slightly different interaction structures adopted for uploading
-messages. If Gmail didn't exist yet, by using MIGTOOL its designers
+messages. If Gmail didn't exist yet, by using \migtool its designers
 could obtain these answers well before developing prototypes and
@@
 
-(right now these are random notes)
+The literature on using state-transition network representations for
+specifying or analysing the behavior of UIs is vast. We conducted a
+systematic-style literature review, using Google Scholar and queries
+with combinations of these phrases: ``user interface'', ``path
+analysis'', ``user trace'', ``interaction trace'', ``navigation'',
+``markov chain'', ``state transition'', ``statechart'', ``metric'',
+``measure''. For each query we analyzed title and abstract of the
+first 50 hits and appraised their relevance based on whether the paper
+discusses approaches for measuring user interfaces in the context of
+usability and on the basis of a model based on state transitions. This
+resulted in 78 full-text papers that were later on re-analyzed against
+the same criterion, leading to several of papers mentioned below.
+%
+Several of the examined papers deal with testing user interfaces, and
+problems related to generating execution traces as a means to assess
+coverage of a test
+suite~\cite{ammann2008introduction,briand2005:test:uml}. They were
+excluded from the review.
 
-SwingState: work done by 
-Caroline Appert\cite{appert2008swingstates}
+Usage of state-transition networks to model UI behavior in order to
+draw usability conclusions dates back at least to~\cite{parnas69}. In
+it, Parnas claimed that several kinds of usability problems would not
+occur if the designer adopted a design framework where states and
+their transitions are made explicit.
 
-in java, FSA used as a conceptual framework to write the code of
-widgets so that events and event handlers in the ui can more easily be
-conceived, developed and verified.
+In many cases \emph{statecharts} are used (a generalization of finite
+state automata, with an expressive language that includes hierarchic
+levels of abstraction, concurrent regions, states and pseudostates, an
+extended state notion based on an arbitrary computational model).
+Harel, the inventor of statecharts, gives an interesting retrospective
+view of how they were invented, back in early eighties, and why they
+were appealing also to non experts~\cite{harel09}.  For the sake of
+brevity, we refrain from describing them here in detail; the
+interested reader is referred to the UML standard~\cite{uml241} or
+some of the textbooks that deal with them, like~\cite{samek09}.
+%
+Horrocks showed how statecharts can be  used to model and
+specify the dynamics of typical desktop UIs~\cite{horrocks99}. While
+providing many interesting insights on how and why one should use
+statecharts to do so, this nice work does not address how such a
+specification could be \emph{automatically} processed.
+%
+This idea was later on greatly expanded by
+Thimbleby~\cite{thimbleby07} , which discusses the use of statecharts
+as meta-models for user interfaces; they are seen as crucial
+representations that allow a designer to fully appreciate how devices
+behave. The overall claim is that ``If you don't understand the logic
+conveyed by a statechart model of a user interface, then you don't
+understand the behavior of the user interface''.
 
-They say: StateCharts were used in the StateMaster User Interface
-Management System [40], and a variant of them specifically tailored to
-designing user interfaces was used in the more recent HsmTk toolkit
-[18]. StateCharts however are significantly more complicated and hard
-to learn than plain state machines, and our experience is that user
-interface designers and developers have difficulties exploiting their
-power. Other approaches include Petri Nets [41], which have also been
-used to specify user interfaces, for example in the PetShop system
-[42]. Here too, the learning curve is steep, making the adoption of
-such a model by developers difficult.
+WebML~\cite{ceri:00} is one of the most successful model-driven
+approaches to web development (UIs and backend systems), with
+industrial traction and a large number of publications. The language
+is based on state transitions and is targeted to automatic generation
+of data-intensive web applications.  %WebML encompasses different
+%meta-models, independent from the technologies used to deliver the
+%final application. % The WebML
+% meta-model provides relatively few primitive units that can be
+% combined to generate quite complex interactions; the modeling language
+% is thus powerful and concise.  
+Many other similar approaches involve or are based on task or activity
+models~\cite{mori02,silva03:umli,koch02:uwe,gomez01,melia08}.
+%
+A recent OMG standard, called Interaction Flow Modeling Language
+(IFML)~\cite{ifml13}, derives from WebML and focuses specifically on
+user interaction. IFML is a language for specifying the structure of a
+user interface and its behavior. It offers most of the abstractions
+that are available in statecharts, mixed with the ability to specify
+so-called ``components'' that are used to display and manipulate data
+(to display details of a item, to display or select lists of items, to
+input an item).
 
-With SwingStates, any number of state machines can run simulateneously. A state machine
-can be active, i.e., handling the events it receives, or inactive,
-i.e., ignoring events
+A rather different route for the problem of generating UIs is followed
+in~\cite{heymann2007formal}. They assume that the UI to be generated
+is used to supervise and to monitor an underlying machine (\eg,
+autopilot of a plane) which is modeled as a statechart. After assuming
+that the behavior of the UI can be modeled as a statechart, they
+devised an algorithm that checks whether the two models are
+compatible, and that refines the UI model so that its states and
+transitions are minimized while still allowing a correct manipulation
+of the underlying machine. Application of such a technique leads to
+UIs that are correct by-construction.
 
-We adopt a similar apporach, but using UML state machine (which are
-more powerful than FSA) for conceiving, guiding development, analysis
-and verification of the behavior of the entire app.
+In~\cite{conversy07} the Interactive Cooperative Objects formalism is
+discussed, which is based on Petri nets, yet another state transition
+formalism.  It is used to enrich the ARINC 661 specification of
+Cockpit Display System used in interactive cockpits of airplanes, so
+that the semantics of widgets can be expressed and the behavior of a
+UI be formally analyzed.
 
-Nebeling says:\cite{nebeling2011metrics}
+%  that are now under provides precise information
+% for communication protocol between application (called User
+% Applications) and user interface components (called widgets) as well
+% as precise information about the widgets themselves. However, in ARINC
+% 661, no information is given about the behaviour of these widgets and
+% about the behaviour of an application made up of a set of such
+% widgets.  The Interactive Cooperative Objects (ICOs) formalism is a
+% formal description technique dedicated to the specification of
+% interactive systems [4, 11]. It uses concepts borrowed from the
+% object-oriented approach to describe the structural or static aspects
+% of systems, and uses high-level Petri nets [8] to describe their
+% dynamic or behavioural aspects.
 
-Despite the fact that screen sizes and average screen resolutions have
-dramatically increased over the past few years, little attention has
-been paid to the design of web sites for large, high-resolution
-displays that are now becoming increasingly used both in enterprise
-and consumer spaces. We present a study of how the visual area of the
-browser window is currently utilised by news web sites at different
-widescreen resolutions. The analysis includes measurements of space
-taken up by the article content, embedded ads and the remaining
-components as they appear in the viewport of the web browser. The
-results show that the spatial distribution of page elements does not
-scale well with larger viewing sizes, which leads to an increasing
-amount of unused screen real estate and unnecessary scrolling. We
-derive a number of device-sensitive metrics to measure the quality of
-web page layout in different viewing contexts, which can guide the
-design of flexible layout templates that scale effectively on large
-screens.
+Finite state representations have been used also as a conceptual
+framework to write the code of widgets so that events and event
+handlers in the UI can more easily be conceived, developed and
+verified (\eg, \cite{appert2008swingstates}).
 
-Stephane Conversy\cite{conversy07} says: (friend of nicolas roussel in
-the committee)
+% in java, FSA used as a conceptual framework to write the code of
+% widgets so that events and event handlers in the ui can more easily be
+% conceived, developed and verified.
 
-ARINC 661 provides precise information for communication protocol
-between application (called User Applications) and user interface
-components (called widgets) as well as precise information about the
-widgets themselves. However, in ARINC 661, no information is given
-about the behaviour of these widgets and about the behaviour of an
-application made up of a set of such widgets.
+% They say: StateCharts were used in the StateMaster User Interface
+% Management System [40], and a variant of them specifically tailored to
+% designing user interfaces was used in the more recent HsmTk toolkit
+% [18]. StateCharts however are significantly more complicated and hard
+% to learn than plain state machines, and our experience is that user
+% interface designers and developers have difficulties exploiting their
+% power. Other approaches include Petri Nets [41], which have also been
+% used to specify user interfaces, for example in the PetShop system
+% [42]. Here too, the learning curve is steep, making the adoption of
+% such a model by developers difficult.
 
-The purpose of ARINC 661 specification (ARINC 661, 2002) is to define
-interfaces to a Cockpit Display System (CDS) used in interactive
-cockpits that are now under
+% With SwingStates, any number of state machines can run simulateneously. A state machine
+% can be active, i.e., handling the events it receives, or inactive,
+% i.e., ignoring events
 
-application of a formal description technique to the various elements
-of ARINC 661 specification within an industrial project. This formal
-description technique called Interactive Cooperative Objects defines
-in a precise and non-ambiguous way all the elements of ARINC 661
-specification. The application of the formal description techniques is
-shown on an interactive application called MPIA (Multi Purpose
-Interactive Application). Within this application, we present how ICO
-are used for describing interactive widgets, User Applications and
-User Interface servers (in charge of interaction techniques). The
-emphasis is put on the model-based management of the feel of the
-applications allowing rapid prototyping of the external presentation
-and the interaction techniques.
+% We adopt a similar apporach, but using UML state machine (which are
+% more powerful than FSA) for conceiving, guiding development, analysis
+% and verification of the behavior of the entire app.
 
-The Interactive Cooperative Objects (ICOs) formalism is a formal
-description technique dedicated to the specification of interactive
-systems [4, 11]. It uses concepts borrowed from the object-oriented
-approach to describe the structural or static aspects of systems, and
-uses high-level Petri nets [8] to describe their dynamic or
-behavioural aspects.
+% Nebeling says:\cite{nebeling2011metrics}
 
-The Interactive Cooperative Objects (ICOs) formalism is a formal
-description technique dedicated to the specification of interactive
-systems [4, 11]. It uses concepts borrowed from the object-oriented
-approach to describe the structural or static aspects of systems, and
-uses high-level Petri nets [8] to describe their dynamic or
-behavioural aspects.
+% Despite the fact that screen sizes and average screen resolutions have
+% dramatically increased over the past few years, little attention has
+% been paid to the design of web sites for large, high-resolution
+% displays that are now becoming increasingly used both in enterprise
+% and consumer spaces. We present a study of how the visual area of the
+% browser window is currently utilised by news web sites at different
+% widescreen resolutions. The analysis includes measurements of space
+% taken up by the article content, embedded ads and the remaining
+% components as they appear in the viewport of the web browser. The
+% results show that the spatial distribution of page elements does not
+% scale well with larger viewing sizes, which leads to an increasing
+% amount of unused screen real estate and unnecessary scrolling. We
+% derive a number of device-sensitive metrics to measure the quality of
+% web page layout in different viewing contexts, which can guide the
+% design of flexible layout templates that scale effectively on large
+% screens.
 
-Voida\cite{voida} says:
-We describe an alternative model for organizing the
-desktop interface-activity-based computing-and
-identify a series of high-level system requirements for
-interfaces that use activity as their primary organizing
-principle.
+% The purpose of ARINC 661 specification (ARINC 661, 2002) is to define
+% interfaces to a Cockpit Display System (CDS) used in interactive
+% cockpits that are now under
 
-We could mention this work when we introduce scenarios, and say that
-people inventing these new metaphors could take advantage of our
-metrics if they model 2+ apps.
+% application of a formal description technique to the various elements
+% of ARINC 661 specification within an industrial project. This formal
+% description technique called Interactive Cooperative Objects defines
+% in a precise and non-ambiguous way all the elements of ARINC 661
+% specification. The application of the formal description techniques is
+% shown on an interactive application called MPIA (Multi Purpose
+% Interactive Application). Within this application, we present how ICO
+% are used for describing interactive widgets, User Applications and
+% User Interface servers (in charge of interaction techniques). The
+% emphasis is put on the model-based management of the feel of the
+% applications allowing rapid prototyping of the external presentation
+% and the interaction techniques.
 
-Eg. they say: Requirement 2. Activity-based systems should provide
-lightweight mechanisms to create, change, and alter
-activities, since heavyweight interaction techniques are
-likely to deter adoption and use.
+% The Interactive Cooperative Objects (ICOs) formalism is a formal
+% description technique dedicated to the specification of interactive
+% systems [4, 11]. It uses concepts borrowed from the object-oriented
+% approach to describe the structural or static aspects of systems, and
+% uses high-level Petri nets [8] to describe their dynamic or
+% behavioural aspects.
 
-eg Requirement 7. Because information sharing is a
-“common case” in knowledge work, lightweight sharing
-capabilities should be integrated directly as a first-class
-interaction technique.
+% The Interactive Cooperative Objects (ICOs) formalism is a formal
+% description technique dedicated to the specification of interactive
+% systems [4, 11]. It uses concepts borrowed from the object-oriented
+% approach to describe the structural or static aspects of systems, and
+% uses high-level Petri nets [8] to describe their dynamic or
+% behavioural aspects.
+
+The issue of measuring UI/interaction properties based on state
+transition representations is not so well covered in the literature. A
+review of usability measuring practices~\cite{hornbaek2006current},
+under the headings \emph{measures of efficiency/usage patterns}, lists
+only number of keystrokes, mouse clicks, and visited objects as
+possible metrics. The notion of \emph{deviation from optimal solution}
+is discussed only in the context of tools for route planning in 3D
+navigation. In this paper we provide our own definition of deviation,
+which applies to interaction with the UI; we provide also our notion
+of execution traces across states of the UI, and the length of these
+traces can be used as a measure of efficiency.
+
+Lostness~\cite{smith1996lostness,otter2000lostness} is one of the few
+metrics that were defined to measure the degree to which users become
+lost in the information space, and therefore considers the notion of
+\emph{deviation}. Defined specifically for hypertext systems, lostness
+is a user performance measure that considers among others the number
+of visited nodes, the number of different nodes that were visited
+nodes, and the number of required nodes. This measure of efficiency is
+usually applied to traces of actual users, and is argued to be
+suitable for hypertext systems because the predominant task is
+browsing information, rather than trying to achieve specific goals. It
+therefore relaxes three assumptions: that there is a task to complete,
+that there is a correct way to carry it out, that the purpose of the
+system is to support users to carry out their tasks. The lostness
+metric can be easily computed on top of the data that we use with our
+approach; it would be yet another metric that can be provided to
+assess quality of a UI. Notice that in our case we do not need to user
+data to compute it.
+
+A usability analysis method capable to analytically predict task
+completion times from a storyboard of the UI is
+CogTool~\cite{bellamy2011cogtool}. CogTool relies on the ACT-R
+cognitive modeling engine, and allows a designer to setup a
+low-fidelity prototype of the UI. After deciding which interaction
+modality and which widgets are used to implement actions, the designer
+gets a validated estimation of how long an experienced user would
+spend on each step. CogTool takes care of adding extra ``mental''
+steps, such as ``think'' steps, before certain patterns of provided
+steps, according to the cognitive theory underlying ACT-R. As a
+result, users of CogTool obtain the breakdown of the times required by
+each of the steps. Our method is weaker in terms of cognitive validity
+and in terms of precision of the output: it does not provide expected
+completion times. However, with our method one can analyze a large
+part of the UI, get information about possible problems in some areas,
+and only then devolve more resources in building storyboards and
+in making assumptions regarding widgets so that specific execution
+paths previously identified can be analyzed with CogTool. In a sense,
+the output of our method could be used to make informed decisions as
+to what to analyze with CogTool.
+
+Markov models, \ie\ directed graphs where edges leaving a vertex are
+associated to a probability distribution, were used
+in~\cite{thimbleby2001markov}, as a means to perform usability
+analysis as early as possible, even before building prototypes of the
+UI. Vertices represent states of the UI and edges correspond to user
+actions (such as pressing a push-down button).  Probabilities can be
+used as a model of the skills or knowledge of the user: equiprobable
+actions correspond to a knowledge-free user, whereas when some actions
+have a very low probability it means that for that user the action is
+unlikely to be executed (perhaps because little known). Simple
+mathematical operations on the transition matrix of the model give the
+probability that after $n$ steps from a given initial state, the UI is
+in a given state. With Markov models, by manipulating probabilities,
+the designer can plot the number of required steps as a function of
+how close the probabilities are to the designer's ``perfect''
+knowledge.  Examples discussed in the paper cover several devices,
+ranging from a simple torch (with 4 states), a microwave cooker (6
+states), a mobile phone (152 states), but all being push-down devices
+with a fixed set of buttons. This is obviously not the case for UIs of
+information systems, where some buttons may or may not be ``painted''
+on the screen. This makes it more difficult to specify the transition
+matrix. Another limitation of this approach is that probabilities are
+assumed to be constant, corresponding to users that are not learning
+while interacting with the system.
+%
+Our approach is based on statecharts, a language that in practice is
+much more concise than Markov models, making it easier to specify the
+UI behavior, especially in cases where the set of buttons change over
+time. While our examples do not make use of probabilities, this is
+very easy to cope with (see the Discussion at the end of the paper for
+some of the benefits that doing this could bring). Similarly
+to~\cite{thimbleby2001markov}, our approach could be used when
+conservative results that do not rely on psychological assumptions are
+sought.
+
+centrality measures
+
+A discussion of social network analysis metrics applied to interaction
+design is provided in~\cite{thimbleby09:sna:eics09}. Once more, a UI
+is modeled in terms of directed graphs (vertices are states and edges
+are actions), and various centrality metrics are used to draw
+conclusions that bear upon usability. For example, centrality measures
+(such as the Sabidussi,  Jordan, or betweenness) can be used to
+identify states that are good places to start from to get to other
+states. Other metrics, such as edge betweenness, can be used to
+identify actions that are important because most of the shortest
+paths go through these actions. The paper presents compelling examples
+of using this technique to identify shortcomings in the design of
+infusion pumps. 
+%
+In our work we automatically generate graphs from statechart models,
+and tried to use these metrics. We were not able to draw sensible
+conclusions from the values we obtained (for example from the models
+presented below). One possible explanation may rest on the different
+types of models: in our case they reflect the variety and flexibility
+with which ``buttons'' can be used in modern web applications. For
+infusion pumps the UI is more constrained in how a task can be carried
+on.
+
+% Voida\cite{voida} says:
+% We describe an alternative model for organizing the
+% desktop interface-activity-based computing-and
+% identify a series of high-level system requirements for
+% interfaces that use activity as their primary organizing
+% principle.
+
+% We could mention this work when we introduce scenarios, and say that
+% people inventing these new metaphors could take advantage of our
+% metrics if they model 2+ apps.
+
+% Eg. they say: Requirement 2. Activity-based systems should provide
+% lightweight mechanisms to create, change, and alter
+% activities, since heavyweight interaction techniques are
+% likely to deter adoption and use.
+
+% eg Requirement 7. Because information sharing is a
+% “common case” in knowledge work, lightweight sharing
+% capabilities should be integrated directly as a first-class
+% interaction technique.
 
@@
 
+outline:
+- modeling, statecharts
+- flattening, ixgraph generation
+- scenario def
+- generation of traces
+- analysis
+
+modeling:
+- statecharts details
+
+flattening state machine \cite{briand2005:test:uml}
+-
+
 Notice that a scenario specifies only the desired occurrences of user
@@
 in order to view a conversation while reading another one, one has to
-\texttt{goBack} to the inbox first. It is the task of MIGTOOL to
+\texttt{goBack} to the inbox first. It is the task of \migtool to
 unfold cyles in the graph and compute all the  paths that connect the
@@
 
-feasibility:
+An important issue underlying \migtool is the modeling effort that is
+needed upfront. Our experience, based on several case studies and some
+industrial examples, is that models do not need to be complete
+representations of the behavior of the application under
+study. By following an agile modeling approach~\cite{ambler02}, models can
+be easily developed by one person in less than one day, using any UML
+capable design tool. Even more complex models (in our experience up to
+150 states and 350 transitions) can be developed and verified in 2-3
+days by one person.
+Experience in using statecharts to model behavior of UIs is needed
+though. Useful suggestions are given in~\cite{horrocks99,thimbleby07}.
 
-sensitivity with respect to models
+UML state diagrams provide a very expressive language, well suited to
+specify behavior of UIs based on descrete events. Evidence of this can
+also be found in recent OMG standards, such as
+IFML~\cite{ifml13}. Even though there are fundamental limits
+(inability to handle undo/redo's - because that goes beyond a finite
+state representation; inability to handle customizable toolbars -
+because that requires models that change at runtime; inability to
+handle perceptive UIs - because they are not well suited to be
+described in terms of descrete states), in many practical cases they
+can be be isolated and/or left aside).
 
-dire di horrocks \cite{horrocks99}
+Expressivity of the modeling language means that different designers
+are likely to produce different models for the same UI. As a
+consequence, metrics computed by \migtool do depend on different
+modeling choices. It is worth mentioning, though, that because of the
+flattening process, several differences are reduced (for example,
+those dealing with using a different hierarchy of states, or with
+distributing differently concurrent regions across states), and such a
+sensitivity is correspondingly reduced. At the moment, however, we do
+not have hard evidence to back this claim.
 
-dire di agile modeling \cite{ambler02}
+Differently from other model-based approaches, such as those based on
+IFML, \migtool uses \emph{only} a model of the behavior of the
+UI. Designers do not need to cope with data modeling, nor with
+decisions dealing with presentation. In a sense, \migtool uses only the
+\emph{controller} part of the Model-View-Controller paradigm that is
+adopted when developing UIs. Therefore, the designer using \migtool to
+perform analysis is free from other concerns that in the end affect
+usability, and the conclusions that are derived with \migtool can be
+combined with other results \emph{after} the analysis is
+performed. This also means that no effort needs to be directed on
+developing data and presentation specifications/implementations.
 
-\hl{dire} di SEPARATING concerns, ad es M-V-C, per vedere le
-  implicazioni di ciascuna e non soffrire di ``saturation'', vedi adams
+In terms of validity of conclusions obtained through \migtool, we can
+say that because they are devoid of user behavior assumptions (such as
+preferences, skills, interpretations, ergonomic constraints) they are
+very general. On the other hand, predictions based on \migtool metrics
+are also generic because they do not consider data and presentation
+aspects, nor user-related factors. For example, it is unfeasible to
+use \migtool to predict the time needed by a user to complete a
+scenario. However, \migtool can be used to analyze the whole
+interaction design and gather data to inform more specific analyses
+that could be performed, for example, with CogTool~\cite{john04cogtool}.
 
-valdiity of results:
-
-dire che le ns analisi sono prive di assunzioni su conoscenze
-abilita'; le probability sono puramente sulla base delle frequenze.
-
-dire di cogtools: noi non prevediamo i tempi (richiede modeling
-psicologico e dei specifici widgets). pero' forniamo un'analisi che
-puo' guidare chi applica cogtool.
-
-MIGTOOL is part of a suite of model-based tools for designing,
+\migtool is part of a suite of model-based tools for designing,
 analysing and testing user interfaces developed by the company
-ANONYMIZED. Development of a web-based user interface of MIGTOOL is
+ANONYMIZED. Development of a web-based user interface of \migtool is
 underway; the company plans to make it freely available for research

### 2015-09-24 paper.tex
completed "how it works" section

@@
 \newcommand{\eg}{\emph{e.g.}}
-\newcommand{\migtool}{MIG-Tool\ }
+\newcommand{\migtool}{MIGtool\ }
 \usepackage{soul}
@@
 
-\keywords{Experimental; Evaluation; Statecharts: UML; UML-IDEA; Testing.}
+\keywords{Experimental; Evaluation; Statecharts: UML; Metrics; Testing.}
 
@@
 
-In many cases \emph{statecharts} are used (a generalization of finite
-state automata, with an expressive language that includes hierarchic
-levels of abstraction, concurrent regions, states and pseudostates, an
-extended state notion based on an arbitrary computational model).
-Harel, the inventor of statecharts, gives an interesting retrospective
-view of how they were invented, back in early eighties, and why they
-were appealing also to non experts~\cite{harel09}.  For the sake of
-brevity, we refrain from describing them here in detail; the
-interested reader is referred to the UML standard~\cite{uml241} or
-some of the textbooks that deal with them, like~\cite{samek09}.
+In many cases \emph{statecharts} are used, a generalization of finite
+state automata, which is a more expressive language.
 %
@@
 %
-This idea was later on greatly expanded by
+This idea was later on expanded by
 Thimbleby~\cite{thimbleby07} , which discusses the use of statecharts
@@
 Our approach is based on statecharts, a language that in practice is
-much more concise than Markov models, making it easier to specify the
+more expressive than Markov models, making it easier to specify the
 UI behavior, especially in cases where the set of buttons change over
@@
 sought.
-
-centrality measures
 
@@
 
+vedere \cite{hilbert2000}
+\mytodo{However, because user interface
+  events are typically voluminos and rich in detail, automated support
+  is generally required to extract information at a level of
+  abstraction that is useful to investigators interested in analyzing
+  application usage or evaluating usability. This survey examines
+  computer-aided techniques used by HCI practitioners and researchers
+  to extract usability-related information from user interface
+  events. A framework is presented to help HCI practitioners and
+  researchers categorize and compare the approaches that have been, or
+  might fruitfully be, applied to this problem.
+
+  They - mention levels of abstractions in user events; what we deal
+  with here are ``abstract interaction level'' events - they mention
+  the problem of detecting sequences from event streams, which in a
+  sense is the converse of what we do - they discuss partial matches
+  between event streams and ``target'' sequences; we do the converse:
+  generation of event streams with detours from target sequences.  }
+
 % Voida\cite{voida} says:
@@
 
-\section{Generation of interaction paths}
+\section{Generation of interaction traces}
 \label{sec:traces}
 
-\begin{quote}
-  here we describe how we transform state machines into graphs, how we
-  specify scenarios, how the system computes traces, how metrics are
-  computed. 
-\end{quote}
-
-outline:
-- modeling, statecharts
-- flattening, ixgraph generation
-- scenario def
-- generation of traces
-- analysis
-
-modeling:
-- statecharts details
-
-flattening state machine \cite{briand2005:test:uml}
--
-
-Notice that a scenario specifies only the desired occurrences of user
-actions, not the necessary ones. For example, the model specifies that
-in order to view a conversation while reading another one, one has to
-\texttt{goBack} to the inbox first. It is the task of \migtool to
-unfold cyles in the graph and compute all the  paths that connect the
-desired user actions.
-
-Actions specified in a scenario are called \texttt{bridges}, because
-they represent one or more edges in the graph that have to be followed
-when carrying out the
-
-issues dealing with complexity of searching paths
+Interaction traces are paths (\ie, sequences of connected states
+of the model) that users can follow to perform a given scenario.  The
+generation process includes the following steps:
+\begin{inparaenum}[(1)]
+\item automatic flattening of the statechart model;
+\item manual definition of the interaction scenario;
+\item automatic generation of execution traces; and
+\item manual interactive analysis of results.
+\end{inparaenum}
+
+\subsection{Processing models }
+
+\migtool\ takes as input UML statecharts, which are a
+generalization of finite state automata (FSA), with an expressive language
+that includes hierarchic levels of abstraction, concurrent regions,
+states and pseudostates, an extended state notion based on an
+arbitrary computational model.  Harel, the inventor of statecharts,
+gives an interesting retrospective view of how they were invented,
+back in early eighties, and why they were appealing also to non
+experts~\cite{harel09}.  For the sake of brevity, we refrain from
+describing them here in detail; the interested reader is referred to
+the UML standard~\cite{uml25} or some of the textbooks that deal with
+them, like~\cite{samek09}.
+
+Using statecharts, behavior of UIs can be represented by associating
+states to screens and particular configuration of widgets, and
+transitions to actions performed by users or by the system itself.
+
+Because statechart models take advantage of abstraction features and a
+rich set of connecting pseudostates, they are not suitable to be
+directly processed to compute metrics. For this reason, \migtool first
+\emph{flattens} the model. Flattening is a process often used when
+statecharts have to be automatically
+processed~\cite{briand2005:test:uml}, and it means to produce a FSA
+that is behaviorally equivalent to the original statechart, where
+hierarchy between states and concurrent regions are removed. In most
+cases this leads to an exponential number of states and transitions,
+but because the process is completely automatic and in our case there
+is no need to manually inspect the resulting FSA, this aspect is in
+many cases not relevant.  Flattening statecharts is an often used
+process when they have to be automatically
+processed~\cite{briand2005:test:uml}.  \migtool\ produces an XML
+representation (graphML) of the resulting FSA, the \emph{interaction
+  graph}. It is a directed multigraph, potentially with cycles and
+loops, with edges that are labeled with the name of the corresponding
+action\footnote{A directed multigraph is a graph such that there is at
+  least a pair of vertices that are connected by 2 or more edges that
+  have the same direction. Cycles are paths that include 2 or more
+  occurrences of the same edge. Loops are edges that start and end on
+  the same vertex.}. In the following we will be referring to the
+interaction graph, and will be using as synonymous the terms ``state''
+and ``vertex'', and ``action'' and ``edge''.
+
+\subsection{Defining usage scenarios}
+
+Because in all but the most trivial interaction graphs there are
+cycles, the set of possible interaction traces is usually
+infinite. For this reason, scenarios need to be defined as constraints
+on the possible execution traces that can be generated.  Users of
+\migtool\ define interaction scenarios by specifying key steps (called
+\emph{bridge sets}) that users are expected to go through; we call
+this process \emph{grounding usage scenarios on models}. Each bridge
+set is specified by selecting a subset of the edges of the interaction
+graph. Often a scenario might require also optional steps. In such a
+case, the specification includes an initial state and a sequence of
+bridge sets. For example, to specify a scenario for replying to an
+email message, one could select all the edges associated to the action
+``reply'' (bridge set 1), followed by edges labeled with ``typeBody''
+(bridge set 2), followed by edges labeled with ``send'' (bridge set
+3). A well formed bridge set would consist of 1 or more edges (a
+bridge set with 0 edges would make the scenario unviable). In this way
+scenarios with cycles can be formulated (\eg, reply to two
+messages). Each pair of consecutive bridge sets makes a \emph{stage}
+of the scenario.
+%
+To cope with multi edges, the interaction graph is \emph{simplified}:
+all edges between  pairs of vertices are merged into a single one,
+whose label includes the original ones.
+
+\subsection{Searching traces}
+
+Quite expressive languages can be conceived for grounding scenarios
+(\eg, languages based on regular expressions). But such expressivity
+bonus needs to be balanced with computational tractability: even
+models with 50 states might lead to interaction graphs with several
+hundred states and several thousand edges, leading to an enormous
+number of possible paths to filter even for scenarios with just a few
+stages.
+
+To cope with this we implemented a path searching algorithm that
+processes each of the stages sequentially, starting from the initial
+initial state. Given a bridge set $B_i$, the algorithm does a
+breadth-first search of all the geodesic paths\footnote{``Geodesic''
+  is synonymous with ``shortest''.}  that connect each of the ending
+vertices of edges in the bridge set $B_i$ with some of the starting
+vertices of edges in bridge set $B_{i+1}$. If some of the starting
+vertices cannot be reached, then the bridge is dropped. \migtool\
+creates a new graph from the geodesic paths found for each stage, and
+then joins these graphs so that geodesic paths found for stage $i$ are
+joined with those of ${i+1}$. These global paths, connecting the
+initial state and the remaining bridges of the last bridge set of the
+scenario are called \emph{execution traces}.
+
+Notice that a scenario specifies only the desired occurrences of
+actions, not all the necessary ones. For example, if the model
+prescribes that in order to \texttt{view(message)} while reading
+another one, one has to \texttt{goBack} to the inbox first, a scenario
+specifying two consecutive \texttt{view(message)} would lead to
+geodesic paths that include also the \texttt{goBack} action, even if
+it is not specified in the scenario. It is the task of \migtool to
+unfold cyles in the graph and search all the geodesic paths that
+connect the desired user actions.
+
+\subsection{Detour traces}
+The algorithm described so far finds (some of) the global paths
+connecting the initial state to one or more bridges for each of the
+specified stages. The globally shortest path is identified, together
+with other viable alternatives. These traces are called
+\emph{detour order 0} traces (and states).
+
+For each stage, up to a maximum detour order $H$, the search algorithm
+creates traces with detour order $k+1$ by collecting states $D_k$
+with order $0$ up to $k$, and by identifying the neighbours ($N_k$)
+of $D_k$ (which are the states not included in $D_k$ that are
+connected through an edge to some state in $D_k$). Edges connecting
+$D_k$ with $N_k$ represent deviations that users might follow, and
+geodesic paths between $N_k$ and $D_k$ constitute recovery paths that
+users might follow to complete the scenario from states in
+$N_k$. These deviations and recovery paths are joined and constitute
+the traces of order $k+1$.  It could happen that
+for some state in $N_k$ there is no path leading to any vertex in
+$D_k$: in such a case the deviation leads to a dead end that prevents
+the user to complete the scenario.
+
+In general, \migtool is used to process a model against a scenario
+and to generate execution traces of detour order = 0,
+..., H. The time complexity of the search algorithm is
+$\mathcal{O}(KM(E+V)))$, where there are $K$ bridge sets, the mean
+size of them is $M$, the interaction graph has $V$ vertices and $E$
+edges. Therefore it scales pretty well with the size of the
+interaction graph and/or complexity of the scenarios. In practice for
+interaction graphs consisting of about 10000 edges and a dozen of
+bridge sets, on a low-cost PC it takes about 15 seconds to generate
+traces of order 0 to 3.
 
@@
 analysing and testing user interfaces developed by the company
-ANONYMIZED. Development of a web-based user interface of \migtool is
-underway; the company plans to make it freely available for research
+ANONYMIZED. \migtool is implemented partly in Java (model processing)
+and partly in R/iGraph. Development of a web-based UI of \migtool is
+underway; \migtool will be made freely available for research
 purposes.

### 2015-09-25 paper.tex
half way in the revision process

--- 2015-09-24 paper.tex
@@
 \newcommand{\eg}{\emph{e.g.}}
-\newcommand{\migtool}{MIG-Tool\ }
+\newcommand{\migtool}{MIGtool}
 \usepackage{soul}
@@
 \begin{abstract}
-  Early prototyping of user interfaces is an established good
-  practice. However, because prototypes cover only some usage
-  scenarios, the whole picture of the user interface is easily
-  lost.  Questions dealing with number of required steps,
-  possible interaction paths or impact of possible user errors can be
-  answered only for the specific scenarios being analysed and after
-  tedious manual inspection.
-
-  We present a tool that transforms models of a user interface into a
-  graph, upon which usage scenarios can be easily specified, and used
-  to compute possible interaction paths. Metrics based on possible
-  paths, with or without possible user errors, can be easily computed.
-  For example, when applying them to mail applications, we learn that
-  Gmail has 3 times more shortest routes, has 2x more routes that
-  include one user error, has routes with 13\% fewer steps, but has
-  the smallest probability to hit an optimal route.
+  Early prototyping of user interfaces is an established good practice
+  in interactive system development. However, prototypes cover only
+  some usage scenarios, and questions dealing with number of required
+  steps, possible interaction paths or impact of possible user errors
+  can be answered only for the specific scenarios and only after tedious
+  manual inspection.
+
+  We present a tool (\migtool) that transforms models of a user
+  interface into a graph, upon which usage scenarios can be easily
+  specified, and used to compute possible interaction paths. Metrics
+  based on possible paths, with or without user errors, can be easily
+  computed.  For example, when applying them to compare mail
+  applications, we show that Gmail has 3 times more shortest routes,
+  has 2x more routes that include a single user error, has routes with
+  13\% fewer steps, but optimal routes have the smallest probability
+  to be chosen.
 
@@
 
-\begin{quote}
-\mytodo{ Subcommittee: Technology, Systems, and Engineering
-
-People we should pay particular attention to:
-Caroline Appert
-Conversy, Stéphane
-Nebeling, Michael
-
-Keyword we need to pay attention to:
-This subcommittee will focus on technology, systems and engineering
-contributions that enable, improve, or advance interaction. This will
-include software and hardware technologies and systems that enable and
-demonstrate novel interactive capabilities, as well as languages,
-methods and tools for \hl{CONSTRUCTION AND ENGINEERING} of interactive
-systems. Engineering contributions should clearly demonstrate how they
-address interactive systems concerns such as, for example,
-scalability, reliability, interoperability, testing, and
-performance. Systems and technology contributions will be judged by
-their technical innovation and/or ability to connect, SIMPLIFY or
-enrich interactions, for example in intelligent interfaces and
-mobile/ubiquitous computing.
-}
-
-\end{quote}
+% \mytodo{ Subcommittee: Technology, Systems, and Engineering
+
+% People we should pay particular attention to:
+% Caroline Appert
+% Conversy, Stéphane
+% Nebeling, Michael
+
+% Keyword we need to pay attention to:
+% This subcommittee will focus on technology, systems and engineering
+% contributions that enable, improve, or advance interaction. This will
+% include software and hardware technologies and systems that enable and
+% demonstrate novel interactive capabilities, as well as languages,
+% methods and tools for \hl{CONSTRUCTION AND ENGINEERING} of interactive
+% systems. Engineering contributions should clearly demonstrate how they
+% address interactive systems concerns such as, for example,
+% scalability, reliability, interoperability, testing, and
+% performance. Systems and technology contributions will be judged by
+% their technical innovation and/or ability to connect, SIMPLIFY or
+% enrich interactions, for example in intelligent interfaces and
+% mobile/ubiquitous computing.
+% }
+% % 
+%\end{quote}
 
 We present an approach that allows a designer to assess interaction
-design (IxD) qualities such as efficiency,
-error-proneness and recovery from premature committment. Key
-importance is given to the ability to quickly
+design (IxD) qualities such as efficiency, error-proneness and
+recovery from errors. Key importance is given to the ability to
 \begin{inparaenum}[(a)]
@@
 \end{inparaenum}
+All this can be done before developing prototypes of the UI.
 
@@
 ends. We have chosen this domain because it is easy to understand and
-yet several trivial questions are not easily answered. We applied
-the same techniques also in other domains, such as HVAC and other
-embedded UIs.
+yet several questions cannot be easily answered. We applied the same
+techniques also in other domains, such as HVAC and other embedded UIs.
 
@@
 
-Even worse, there might be questions that cannot be easily
+Even worse, several questions cannot be easily
 answered. Given one or more potential designs and some usage
-scenarios, interesting questions could include: ``How many different
+scenarios, interesting questions include: ``How many different
 routes can be followed by the user to carry out the scenario?'',
@@
 he or she be able to recover?'', ``How many steps would the recovery
-require?''. In a different setting, when designing and evaluating
+require?''. When designing and evaluating
 embedded UIs (such as when dealing with plane's
-cockpits~\cite{conversy07}), other relevant questions might include
+cockpits~\cite{conversy07}), other relevant questions include
 ``How would the above properties change if we add a certain a
@@
 
-This should not be the case, however. Answers to these questions could
-be automatically computed. In this way a system could provide
-important insights to a designer, and  support assements of
+This should not be the case, however, because answers to these
+questions can be automatically computed. In this way a system can
+provide important insights to a designer, and support assements of
 potential user flexibility, user efficiency, error proneness, ability
@@
 graphs are then used to specify interaction scenarios and to unfold
-the possible interaction paths (called \emph{execution traces})
-determined by the specific scenarios being considered. Traces are fed
-to several graph-theoretic computations which produce a dashboard with
-different results. Except for development of models and specification
-of the desired scenarios, which have to be done manually, the other
-steps are totally automatic; models of a design (such as the ones
-shown below) can be developed in a matter of a couple of hours.
+the possible interaction paths (called \emph{execution traces}) that
+are compatible with the specific scenarios being considered. Traces
+are fed to graph-theoretic computations which produce a
+dashboard with different results. Except for development of models and
+specification of the desired scenarios, which have to be done
+manually, the other steps are totally automatic; models of a design
+(such as the ones shown below) can be developed in a matter of a
+couple of hours.
 
@@
 \begin{inparaenum}[(a)]
-\item the idea of using a model to specify interaction scenarios,
-\item the development of a tool (MIG-TOOL, Measuring Interaction
+\item the idea of using a UML state machine model to specify
+  interaction scenarios,
+\item the development of a tool (\migtool, Measuring Interaction
   Graphs Tool) that transforms models and scenario specifications into
@@
 \end{inparaenum}
-The examples included in the paper show that among four web mail
+The examples presented below show that among four web mail
 applications, and with respect to a typical usage scenario, Gmail is
@@
 slightly different interaction structures adopted for uploading
-messages. If Gmail didn't exist yet, by using \migtool its designers
+messages. If Gmail didn't exist yet, by using \migtool\ its designers
 could obtain these answers well before developing prototypes and
@@
 
+Other examples shown below show what is gained when a new feature is
+added to an embedded system.
+
 \section{Background}
@@
 
-The literature on using state-transition network representations for
-specifying or analysing the behavior of UIs is vast. We conducted a
-systematic-style literature review, using Google Scholar and queries
-with combinations of these phrases: ``user interface'', ``path
-analysis'', ``user trace'', ``interaction trace'', ``navigation'',
-``markov chain'', ``state transition'', ``statechart'', ``metric'',
+The literature on using state-transition networks for specifying or
+analysing the behavior of UIs is vast. We conducted a systematic-style
+literature review, using Google Scholar and queries with combinations
+of these phrases: ``user interface'', ``path analysis'', ``user
+trace'', ``interaction trace'', ``navigation'', ``markov chain'',
+``markov model'', ``state transition'', ``statechart'', ``metric'',
 ``measure''. For each query we analyzed title and abstract of the
@@
 discusses approaches for measuring user interfaces in the context of
-usability and on the basis of a model based on state transitions. This
-resulted in 78 full-text papers that were later on re-analyzed against
-the same criterion, leading to several of papers mentioned below.
+usability and on the basis of a state transition model. This resulted
+in 78 full-text papers that were later on re-analyzed against the same
+criterion, leading to several of papers mentioned below.
 %
-Several of the examined papers deal with testing user interfaces, and
-problems related to generating execution traces as a means to assess
-coverage of a test
+Many papers deal with testing user interfaces, and problems related to
+generating execution traces as a means to assess coverage of a test
 suite~\cite{ammann2008introduction,briand2005:test:uml}. They were
@@
 
-In many cases \emph{statecharts} are used (a generalization of finite
-state automata, with an expressive language that includes hierarchic
-levels of abstraction, concurrent regions, states and pseudostates, an
-extended state notion based on an arbitrary computational model).
-Harel, the inventor of statecharts, gives an interesting retrospective
-view of how they were invented, back in early eighties, and why they
-were appealing also to non experts~\cite{harel09}.  For the sake of
-brevity, we refrain from describing them here in detail; the
-interested reader is referred to the UML standard~\cite{uml241} or
-some of the textbooks that deal with them, like~\cite{samek09}.
+In many cases \emph{statecharts} are used, a generalization of finite
+state automata.  Horrocks showed how statecharts can be used to model
+and specify the dynamics of typical desktop
+UIs~\cite{horrocks99}. While providing many interesting insights on
+how and why one should use statecharts to do so, this nice work does
+not address how such a specification could be \emph{automatically}
+processed.
 %
-Horrocks showed how statecharts can be  used to model and
-specify the dynamics of typical desktop UIs~\cite{horrocks99}. While
-providing many interesting insights on how and why one should use
-statecharts to do so, this nice work does not address how such a
-specification could be \emph{automatically} processed.
-%
-This idea was later on greatly expanded by
-Thimbleby~\cite{thimbleby07} , which discusses the use of statecharts
-as meta-models for user interfaces; they are seen as crucial
-representations that allow a designer to fully appreciate how devices
-behave. The overall claim is that ``If you don't understand the logic
-conveyed by a statechart model of a user interface, then you don't
-understand the behavior of the user interface''.
+This idea was later on expanded by Thimbleby~\cite{thimbleby07};
+statecharts are seen as crucial representations that allow a designer to
+fully appreciate how devices behave. The overall claim is that ``If
+you don't understand the logic conveyed by a statechart model of a
+user interface, then you don't understand the behavior of the user
+interface''.
 
@@
 (to display details of a item, to display or select lists of items, to
-input an item).
+input an item). None of these approaches focus directly on measures of
+usability.
 
 A rather different route for the problem of generating UIs is followed
-in~\cite{heymann2007formal}. They assume that the UI to be generated
+in~\cite{heymann2007formal}. Authors assume that the UI to be generated
 is used to supervise and to monitor an underlying machine (\eg,
 autopilot of a plane) which is modeled as a statechart. After assuming
-that the behavior of the UI can be modeled as a statechart, they
+that the behavior of the UI can also be modeled as a statechart, they
 devised an algorithm that checks whether the two models are
@@
 formalism.  It is used to enrich the ARINC 661 specification of
-Cockpit Display System used in interactive cockpits of airplanes, so
-that the semantics of widgets can be expressed and the behavior of a
-UI be formally analyzed.
+Cockpit Display System so that the semantics of widgets can be
+expressed and the behavior of a UI be formally analyzed.
 
@@
 Finite state representations have been used also as a conceptual
-framework to write the code of widgets so that events and event
+framework for writing the code of widgets so that events and event
 handlers in the UI can more easily be conceived, developed and
@@
 
-The issue of measuring UI/interaction properties based on state
-transition representations is not so well covered in the literature. A
-review of usability measuring practices~\cite{hornbaek2006current},
-under the headings \emph{measures of efficiency/usage patterns}, lists
-only number of keystrokes, mouse clicks, and visited objects as
-possible metrics. The notion of \emph{deviation from optimal solution}
-is discussed only in the context of tools for route planning in 3D
-navigation. In this paper we provide our own definition of deviation,
-which applies to interaction with the UI; we provide also our notion
-of execution traces across states of the UI, and the length of these
-traces can be used as a measure of efficiency.
+A review of usability measuring practices~\cite{hornbaek2006current}
+lists, under the headings \emph{measures of efficiency/usage
+  patterns}, number of keystrokes, mouse clicks, and visited objects
+as possible metrics. The notion of \emph{deviation from optimal
+  solution} is discussed only in the context of tools for route
+planning in 3D navigation. In this paper we provide our own definition
+of deviation, which applies to interaction with the UI; we provide
+also our notion of execution traces across states of the UI, and the
+length of these traces can be used as a measure of efficiency.
 
 Lostness~\cite{smith1996lostness,otter2000lostness} is one of the few
-metrics that were defined to measure the degree to which users become
-lost in the information space, and therefore considers the notion of
+metrics for measuring the degree to which users become lost in the
+information space, and therefore considers the notion of
 \emph{deviation}. Defined specifically for hypertext systems, lostness
-is a user performance measure that considers among others the number
-of visited nodes, the number of different nodes that were visited
-nodes, and the number of required nodes. This measure of efficiency is
+is a user performance measure which is afuncion of the number of
+visited nodes, the number of different nodes that were visited nodes,
+and the number of required nodes. This measure of efficiency is
 usually applied to traces of actual users, and is argued to be
@@
 browsing information, rather than trying to achieve specific goals. It
-therefore relaxes three assumptions: that there is a task to complete,
-that there is a correct way to carry it out, that the purpose of the
-system is to support users to carry out their tasks. The lostness
-metric can be easily computed on top of the data that we use with our
-approach; it would be yet another metric that can be provided to
-assess quality of a UI. Notice that in our case we do not need to user
-data to compute it.
+is suitable therefore whe some of the following three assumptions can
+be relaxed: that there is a task to complete, that there is a correct
+way to carry it out, that the purpose of the system is to support
+users to carry out their tasks.
 
@@
 modality and which widgets are used to implement actions, the designer
-gets a validated estimation of how long an experienced user would
+gets an estimation of how long an experienced user would
 spend on each step. CogTool takes care of adding extra ``mental''
@@
 actions (such as pressing a push-down button).  Probabilities can be
-used as a model of the skills or knowledge of the user: equiprobable
-actions correspond to a knowledge-free user, whereas when some actions
-have a very low probability it means that for that user the action is
-unlikely to be executed (perhaps because little known). Simple
-mathematical operations on the transition matrix of the model give the
-probability that after $n$ steps from a given initial state, the UI is
-in a given state. With Markov models, by manipulating probabilities,
-the designer can plot the number of required steps as a function of
-how close the probabilities are to the designer's ``perfect''
-knowledge.  Examples discussed in the paper cover several devices,
-ranging from a simple torch (with 4 states), a microwave cooker (6
-states), a mobile phone (152 states), but all being push-down devices
-with a fixed set of buttons. This is obviously not the case for UIs of
-information systems, where some buttons may or may not be ``painted''
-on the screen. This makes it more difficult to specify the transition
-matrix. Another limitation of this approach is that probabilities are
-assumed to be constant, corresponding to users that are not learning
-while interacting with the system.
+used as a model of user knowledge: equiprobable actions correspond to
+a knowledge-free user, whereas when some actions have a very low
+probability it means that for that user the action is unlikely to be
+executed. Simple mathematical operations on the transition matrix of
+the model give the probability that after $n$ steps from a given
+initial state the UI is in a given state. With Markov models, by
+manipulating probabilities, the designer can plot the number of
+required steps as a function of how close the probabilities are to the
+designer's ``perfect'' knowledge.  Examples discussed in the paper
+cover several devices, ranging from a simple torch (with 4 states), a
+microwave cooker (6 states), a mobile phone (152 states); those are
+all push-down devices with a fixed set of buttons. This is obviously
+not the case for UIs of information systems, where buttons may change
+screen by screen. This makes it more difficult to specify the
+transition matrix. 
 %
 Our approach is based on statecharts, a language that in practice is
-much more concise than Markov models, making it easier to specify the
-UI behavior, especially in cases where the set of buttons change over
+more powerful than Markov models, making it easier to specify the UI
+behavior, especially in cases where the set of buttons change over
 time. While our examples do not make use of probabilities, this is
@@
 
-centrality measures
-
 A discussion of social network analysis metrics applied to interaction
@@
 conclusions that bear upon usability. For example, centrality measures
-(such as the Sabidussi,  Jordan, or betweenness) can be used to
-identify states that are good places to start from to get to other
+(such as Sabidussi, eccentricity, betweenness) can be used to identify
+states that are good places to start from to get to other
 states. Other metrics, such as edge betweenness, can be used to
-identify actions that are important because most of the shortest
-paths go through these actions. The paper presents compelling examples
-of using this technique to identify shortcomings in the design of
+identify actions that are important because most of the shortest paths
+go through these actions. The paper presents compelling examples of
+using this technique to identify shortcomings in the design of
 infusion pumps.
@@
 In our work we automatically generate graphs from statechart models,
-and tried to use these metrics. We were not able to draw sensible
-conclusions from the values we obtained (for example from the models
-presented below). One possible explanation may rest on the different
-types of models: in our case they reflect the variety and flexibility
-with which ``buttons'' can be used in modern web applications. For
-infusion pumps the UI is more constrained in how a task can be carried
-on.
+and on some of them computed these metrics. We were not able to draw
+sensible conclusions from the values we obtained (for example from the
+models presented below). One possible explanation rests on the
+different types of models: in our case they reflect the variety and
+flexibility with which ``buttons'' can be used in modern web
+applications. For infusion pumps the UI is more constrained in how a
+task can be carried on, and this difference might reflect on the
+usefulness of those metrics.
+
+In~\cite{hilbert2000} several approaches to analyse streams of user
+events are discussed and compared. It is interesting to realize that
+this is, in a sense, the inverse problem of the one we tackle in this
+paper: we want to compute a subset of the possible streams of user
+events given a specification of the UI, rather than trying to abstract
+general properties from streams of events.
 
@@
 
-\section{Generation of interaction paths}
+\section{Generation of interaction traces}
 \label{sec:traces}
 
-\begin{quote}
-  here we describe how we transform state machines into graphs, how we
-  specify scenarios, how the system computes traces, how metrics are
-  computed. 
-\end{quote}
-
-outline:
-- modeling, statecharts
-- flattening, ixgraph generation
-- scenario def
-- generation of traces
-- analysis
-
-modeling:
-- statecharts details
-
-flattening state machine \cite{briand2005:test:uml}
--
-
-Notice that a scenario specifies only the desired occurrences of user
-actions, not the necessary ones. For example, the model specifies that
-in order to view a conversation while reading another one, one has to
-\texttt{goBack} to the inbox first. It is the task of \migtool to
-unfold cyles in the graph and compute all the  paths that connect the
-desired user actions.
-
-Actions specified in a scenario are called \texttt{bridges}, because
-they represent one or more edges in the graph that have to be followed
-when carrying out the
-
-issues dealing with complexity of searching paths
+In this section we provide a description of how \migtool\ computes
+execution traces.  They are paths (\ie, sequences of connected states
+of the model) that users can follow when performing a given scenario.
+The generation process encompasses the following steps:
+\begin{inparaenum}[(1)]
+\item automatic flattening of the statechart model;
+\item manual definition of the interaction scenario;
+\item automatic generation of execution traces; and
+\item interactive analysis of results.
+\end{inparaenum}
+
+\subsection{Processing models }
+
+\migtool\ takes as input UML statecharts, which are a generalization
+of finite state automata (FSA), with an expressive language that
+includes hierarchic levels of abstraction, concurrent regions, states
+and pseudostates, guards, and an extended state notion based on an
+arbitrary underlying computational model.  Harel, the inventor of
+statecharts, gives an interesting retrospective view of how they were
+invented, back in early eighties, and why they were appealing also to
+non experts~\cite{harel09}.  For the sake of brevity, we refrain from
+describing them here in detail; the interested reader is referred to
+the UML standard~\cite{uml25} or some of the textbooks that deal with
+them, like~\cite{samek09}.
+
+By using statecharts, the behavior of UIs can be represented by
+associating states to screens and particular configurations of
+widgets, and transitions to actions performed by users or by the
+system itself. In the classification reported in~\cite{hilbert2000},
+actions belong to the ``abstract interaction events'' category.
+
+Because statechart models take advantage of abstraction features and a
+rich set of connecting pseudostates, they are not suitable to be
+directly processed to compute metrics. For this reason, \migtool\
+first \emph{flattens} the model. Flattening is a process often used
+when statecharts have to be automatically
+processed~\cite{briand2005:test:uml}, and it means to produce a FSA
+that is behaviorally equivalent to the original statechart, with no
+hierarchy between states and no concurrent regions. In most cases this
+leads to an exponential number of states and transitions in the FSA,
+but because the process is completely automatic and there is no need
+to manually inspect the resulting FSA, this aspect is in many cases
+not relevant.  \migtool\ produces an XML representation (graphML) of
+the resulting FSA, the \emph{interaction graph}. It is a directed
+multigraph\footnote{A directed multigraph is a graph such that there
+  are 2 or more edges that have the same end points. Cycles are paths
+  that include 2 or more occurrences of the same vertex. Loops are
+  edges that start and end on the same vertex. ``Geodesic path'' is a
+  synonymous term with ``shortest path''.}, potentially with
+cycles and loops, with edges that are labeled with the name of the
+corresponding action. In the following we will be using as synonymous
+the terms state/vertex, and action/edge.
+
+\subsection{Defining usage scenarios}
+
+Because in all but the most trivial interaction graphs there are
+cycles, the set of possible interaction traces is infinite. For this
+reason, scenarios need to be defined and used as constraints on the
+possible execution traces that can be generated.  Users of \migtool\
+define interaction scenarios by specifying key steps (called
+\emph{bridge sets}) that users are expected to go through; we call
+this process \emph{grounding usage scenarios on models}. Each bridge
+set is specified by selecting a subset of the edges of the interaction
+graph. In general a specification includes an initial state and a
+sequence of bridge sets. For example, to specify a scenario for
+replying to an email message, one could select all the edges
+associated to the action \texttt{reply} (bridge set 1), followed by
+edges labeled with \texttt{typeBody} (bridge set 2), followed by edges
+labeled with \texttt{send} (bridge set 3). A well formed bridge set
+consists of 1 or more edges (an empty bridge set would make the
+scenario unviable). In this way scenarios with cycles can be
+formulated (\eg, reply twice to two messages). A \emph{stage} of a
+scenario comprises two consecutive bridge sets.
+%
+To cope with multi edges, the interaction graph is \emph{simplified}:
+all edges between a   pair of vertices are merged into a single one,
+whose label includes the original ones.
+
+\subsection{Searching traces}
+
+Quite expressive languages can be conceived for grounding scenarios
+(\eg, regular expressions on sequences of action names). But such
+expressivity bonus needs to be balanced with computational
+tractability: even models with a dozen states might correspond to interaction
+graphs with several hundred states and several thousand edges, leading
+to an enormous number of possible traces to filter even for scenarios
+with just a few stages.
+
+To cope with this we implemented a trace searching algorithm that
+processes each of the stages sequentially, starting from the initial
+initial state. Given a bridge set $B_i$, the algorithm does a
+breadth-first search of all the geodesic paths that connect
+\emph{each} of the ending vertices of edges in $B_i$ with \emph{some}
+of the starting vertices of edges in $B_{i+1}$. If some bridge in
+$B_{i+1}$ cannot be reached, then it is dropped from further
+searches. \migtool\ creates a new graph from the geodesic paths found
+for each stage, and then joins these graphs so that geodesic paths
+found for stage $i$ are joined with those of ${i+1}$. These global
+paths, connecting the initial state to reachable bridges of the
+last bridge set of the scenario are called \emph{execution traces}.
+
+Notice that a scenario specifies only the desired occurrences of
+actions, not all the necessary ones. For example, if the model
+prescribes that in order to \texttt{view(message)} while reading
+another one, one has to \texttt{goBack} to the inbox first, a scenario
+specifying two consecutive \texttt{view(message)} would lead to traces
+that include also the \texttt{goBack} action, even if that step is not
+explicitly specified in the scenario. It is the task of \migtool\ to
+unfold paths in the graph and search all the geodesic paths that
+connect the desired user actions.
+
+\subsection{Detour traces}
+The algorithm described so far finds (some of) the global paths
+connecting the initial state to one or more bridges for each of the
+specified stages. The globally shortest paths are included in the
+solution, together with other viable alternatives. These traces are
+called \emph{detour order 0} traces.
+
+\migtool\ is used to process a model and a scenario
+and to generate execution traces of detour order = 0,
+..., H. 
+For each stage, up to a maximum detour order $H$, the search algorithm
+creates traces with detour order $k+1$ by collecting the set $D_k$ of states
+with order $0$ up to $k$, and by identifying the neighbours ($N_k$)
+of $D_k$ ($N_k$ is the set of states  not included in $D_k$ that are
+connected through an edge to some state in $D_k$). Edges connecting
+$D_k$ with $N_k$ represent deviations that users might follow, and
+geodesic paths between $N_k$ and $D_k$ constitute recovery paths that
+users might follow to complete the scenario from states in
+$N_k$. These deviations and recovery paths are joined and constitute
+the traces of order $k+1$.  It could happen that
+for some state in $N_k$ there is no path leading to any vertex in
+$D_k$: in such a case the deviation leads to a dead end that prevents
+the user to complete the scenario.
+
+The time complexity of the search algorithm is
+$\mathcal{O}(KM(E+V)))$, where there are $K$ bridge sets, their mean
+size is $M$, the interaction graph has $V$ vertices and $E$
+edges. Therefore it scales well with the size of the
+interaction graph and/or complexity of the scenarios. In practice for
+interaction graphs consisting of about 10000 edges and a dozen of
+bridge sets, on a low-cost PC it takes about 20 seconds to generate
+traces of order 0 to 3.
 
@@
 
-In this section we describe some examples based on well known web mail
-clients, namely Gmail, Horde, SquirrelMail and Roundcube. We chose
-these examples because they are very well known, and therefore are
-easy to describe and understand. And yet, despite email being a very
-well understood domain, the kind of questions that can be posed and
-the answers that are found provide an insight on some of the usability
-properties of these applications.
+In this section we describe some examples of the results that can be
+obtained with \migtool, based on well known web mail clients, namely
+Gmail, Horde, SquirrelMail and Roundcube. We chose these examples
+because they are well known, and therefore are easy to describe and
+understand. And yet, despite email being a very well understood
+domain, the kind of questions that can be posed and the answers that
+are found provide interesting insights on some of the usability properties of
+these applications.
 
@@
 
-Models of the four applications have been manually developed using UML
-state machines. In order to support a fair comparison, all four models
-cover the same set of use cases: listing the content of the inbox,
-reading a message or conversation, replying to a message, composing
-and sending a new message.
+Models of the four applications have been manually defined. In order
+to support a fair comparison, all four models cover the same set of
+use cases: listing the content of the inbox, reading a message or
+conversation, replying to a message, composing and sending a new
+message.
 
 Figure~\ref{fig:model} shows part of the model of Gmail. At some
-point, the user might be viewing all the conversations of the inbox
-(state \texttt{viewingConversations}); available actions include
-moving to the next or previuous block of conversations, refreshing the
-list, or opening a specific conversation (transition
+point, a user is viewing all the conversations of the inbox (state
+\texttt{viewingConversations}); available actions include moving to
+the next or previuous block of conversations, refreshing the list, or
+opening a specific conversation (transition
 \texttt{open(conversation)}). This transition (which is assumed to
@@
 action the UI moves to a state called \texttt{replying}, where the
-body of the message can be typed, the subject can be changed, or
-another addressee can be added.
+body of the reply can be typed, the subject can be changed, or another
+addressee can be added.
 
@@
 The model that we show here is part of what we used in the examples
-reported below, and that is only a part of what the real Gmail
-application supports. The actual model that we used consists of 
+reported below. The actual model that we used consists of 
 24 states, 12 pseudo states, 12 regions, and 61 transitions.
@@
 
-An important issue underlying \migtool is the modeling effort that is
+An important issue underlying \migtool\ is the modeling effort that is
 needed upfront. Our experience, based on several case studies and some
@@
 are likely to produce different models for the same UI. As a
-consequence, metrics computed by \migtool do depend on different
+consequence, metrics computed by \migtool\ do depend on different
 modeling choices. It is worth mentioning, though, that because of the
@@
 Differently from other model-based approaches, such as those based on
-IFML, \migtool uses \emph{only} a model of the behavior of the
+IFML, \migtool\ uses \emph{only} a model of the behavior of the
 UI. Designers do not need to cope with data modeling, nor with
-decisions dealing with presentation. In a sense, \migtool uses only the
+decisions dealing with presentation. In a sense, \migtool\ uses only the
 \emph{controller} part of the Model-View-Controller paradigm that is
-adopted when developing UIs. Therefore, the designer using \migtool to
+adopted when developing UIs. Therefore, the designer using \migtool\ to
 perform analysis is free from other concerns that in the end affect
-usability, and the conclusions that are derived with \migtool can be
+usability, and the conclusions that are derived with \migtool\ can be
 combined with other results \emph{after} the analysis is
@@
 preferences, skills, interpretations, ergonomic constraints) they are
-very general. On the other hand, predictions based on \migtool metrics
+very general. On the other hand, predictions based on \migtool\ metrics
 are also generic because they do not consider data and presentation
 aspects, nor user-related factors. For example, it is unfeasible to
-use \migtool to predict the time needed by a user to complete a
-scenario. However, \migtool can be used to analyze the whole
+use \migtool\ to predict the time needed by a user to complete a
+scenario. However, \migtool\ can be used to analyze the whole
 interaction design and gather data to inform more specific analyses
@@
 
-\migtool is part of a suite of model-based tools for designing,
+\migtool\ is part of a suite of model-based tools for designing,
 analysing and testing user interfaces developed by the company
-ANONYMIZED. Development of a web-based user interface of \migtool is
-underway; the company plans to make it freely available for research
-purposes.
+ANONYMIZED. \migtool\ is implemented partly in Java (model processing)
+and partly in R/iGraph. Development of a web-based UI of \migtool\ is
+underway; it will be freely available for research purposes.
 

### 2015-09-25 paper.tex
another step

@@
   \centering
-  \includegraphics[width=\linewidth]{figures/gmail_2.png}
+  \includegraphics[width=1.05\linewidth]{figures/gmail_2.png}
   \caption{Part of the Gmail model.}\label{fig:model}
@@
 
-The flattening process, which takes a fraction of a second on a
-low-cost PC, produces a directed multigraph with 47 vertices and 634
-edges; when simplified by collapsing multiple edges between any pair of
-vertices, the graph includes 312 edges. Each vertex represents one of
-the possible combinations of simple states in any of the regions that
-can be active at the same time.
+The flattening process, which takes a couple seconds on a low-cost PC,
+produces a directed multigraph with 47 vertices and 634 edges; when
+simplified by collapsing multiple edges, the number of edges drops to
+312. Each vertex represents one of the possible combinations of simple
+states in any of the regions that can be active at the same time.
 
@@
 Inspection of the interaction graph is not particularly useful because
-even for small graphs like the one  obtained from our Gmail model
-no particular structure is evident. Figure~\ref{fig:gom} shows a plot
-using a circular layout of the 47 states: because of the large
-number of cycles that exist among states, edges form an intricated web
-of possible action sequences. This in practice greatly reduces usefulness of
-typical network analysis metrics, such as \emph{betweenness,
-  eccentricity, page-rank, eigenvalue} centrality measures.
+even for small graphs like the one obtained from our Gmail model no
+particular structure is evident that was not known from the
+model. Figure~\ref{fig:gom} shows a plot using a circular layout of
+the 47 states: because of the large number of cycles that exist among
+states, edges form an intricated web of possible action sequences. We
+believe this in practice greatly reduces usefulness of typical network
+analysis metrics, such as \emph{betweenness, eccentricity, page-rank,
+  eigenvalue} centrality measures.
 
@@
   \includegraphics[width=\linewidth]{figures/gom.png}
-  \caption{Graph obtained from the Gmail model.}
+  \caption{Interaction graph obtained from the Gmail model.}
   \label{fig:gom}
@@
 To be able to obtain results that bear upon usability, we process
-further the graph, first by specifying an interaction scenario using
-vertices and edges of the graph, and secondly by computing possible
-execution paths.
-
-Specification of a scenario consists of formulating it using the
-language provided by the graph, i.e. deciding which is the initial
-state and which user actions need to be considered. For example, the
-following fragment of R code specifies that the scenario should
-include, in the given order, two occurrences of
+further the graph, by specifying scenarios and computing traces.  The
+following fragment of R code specifies the bridge sets of the choosen
+scenario: it comprises, in the given order, two occurrences of
 \texttt{open(conversation)}, followed by \texttt{reply}, followed by
@@
 \end{lstlisting}
-In plain language such a scenario means opening one and then a second conversation,
-replying to the last message of the second one by typing the body of
-the response, sending it, and then composing a new message by adding a
-recipient, an attachment, typing the body and subject, and finally
-sending it.
+In plain language such a scenario means opening a conversation,
+opening a second one, replying to the last message of the
+second conversation by typing the body of the response, sending it, and then
+composing a new message by adding a recipient, an attachment, typing
+the body and subject, and finally sending it.
 
@@
 
-Gmail has the highest number of order 0 and 1 paths, and the highest
-difference between the number of order 0 and 1, and between order 1
-and 2 (at least 3 times as many order 0 paths than any of the other
-applications, and at least twice as many order 1 paths as the any of
-the other ones). 
- Gmail offers 3 times as many error-free alternative
-paths to accomplish the scenario, which indicates that users might
-more easily follow one of those paths. 
-\mytodo{result}
-However, Gmail offers also
-almost twice as many order 1 paths, which means that users could be
-more easily induced into an erroneous path than when using another
-system.  
-\mytodo{result}
-Because the number of order 2 or 3 paths decreases, Gmail
-reduces therefore the ``error proneness'' of this UI, for executions
-that involve 2 or 3 errors. Notice that Roundcube has the smallest
-number of order 0 paths (2 of them), which means that users are not
-given much flexibility and freedom in carrying out correctly the
-scenario. 
-\mytodo{result}
-For none of the systems a detour leads to dead-ends preventing the
-user to complete the scenario.
+Gmail has the highest number of order 0 and 1 traces; it has the
+highest difference between the number of order 0 and 1, and between
+order 1 and 2 (at least 3 times as many order 0 paths than any of the
+other applications, and at least twice as many order 1 paths as the
+any of the other applications).  In other words, Gmail offers 3 times
+as many error-free alternative paths to accomplish the scenario, which
+indicates that users might more easily follow one of those paths, than
+when using other applications.  Notice that Roundcube has the smallest number
+of order 0 paths (2 of them), which means that users are not given
+much flexibility and freedom in carrying out correctly the scenario.
+However, Gmail provides also almost twice as many order 1 paths, which
+means that users could be more easily induced into an erroneous path
+than when using another system (or given more flexibility).
+Because the number of order 2 or 3 paths decreases, Gmail reduces
+therefore the ``error proneness'' of this UI, for executions that
+involve 2 or 3 errors.
+
+For none of the systems a detour leads to dead-ends.
 
@@
 error-free paths, but also gives the shortest ones. Users are given
-more flexibility and more efficiency.  \mytodo{result} Because also
-paths with order 1 or more are the shortest ones among the four
-applications, Gmail also makes users more efficient in recovering from
-errors.  \mytodo{result}
+more flexibility and more efficiency.  Because also paths with order 1
+or more are the shortest ones among the four applications, Gmail also
+makes users more efficient in recovering from errors.  \mytodo{result}
 
@@
 To combine these two results, we can easily compute the frequency of
-paths with different length. Figure~\ref{fig:freqopt} shows, for each
-application, the frequency of order 0 paths, the frequency of paths
-with length less or equal to 15 (the minimum length across the four
-applications), and the frequency of optimal paths (the shortest
+paths having different length. Figure~\ref{fig:freqopt} shows, for
+each application, the frequency of order 0 paths, the frequency of
+paths with length less or equal to 15 (the minimum length across the
+four applications), and the frequency of optimal paths (the shortest
 ones). These values show that Gmail users have the lowest probability
 to hit an order 0 path (because of the relative large number of order
-1 paths made available by Gmail), have the highest probability to hit
-a path with length 15 or less, have the lowest probability of hitting
-the shortest paths (10 times smaller than the best of the other
-applications).  Thus, the flexibility and efficiency that can be
-exploited with Gmail are counterbalanced by the required knowledge and
-capability of chosing an optimal path. In particular, Gmail offers
-many detours of order 1 which increase flexibility for some users and
-might decrease effectiveness for less knowledgeable ones.
-\mytodo{result}
+1 paths made available by Gmail), have the lowest probability of
+hitting the shortest paths (10 times smaller than the best of the
+other applications), but have the highest probability to hit a
+path with length 15 or less.  Thus, the flexibility and efficiency
+that can be exploited with Gmail are counterbalanced by the required
+knowledge and capability of chosing an optimal path. In particular,
+Gmail offers many detours of order 1 which increase flexibility for
+some users and might decrease effectiveness for less knowledgeable
+ones.
 
@@
 Another probabilistic analysis can be performed using page rank, which
-can be used to compute the probability that a random walk in a graph
-visits a certain vertex~\cite{page1999pagerank}. We computed the page rank (with a damping
-value of 5\% - meaning that the random walker with probability 5\%
-jumps to an arbitrary state and probability 95\% chooses one of the
-actions available in the current state) for each vertex in the graph,
-and then added the page rank value for vertices with different detour order.
-Figure~\ref{fig:pr} shows the resulting probabilities.
+computes the probability that a random walk in a graph visits a
+certain vertex~\cite{page1999pagerank}. We computed the page rank
+(with a damping value of 5\% - meaning that the random walker with
+probability 5\% jumps to an arbitrary state and probability 95\%
+chooses one of the actions available in the current state) for each
+vertex in the graph, and then summed up page rank values for vertices
+with different detour order.  Figure~\ref{fig:pr} shows the resulting
+sums.
 
@@
 visiting an order 0 or 1 state, the probability increases to 87\%,
-which is the highest. Thus, to compare Gmail with SquirrelMail, a
-completely random usage of SquirrelMail has 10\% more probability of
-hitting an error-free state than Gmail. That advantage is reduced when
-considering order 0 and 1 paths, because with SquirrelMail the
-probability is 85\% and Gmail it is 87\%.  This means that somebody
-with no knowledge on how to use an email front end, when using Gmail
-would have 10\% fewer chances of carrying out the scenario without
-making any error, as opposed to when using SquirrelMail: SquirrelMail
-provides more guidance. Across the four UIs, the probability of making
-at most 1 error is approximately the same\footnote{These results are
-  not very sensitive to the value chosen for the damping factor:
-  differences across the four applications are very similar when $d$
-  ranges in $[0.05,0.20]$.}.
-
-\mytodo{result}
+which is the highest. Thus, in a comparison of Gmail against
+SquirrelMail, a completely random usage of SquirrelMail has 10\% more
+probability of hitting an error-free state than Gmail. That advantage
+is slightly reduced when considering order 0 and 1 paths, because with
+SquirrelMail the probability is 85\% and Gmail it is 87\%.  This means
+that somebody with no knowledge on how to use an email front end, when
+using Gmail would have 10\% fewer chances of carrying out the scenario
+without making any error, as opposed to when using SquirrelMail:
+SquirrelMail provides more guidance. Across the four UIs, the
+probability of making at most 1 error is approximately the
+same\footnote{These results are not very sensitive to the value chosen
+  for the damping factor: differences across the four applications are
+  stable when $d$ ranges in $[0.05,0.20]$.}.
 
@@
 fact that users can start composing a new message while reading a
-conversation, whereas in other applications an explicit ``closing'' of
-the reading activity has to be performed. Another reason is in the
+conversation, whereas in other applications an explicit ``close action'' of
+the reading activity has to be performed. Another reason relies on the
 more streamlined process to attach a file: in Gmail one needs to
@@
 after selecting them.
-
-\mytodo{result}
 
@@
 Figure~\ref{fig:ad} shows the \emph{action density} of the four
-applications, in terms of average number of actions per state involved
+applications, \ie\ the average number of actions per state involved
 in traces, and average number of \emph{unique} actions per state. The
@@
 thus it could be easier to learn.
-
-\mytodo{result}
 

### 2015-09-25 paper.tex
final?

@@
 Developing good UIs for web or mobile applications is a
-complex and expensive endeavour. One reason is the combination of
+complex and expensive endeavor. One reason is the combination of
 devices, interaction modalities and workflows that need to be
@@
 questions can be automatically computed. In this way a system can
-provide important insights to a designer, and support assements of
+provide important insights to a designer, and support assessments of
 potential user flexibility, user efficiency, error proneness, ability
@@
 The literature on using state-transition networks for specifying or
-analysing the behavior of UIs is vast. We conducted a systematic-style
+analyzing the behavior of UIs is vast. We conducted a systematic-style
 literature review, using Google Scholar and queries with combinations
@@
 \emph{deviation}. Defined specifically for hypertext systems, lostness
-is a user performance measure which is afuncion of the number of
+is a user performance measure which is a function of the number of
 visited nodes, the number of different nodes that were visited nodes,
@@
 browsing information, rather than trying to achieve specific goals. It
-is suitable therefore whe some of the following three assumptions can
+is suitable therefore when some of the following three assumptions can
 be relaxed: that there is a task to complete, that there is a correct
@@
 
-In~\cite{hilbert2000} several approaches to analyse streams of user
+In~\cite{hilbert2000} several approaches to analyze streams of user
 events are discussed and compared. It is interesting to realize that
@@
 the resulting FSA, the \emph{interaction graph}. It is a directed
-multigraph\footnote{A directed multigraph is a graph such that there
-  are 2 or more edges that have the same end points. Cycles are paths
-  that include 2 or more occurrences of the same vertex. Loops are
-  edges that start and end on the same vertex. ``Geodesic path'' is a
-  synonymous term with ``shortest path''.}, potentially with
-cycles and loops, with edges that are labeled with the name of the
-corresponding action. In the following we will be using as synonymous
-the terms state/vertex, and action/edge.
+multigraph\footnote{A directed multigraph is a directed graph such
+  that there are 2 or more edges that have the same end points. Cycles
+  are paths that include 2 or more occurrences of the same
+  vertex. Loops are edges that start and end on the same
+  vertex. ``Geodesic path'' is a synonymous term with ``shortest
+  path''.}, potentially with cycles and loops, with edges that are
+labeled with the name of the corresponding action. In the following we
+will be using as synonymous the terms state/vertex, and action/edge.
 
@@
 creates traces with detour order $k+1$ by collecting the set $D_k$ of states
-with order $0$ up to $k$, and by identifying the neighbours ($N_k$)
+with order $0$ up to $k$, and by identifying the neighbors ($N_k$)
 of $D_k$ ($N_k$ is the set of states  not included in $D_k$ that are
@@
 \texttt{viewingConversations}); available actions include moving to
-the next or previuous block of conversations, refreshing the list, or
+the next or previous block of conversations, refreshing the list, or
 opening a specific conversation (transition
@@
 the 47 states: because of the large number of cycles that exist among
-states, edges form an intricated web of possible action sequences. We
+states, edges form an intricate web of possible action sequences. We
 believe this in practice greatly reduces usefulness of typical network
@@
 further the graph, by specifying scenarios and computing traces.  The
-following fragment of R code specifies the bridge sets of the choosen
+following fragment of R code specifies the bridge sets of the chosen
 scenario: it comprises, in the given order, two occurrences of
@@
 or more are the shortest ones among the four applications, Gmail also
-makes users more efficient in recovering from errors.  \mytodo{result}
+makes users more efficient in recovering from errors.
 
@@
 that can be exploited with Gmail are counterbalanced by the required
-knowledge and capability of chosing an optimal path. In particular,
+knowledge and capability of choosing an optimal path. In particular,
 Gmail offers many detours of order 1 which increase flexibility for
@@
 former is an overall measure of the number of options that are made
-available by a UI, the latter can be used to analyse how many
+available by a UI, the latter can be used to analyze how many
 \emph{new} options the user is presented with in any state. For our
@@
 \subsection{Comparing scenarios}
-
-The same kind of analysis can be carried out to assess how suitable a
-design is for a set of different scenarios. For example, a designer
-might be interested in understanding what are some quantitative
-differences in replying to a message as opposed to composing a new
-one.
-
-For Horde, it turns out that composing a new message entails many more
-order 0 paths (154 vs 86), and a comparable number of higher order
-paths. The path length in the best case is the same across the two
-scenarios, but in the average case composing has a length of 7.75
-steps compared to 8.5 for reply. Probability of hitting an optimal
-path is 4 times higher for compose.
-
-This means that users will have a twice as much large choice of
-correct paths when composing a new message rather than ``simply''
+Previous examples show how \migtool\ can be used to compare different
+UIs against the same scenario.  The same kind of analysis can be
+carried out to assess how suitable a design is for different
+scenarios. For example, a designer might be interested in a comparison
+of replying to a message as opposed to composing a new one.
+
+Focusing now on Horde, it turns out that composing a new message
+entails many more order 0 paths (154 vs 86), and a comparable number
+of higher order paths. The path length in the best case is the same
+across the two scenarios, but in the average case composing has a
+length of 7.75 steps compared to 8.5 for reply. Probability of hitting
+an optimal path is 4 times higher for compose.
+
+This means that users will have twice as much many choices between 
+correct paths when composing a new message rather than simply
 replying to a read one. On average, when composing, users could be
@@
 it is for replying to existing ones.
-\todo[inline]{result}
 
@@
 adding a widget or feature to an existing UI.  For example, we studied
-the cruise control features of cars. One of the examples is system S,
-where the driver can engage the system, and once it is engaged, speed
-can be increased or decreased with small or large steps. Of course the
-system can also be disengaged (by pressing the brake pedal, for
-example).
+some cruise control features of cars. One of the examples is system
+S\footnote{There is no need to disclose the actual brand.}, where the
+driver can engage the system, and once it is engaged, speed can be
+increased or decreased with small or large steps. Of course the system
+can also be disengaged (by pressing the brake pedal, for example).
 System A is more elaborate, as it includes also a memory function:
 when it is disengaged it remembers the current speed, which can be
-recalled later on.  There are two ways to re-engage it: one
-by setting a new speed, and one by recalling the previous one. In
-addition,  if the car drives for more than 5 minutes at a higher
-speed than the set one, system A automatically disengages.
+recalled later on.  There are two ways to re-engage it: one by setting
+a new speed, and one by recalling the previous one. In addition, if
+the car drives for more than 5 minutes at a higher speed than the set
+one, system A automatically disengages.
 
@@
 these functionalities?'' in a typical driving scenario where a speed
-is selected, then the system is disengaged, and then the same speed
-needs to be set.
-
-System S (where we assumed that re-setting the speed is done manually
-by the driver with 4 actions on ``up'' and ``down'' to
-approximately get the same speed) leads to 1 order 0 path requiring 7
-steps; there are 8 order 1 paths with average length  9.4. The
-probability of hitting the optimal path is 0.11; action density is
-1.75, and unique action density is 1.25.
-
-On the other hand, system A has 2 order 0 paths (average length 4.5),
-and 7 order 1 paths (average length 6.8). 
-The optimal path probability is 0.06; action density is
-2.2, and unique action density is 1.2.
-
-In both cases the reasons for detours deal with the possibility of
-disengaging the system at the wrong moment.
+is selected, then the system is disengaged, and later on the same
+speed needs to be set.
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c|r|r|r|r|r|r|r|r|}\hline
+    System & $N_0$ & $L_0$ & $N_1$ & $L_1$ & p & AD & UAD \\\hline
+    S  & 1 & 7 & 8 & 9.4& 11\% & 1.75 & 1.25 \\
+    A  & 2 & 4.5 & 7 & 6.8 & 6\% & 2.20 & 1.20 \\\hline
+  \end{tabular}\vspace{1mm}
+  \caption{Comparison of the two cruise control designs. $N_i$: number
+    of execution traces with detour order $i$; $L_i$: average length of
+    traces with detour order $i$; $p$: probability to hit the shortest execution
+    trace; AD: actions/state; UAD: unique actions/state.}
+  \label{tab:data}
+\end{table}
+
+When using a scenario for S where we assume that re-setting the speed
+is done manually by the driver with 4 actions on ``up'' and ``down''
+to approximately get the same speed), the comparison produces the
+values shown in Table~\ref{tab:data}. 
+System A has 2 alternative optimal paths, their length is 4.5 (36\%
+shorter than in S); both systems have a comparable number of detour 1
+traces, but again their length in A is  28\% shorter. The probability
+of hitting the optimal trace is however twice as large in S, which
+features also a more compact design. 
+In both cases detours are caused by the possibility of disengaging the
+system at the wrong moment.
 
 Therefore we can conclude that:
-\begin{inparaenum}
-\item system A makes users more efficient;
+\begin{inparaenum}[(1)]
+\item system A makes users more efficient (an increase of 56\%) for
+  the considered scenario;
 \item with A there are two possible ways to achieve the scenario, thus
@@
 \item system A features a more compact design, with fewer action types
-  to be performed at each moment, being therefore potentially easier
-  to learn and remember.
+  to be performed at each moment.
 \end{inparaenum}
-\todo[inline]{result}
 
@@
 UML state diagrams provide a very expressive language, well suited to
-specify behavior of UIs based on descrete events. Evidence of this can
+specify behavior of UIs based on discrete events. Evidence of this can
 also be found in recent OMG standards, such as
@@
 handle perceptive UIs - because they are not well suited to be
-described in terms of descrete states), in many practical cases they
-can be be isolated and/or left aside).
+described in terms of discrete states), in many practical cases they
+can be be isolated and/or ignored.
 
@@
 are likely to produce different models for the same UI. As a
-consequence, metrics computed by \migtool\ do depend on different
+consequence, it is possible that metrics computed by \migtool\ do depend on different
 modeling choices. It is worth mentioning, though, that because of the
@@
 those dealing with using a different hierarchy of states, or with
-distributing differently concurrent regions across states), and such a
+distributing differently concurrent regions across states), and 
 sensitivity is correspondingly reduced. At the moment, however, we do
@@
 combined with other results \emph{after} the analysis is
-performed. This also means that no effort needs to be directed on
-developing data and presentation specifications/implementations.
+performed.
 
@@
 preferences, skills, interpretations, ergonomic constraints) they are
-very general. On the other hand, predictions based on \migtool\ metrics
-are also generic because they do not consider data and presentation
-aspects, nor user-related factors. For example, it is unfeasible to
-use \migtool\ to predict the time needed by a user to complete a
-scenario. However, \migtool\ can be used to analyze the whole
-interaction design and gather data to inform more specific analyses
-that could be performed, for example, with CogTool~\cite{john04cogtool}.
+very general and conservative. On the other hand, predictions based on
+\migtool\ metrics are also generic because they do not consider data
+and presentation aspects, nor user-related factors. For example, it is
+unfeasible to use \migtool\ to predict the time needed by a user to
+complete a scenario. However, as mentioned before, \migtool\ can be
+used to analyze the whole interaction design and gather data to inform
+more specific analyses that could be performed, for example, with
+CogTool.
+
+At the moment \migtool\ does not use weighted edges in the interaction
+graph. It is easy to extend it to search paths that minimize the total
+weight, and therefore in such a way to perform analyses that are
+similar to the ones suggested in~\cite{thimbleby2001markov}.
+
+Likewise it is possible to extend \migtool\ to compute also the
+lostness metric~\cite{smith1996lostness}, based on states with
+different detour orders. This would give yet another objective metric
+to compare different designs. We opted for non doing it in light of
+the fact that lostness was designed for exploratory activities, not
+for goal oriented scenarios like the ones we considered in previous
+examples.
 
 \migtool\ is part of a suite of model-based tools for designing,
-analysing and testing user interfaces developed by the company
+analyzing and testing user interfaces developed by the company
 ANONYMIZED. \migtool\ is implemented partly in Java (model processing)
@@
 \label{sec:conclusion}
+
+We presented a method that can be used to support analysis of user
+interfaces even before they are built or prototyped. UML statecharts
+representing the intended behavior of user interfaces are processed
+and matched to interaction scenarios, producing a number of
+graph-theoretic metrics defined on possible execution traces.
+
+We showed that with this approach one can compare different designs
+against the same scenario, or the same design against different
+scenarios. In both situations precise, quantitative and objective
+measures can be generated regarding flexibility offered to users,
+their potential efficiency, and error proneness of the user interface.
+
+The method should not be used to draw final usability conclusions, as
+it is devoid of any concerns dealing with what is presented to users,
+how they could perceive that, understand that, and manipulate that.
+But, because the approach can be applied also when only a
+specification of the user interface is available, it supports
+construction and engineering of interactive systems and could help in
+iteratively simplifying or improving a design.
 

### 2015-10-15 paper.tex

@@
 
-  We present a tool (\migtool) that transforms models of a user
+  We present a tool (MIGTool) that transforms models of a user
   interface into a graph, upon which usage scenarios can be easily
@@
 % References must be the same font size as other body text.
-\bibliographystyle{SIGCHI-Reference-Format}
-\bibliography{gb,database}
+%\bibliographystyle{SIGCHI-Reference-Format}
+%\bibliography{gb,database}
+%%% -*-BibTeX-*-
+%%% Do NOT edit. File created by BibTeX with style
+%%% ACM-Reference-Format-Journals [18-Jan-2012].
+
+\begin{thebibliography}{00}
+
+%%% ====================================================================
+%%% NOTE TO THE USER: you can override these defaults by providing
+%%% customized versions of any of these macros before the \bibliography
+%%% command.  Each of them MUST provide its own final punctuation,
+%%% except for \shownote{}, \showDOI{}, and \showURL{}.  The latter two
+%%% do not use final punctuation, in order to avoid confusing it with
+%%% the Web address.
+%%%
+%%% To suppress output of a particular field, define its macro to expand
+%%% to an empty string, or better, \unskip, like this:
+%%%
+%%% \newcommand{\showDOI}[1]{\unskip}   % LaTeX syntax
+%%%
+%%% \def \showDOI #1{\unskip}           % plain TeX syntax
+%%%
+%%% ====================================================================
+
+\ifx \showCODEN    \undefined \def \showCODEN     #1{\unskip}     \fi
+\ifx \showDOI      \undefined \def \showDOI       #1{{\tt DOI:}\penalty0{#1}\ }
+  \fi
+\ifx \showISBNx    \undefined \def \showISBNx     #1{\unskip}     \fi
+\ifx \showISBNxiii \undefined \def \showISBNxiii  #1{\unskip}     \fi
+\ifx \showISSN     \undefined \def \showISSN      #1{\unskip}     \fi
+\ifx \showLCCN     \undefined \def \showLCCN      #1{\unskip}     \fi
+\ifx \shownote     \undefined \def \shownote      #1{#1}          \fi
+\ifx \showarticletitle \undefined \def \showarticletitle #1{#1}   \fi
+\ifx \showURL      \undefined \def \showURL       #1{#1}          \fi
+
+\bibitem{ambler02}
+{S. Ambler}. 2002.
+\newblock {\em Agile Modeling: Effective Practices for eXtreme Programming and
+  the Unified Process}.
+\newblock Wiley.
+\newblock
+\showISBNx{0471202827}
+
+\bibitem{ammann2008introduction}
+{P. Ammann} {and} {J. Offutt}. 2008.
+\newblock {\em Introduction to software testing}.
+\newblock Cambridge University Press.
+\newblock
+
+\bibitem{appert2008swingstates}
+{C. Appert} {and} {M. Beaudouin-Lafon}. 2008.
+\newblock \showarticletitle{SwingStates: adding state machines to Java and the
+  Swing toolkit}.
+\newblock {\em Software: Practice and Experience\/} {38}, 11 (2008),
+  1149--1182.
+\newblock
+
+\bibitem{conversy07}
+{E. Barboni}, {S. Conversy}, {D. Navarre}, {and} {P. Palanque}. 2007.
+\newblock \showarticletitle{Model-based engineering of widgets, user
+  applications and servers compliant with ARINC 661 specification}.
+\newblock In {\em Interactive Systems. Design, Specification, and
+  Verification}. Springer, 25--38.
+\newblock
+
+\bibitem{bellamy2011cogtool}
+{R. Bellamy}, {B. John}, {and} {S. Kogan}. 2011.
+\newblock \showarticletitle{Deploying CogTool: integrating quantitative
+  usability assessment into real-world software development}. In {\em Software
+  Engineering (ICSE), 2011 33rd International Conference on}. IEEE, 691--700.
+\newblock
+
+\bibitem{briand2005:test:uml}
+{L.~C Briand}, {Y. Labiche}, {and} {J. Cui}. 2005.
+\newblock \showarticletitle{Automated support for deriving test requirements
+  from UML statecharts}.
+\newblock {\em Software \& Systems Modeling\/} {4}, 4 (2005), 399--423.
+\newblock
+
+\bibitem{buxton07}
+{B. Buxton}. 2007.
+\newblock {\em User Experience: Getting the Design Right and the Right Design}.
+\newblock Morgan Kaufmann.
+\newblock
+
+\bibitem{ceri:00}
+{S. Ceri}, {P. Fraternali}, {and} {A. Bongio}. 2000.
+\newblock \showarticletitle{{Web Modeling Language (WebML)}: a modeling
+  language for designing web sites}.
+\newblock {\em Computer Networks\/}  {33} (2000), 137--157.
+\newblock
+
+\bibitem{constantine99}
+{L.L. Constantine} {and} {L.A.D. Lockwood}. 1999.
+\newblock {\em Software for use: a practical guide to the models and methods of
+  usage-centered design}.
+\newblock Addison-Wesley.
+\newblock
+
+\bibitem{gomez01}
+{J. G\'{o}mez}, {C. Cachero}, {and} {O. Pastor}. 2001.
+\newblock \showarticletitle{Conceptual Modeling of Device-Independent Web
+  Applications}.
+\newblock {\em IEEE MultiMedia\/} {8}, 2 (2001), 26--39.
+\newblock
+\showISSN{1070-986X}
+\showDOI{%
+\url{http://dx.doi.org/10.1109/93.917969}}
+
+\bibitem{harel09}
+{D. Harel}. 2009.
+\newblock \showarticletitle{Statecharts in the making: a personal account}.
+\newblock {\em CACM\/} {52}, 3 (March 2009), 67--75.
+\newblock
+
+\bibitem{heymann2007formal}
+{M. Heymann} {and} {A. Degani}. 2007.
+\newblock \showarticletitle{Formal analysis and automatic generation of user
+  interfaces: approach, methodology, and an algorithm}.
+\newblock {\em Human Factors: The Journal of the Human Factors and Ergonomics
+  Society\/} {49}, 2 (2007), 311--330.
+\newblock
+
+\bibitem{hilbert2000}
+{D.~M Hilbert} {and} {D.F. Redmiles}. 2000.
+\newblock \showarticletitle{Extracting usability information from user
+  interface events}.
+\newblock {\em ACM Computing Surveys (CSUR)\/} {32}, 4 (2000), 384--421.
+\newblock
+
+\bibitem{hornbaek2006current}
+{K. Hornb{\ae}k}. 2006.
+\newblock \showarticletitle{Current practice in measuring usability: Challenges
+  to usability studies and research}.
+\newblock {\em International Journal of Human-Computer Studies\/} {64}, 2
+  (2006), 79--102.
+\newblock
+
+\bibitem{horrocks99}
+{Ian Horrocks}. 1999.
+\newblock {\em Constructing the User Interface with Statecharts}.
+\newblock Addison-Wesley Longman Publishing Co., Inc., Boston, MA, USA.
+\newblock
+\showISBNx{0201342782}
+
+\bibitem{john04cogtool}
+{B.E. John}, {K. Prevas}, {D.D. Salvucci}, {and} {K. Koedinger}. 2004.
+\newblock \showarticletitle{Predictive human performance modeling made easy}.
+  In {\em Proc. of the Int. Conf. on Human Factors in Computing Systems}. ACM,
+  New York, NY, USA, 455--462.
+\newblock
+
+\bibitem{koch02:uwe}
+{N. Koch} {and} {A. Kraus}. 2002.
+\newblock \showarticletitle{{The expressive power of {UML}-based web
+  engineering}}. In {\em Second International Workshop on Web-Oriented Software
+  Technology (IWWOST02)}, {D.~Schwabe}, {O.~Pastor}, {G.~Rossi}, {and}
+  {L.~Olsina} (Eds.). 105--199.
+\newblock
+
+\bibitem{mccurdy06}
+{M. McCurdy}, {C. Connors}, {G. Pyrzak}, {B. Kanefsky}, {and} {A. Vera}. 2006.
+\newblock \showarticletitle{Breaking the fidelity barrier: an examination of
+  our current characterization of prototypes and an example of a mixed-fidelity
+  success}. In {\em CHI 2006}. ACM, ACM Press, New York, NY, 1233--1242.
+\newblock
+
+\bibitem{melia08}
+{S. Meli\'a}, {J. G\'omez}, {S. P\'erez}, {and} {O. D\'iaz}. 2008.
+\newblock \showarticletitle{A Model-Driven Development for GWT- Based Rich
+  Internet Applications with OOH4RIA}. In {\em Proc. 8th Int'l Conf. Web Eng.
+  (ICWE 2008)}. IEEE CS Press, 13--23.
+\newblock
+
+\bibitem{mori02}
+{G. Mori}, {F. Patern\`o}, {and} {C. Santoro}. 2002.
+\newblock \showarticletitle{{CTTE: Support for Developing and Analysing Task
+  Models for Interactive System Design}}.
+\newblock {\em IEEE Transactions on Software Engineering\/} {28}, 8 (August
+  2002), 797--813.
+\newblock
+
+\bibitem{ifml13}
+{{OMG}}. 2013.
+\newblock {\em Interaction Flow Modeling Language ({IFML}), FTF -- Beta 1\/}
+  (omg document number: ptc/2013-03-08 ed.).
+\newblock {T}echnical {R}eport. OMG.
+\newblock
+\showURL{%
+\url{http://www.omg.org/spec/IFML/1.0}}
+
+\bibitem{uml25}
+{{Object Management Group} OMG}. 2015.
+\newblock {OMG Unified Modeling Language ({OMG UML})} Version 2.5.
+\newblock \url{http://www.omg.org/spec/UML/2.5}.   (March 2015).
+\newblock
+\showURL{%
+\url{http://www.omg.org/spec/UML/2.5/PDF}}
+
+\bibitem{otter2000lostness}
+{M. Otter} {and} {H. Johnson}. 2000.
+\newblock \showarticletitle{Lost in hyperspace: metrics and mental models}.
+\newblock {\em Interacting with Computers\/} {13}, 1 (2000), 1--40.
+\newblock
+
+\bibitem{page1999pagerank}
+{L. Page}, {S. Brin}, {R. Motwani}, {and} {T. Winograd}. 1999.
+\newblock \showarticletitle{The PageRank citation ranking: bringing order to
+  the Web.}
+\newblock  (1999).
+\newblock
+
+\bibitem{parnas69}
+{D.L. Parnas}. 1969.
+\newblock \showarticletitle{On the use of transition diagrams in the design of
+  a user interface for an interactive computer system}. In {\em ACM '69 Proc.
+  of the 1969 24th National Conference}. ACM.
+\newblock
+
+\bibitem{silva03:umli}
+{P. Pinhero~da Silva} {and} {N.W. Paton}. 2003.
+\newblock \showarticletitle{User Interface Modeling in {UMLi}}.
+\newblock {\em IEEE Software\/} (2003), 62--69.
+\newblock
+
+\bibitem{preece02}
+{J. Preece}, {Y. Rogers}, {and} {H. Sharp}. 2002.
+\newblock {\em Interaction design}.
+\newblock John Wiley and Sons.
+\newblock
+
+\bibitem{rubin08}
+{J. Rubin} {and} {D. Chisnell}. 2008.
+\newblock {\em Handbook of Usability Testing\/} (second ed.).
+\newblock Wiley.
+\newblock
+
+\bibitem{samek09}
+{M. Samek}. 2009.
+\newblock {\em Practical {UML} Statecharts in {C/C++}}.
+\newblock Elsevier.
+\newblock
+
+\bibitem{smith1996lostness}
+{P.A. Smith}. 1996.
+\newblock \showarticletitle{Towards a practical measure of hypertext
+  usability}.
+\newblock {\em Interacting with Computers\/} {8}, 4 (1996), 365--381.
+\newblock
+
+\bibitem{thimbleby07}
+{H. Thimbleby}. 2007.
+\newblock {\em Press on: principles of interaction programming}.
+\newblock The MIT Press.
+\newblock
+
+\bibitem{thimbleby2001markov}
+{H. Thimbleby}, {P. Cairns}, {and} {M. Jones}. 2001.
+\newblock \showarticletitle{Usability analysis with Markov models}.
+\newblock {\em ACM Transactions on Computer-Human Interaction (TOCHI)\/} {8}, 2
+  (2001), 99--132.
+\newblock
+
+\bibitem{thimbleby09:sna:eics09}
+{H. Thimbleby} {and} {P. Oladimeji}. 2009.
+\newblock \showarticletitle{Social Network Analysis and Interactive Device
+  Design Analysis}. In {\em Proc. of Engineering Interactive Computing Systems
+  2009}. ACM Press, 91--100.
+\newblock
+
+\end{thebibliography}
 

### 2016-01-09 paper.tex
submitted abstract to eics16
changed the title, extended the abstract, written the contribution.

@@
 
-\title{Measuring Interaction Design}
+\title{Measuring Interaction Design \\before Building the System: a
+  Model-Based Approach}
 
@@
 
+% \author{Giorgio Brajnik\inst{1,2} \and Simon Harper\inst{1}}
+% \institute{\textsuperscript{1}Computer Science School\\
+% University of Manchester, UK\\
+% \email{simon.harper@manchester.ac.uk}\\
+% and\\
+% \inst{2}Dipartimento di Matematica e Informatica\\
+% Universit\`a di Udine, Italy\\
+% \email{brajnik@uniud.it}
+% }
+
 \maketitle
@@
 
-  We present a tool (MIGTool) that transforms models of a user
-  interface into a graph, upon which usage scenarios can be easily
-  specified, and used to compute possible interaction paths. Metrics
-  based on possible paths, with or without user errors, can be easily
-  computed.  For example, when applying them to compare mail
-  applications, we show that Gmail has 3 times more shortest routes,
-  has 2x more routes that include a single user error, has routes with
-  13\% fewer steps, but optimal routes have the smallest probability
-  to be chosen.
+  We present a tool (MIGTool) that transforms models of the behavior
+  of a user interface into a graph, upon which usage scenarios can be
+  easily specified, and used by MIGTool to compute possible
+  interaction paths. Metrics based on possible paths, with or without
+  user navigation errors, can then be computed.  For example, when
+  analyzing four mail applications, we show that Gmail compared to
+  others has 3 times more shortest routes, has 2x more routes that
+  include a single user error, has routes with 13\% fewer steps, but
+  has also optimal routes with the smallest probability to be chosen.
+
+  Without MIGTool, this kind of analysis could only be done after
+  building some prototype of the system, and then only for specific
+  scenarios by manually tracing user actions and relative changes to
+  the screens. With MIGTool the exploration of suitability of a design
+  with respect to different scenarios, or comparison of different
+  design alternatives, can be done after providing only a partial
+  specification of the user interface behavior.
+
+  This is made possible by the ability to associate scenarios steps to
+  required user actions as defined in the model, by an efficient
+  strategy to identify complete execution traces that users can
+  follow, and by computing a range of diverse metrics on these
+  results. The agile modeling approach that was followed implies that
+  relatively little work is needed upfront.
+
+  CONTRIBUTION (asked when submitting the abstract)
+
+  There are three contributions.
+
+  First, we provide means to concisely ground usage scenarios on the
+  basis of actions specified in models. This mechanism is general
+  enough that can be used also to link tasks or use cases to actual
+  behavior of a user interface.
+
+  Second, MIGTool transforms models and scenarios into a graph of
+  possible execution paths. Thanks to the special-purpose language
+  that is provided to specify scenarios, MIGTool is quite efficient,
+  with a time complexiy that is linear with the number of steps in a
+  scenario and the number of edges in the graph. We describe the
+  language and the algorithm.
+
+  Third, we provide definition of a handful of metrics that are
+  objective, precise and that bear upon usability of the resulting
+  user interface. They are very general, can be used for most kinds of
+  user interfaces that can be modeled with discrete events, and are
+  platform independent.
+
+%% EICS submission: #122
+%% https://precisionconference.com/~eics16a/updatePaperResponse?userRef=CILXDODXPNF0VESOAFMMC8I16ZO5FR5J&paperNumber=112&emailSent=&error=
 

### 2016-01-14 paper.tex
revised background and discussion sections.
According to what CHI reviewrs said and considering that EICS is more
focussed on model-based approaches.

--- 2015-09-25 paper.tex
@@
 \hypersetup{%
-  pdftitle={SIGCHI Conference Proceedings Format},
+  pdftitle={Measuring Interaction Design},
   pdfauthor={LaTeX},
-  pdfkeywords={SIGCHI, proceedings, archival format},
+  pdfkeywords={model-based User Interfaces},
   bookmarksnumbered,
@@
 
-  We present a tool (\migtool) that transforms models of a user
-  interface into a graph, upon which usage scenarios can be easily
-  specified, and used to compute possible interaction paths. Metrics
-  based on possible paths, with or without user errors, can be easily
-  computed.  For example, when applying them to compare mail
-  applications, we show that Gmail has 3 times more shortest routes,
-  has 2x more routes that include a single user error, has routes with
-  13\% fewer steps, but optimal routes have the smallest probability
-  to be chosen.
+  We present a tool (MIGTool) that transforms models of the behavior
+  of a user interface into a graph, upon which usage scenarios can be
+  easily specified, and used by MIGTool to compute possible
+  interaction paths. Metrics based on possible paths, with or without
+  user navigation errors, can then be computed.  For example, when
+  analyzing four mail applications, we show that Gmail compared to
+  others has 3 times more shortest routes, has 2x more routes that
+  include a single user error, has routes with 13\% fewer steps, but
+  has also optimal routes with the smallest probability to be chosen.
+
+  Without MIGTool, this kind of analysis could only be done after
+  building some prototype of the system, and then only for specific
+  scenarios by manually tracing user actions and relative changes to
+  the screens. With MIGTool the exploration of suitability of a design
+  with respect to different scenarios, or comparison of different
+  design alternatives, can be done after providing only a partial
+  specification of the user interface behavior.
+
+  This is made possible by the ability to associate scenarios steps to
+  required user actions as defined in the model, by an efficient
+  strategy to identify complete execution traces that users can
+  follow, and by computing a range of diverse metrics on these
+  results. The agile modeling approach that was followed implies that
+  relatively little work is needed upfront.
+
+  % CONTRIBUTION (asked when submitting the abstract)
+
+  % There are three contributions.
+
+  % First, we provide means to concisely ground usage scenarios on the
+  % basis of actions specified in models. This mechanism is general
+  % enough that can be used also to link tasks or use cases to actual
+  % behavior of a user interface.
+
+  % Second, MIGTool transforms models and scenarios into a graph of
+  % possible execution paths. Thanks to the special-purpose language
+  % that is provided to specify scenarios, MIGTool is quite efficient,
+  % with a time complexiy that is linear with the number of steps in a
+  % scenario and the number of edges in the graph. We describe the
+  % language and the algorithm.
+
+  % Third, we provide definition of a handful of metrics that are
+  % objective, precise and that bear upon usability of the resulting
+  % user interface. They are very general, can be used for most kinds of
+  % user interfaces that can be modeled with discrete events, and are
+  % platform independent.
+
+%% EICS submission: #122
+%% https://precisionconference.com/~eics16a/updatePaperResponse?userRef=CILXDODXPNF0VESOAFMMC8I16ZO5FR5J&paperNumber=112&emailSent=&error=
 
@@
 %
-Many papers deal with testing user interfaces, and problems related to
-generating execution traces as a means to assess coverage of a test
-suite~\cite{ammann2008introduction,briand2005:test:uml}. They were
-excluded from the review.
+% Many papers deal with testing user interfaces, and problems related to
+% generating execution traces as a means to assess coverage of a test
+% suite~\cite{ammann2008introduction,briand2005:test:uml}. They were
+% excluded from the review.
 
@@
 statecharts are seen as crucial representations that allow a designer to
-fully appreciate how devices behave. The overall claim is that ``If
+fully appreciate how devices behave. The overall stance is that ``If
 you don't understand the logic conveyed by a statechart model of a
@@
 
-In~\cite{conversy07} the Interactive Cooperative Objects formalism is
-discussed, which is based on Petri nets, yet another state transition
-formalism.  It is used to enrich the ARINC 661 specification of
-Cockpit Display System so that the semantics of widgets can be
-expressed and the behavior of a UI be formally analyzed.
-
+% In~\cite{conversy07} the Interactive Cooperative Objects formalism is
+% discussed, which is based on Petri nets, yet another state transition
+% formalism.  It is used to enrich the ARINC 661 specification of
+% Cockpit Display System so that the semantics of widgets can be
+% expressed and the behavior of a UI be formally analyzed.
+% 
 %  that are now under provides precise information
@@
 handlers in the UI can more easily be conceived, developed and
-verified (\eg, \cite{appert2008swingstates}).
+verified (\eg, \cite{appert2008swingstates}). More sophisticated
+model-based approaches employ statecharts to orchestrate the behavior of
+different components of a user interface \cite{winckler03}. These
+models, however, are not used to support usability analysis of any
+kind.
 
@@
 
-A review of usability measuring practices~\cite{hornbaek2006current}
-lists, under the headings \emph{measures of efficiency/usage
-  patterns}, number of keystrokes, mouse clicks, and visited objects
-as possible metrics. The notion of \emph{deviation from optimal
-  solution} is discussed only in the context of tools for route
-planning in 3D navigation. In this paper we provide our own definition
-of deviation, which applies to interaction with the UI; we provide
-also our notion of execution traces across states of the UI, and the
-length of these traces can be used as a measure of efficiency.
-
 Lostness~\cite{smith1996lostness,otter2000lostness} is one of the few
@@
 
+The notion of \emph{deviation from optimal
+  solution} is discussed only in the context of tools for route
+planning in 3D navigation. In this paper we provide our own definition
+of deviation, which applies to interaction with the UI; we provide
+also our notion of \emph{potential execution traces} across states of
+the UI, and the length of these traces can be used as a measure of
+efficiency.
+
+In \cite{fayollas2014}, hierarchical task models (specified with
+HAMSTERS) are used to specify how pilots are expected to interact with
+airplane cockpit interfaces in tasks such as changing value to some
+flight parameter. The paper in particular discusses how these task
+models can be used to specify alternative ways to achieve te same
+goal, and analyse the consequences on the reliability of the resulting
+system. Compared to MIGTool, that approach cannot take advantage of a
+model of the system and has instead to rely only on models of the
+tasks. The conseguence is that analysing alternative ways to achieve
+the same goal requires developing different models; in addition,
+exploring suitability of a given system with respect to certain tasks
+is complex; and finally, hierarchical task models cannot take
+advantage of contexts and rich control flows that are available with
+state-transition representations, limiting therefore the richness of
+the scenarios being analysed.
+
+A review of usability measuring practices~\cite{hornbaek2006current}
+lists, under the headings \emph{measures of efficiency/usage
+  patterns}, number of keystrokes, mouse clicks, and visited objects
+as possible metrics. Notice that in general the time needed to carry
+out individual interaction steps depends on many aspects, such as the
+time to decide what goal to achieve, to understand how to do it, to
+locate the actions to do it, to perform the action, to attend to the
+feedback. Layout of the user interface, Fitt's law considerations,
+cognitive state of the user, physical and environmental aspects affect
+such times. For these reasons we opted, in \migtool, to only focus on
+required steps to complete a task, not on completion time.
+
 A usability analysis method capable to analytically predict task
@@
 modality and which widgets are used to implement actions, the designer
-gets an estimation of how long an experienced user would
-spend on each step. CogTool takes care of adding extra ``mental''
-steps, such as ``think'' steps, before certain patterns of provided
-steps, according to the cognitive theory underlying ACT-R. As a
-result, users of CogTool obtain the breakdown of the times required by
-each of the steps. Our method is weaker in terms of cognitive validity
-and in terms of precision of the output: it does not provide expected
-completion times. However, with our method one can analyze a large
-part of the UI, get information about possible problems in some areas,
-and only then devolve more resources in building storyboards and
-in making assumptions regarding widgets so that specific execution
-paths previously identified can be analyzed with CogTool. In a sense,
-the output of our method could be used to make informed decisions as
-to what to analyze with CogTool.
+gets an estimation of how long an experienced user would spend on each
+step. CogTool takes care of adding extra ``mental'' steps, such as
+``think'' steps, before certain patterns of provided steps, according
+to the cognitive theory underlying ACT-R. As a result, users of
+CogTool obtain the breakdown of the times required by each of the
+steps. Our method is less precise: it does not provide expected
+completion times that are cognitively justified. However, with our
+method, you can analyze a large part of the UI, get information about
+possible problems in some areas, and only then devolve more resources
+in building storyboards and in making assumptions regarding widgets so
+that specific execution paths previously identified can be analyzed
+with CogTool. In a sense, the output obtained with our method could be
+used to make informed decisions as to what to analyze next with
+CogTool. Approaches like \cite{swearngin13} start from the assumption
+that the user interface exists already, and that it can \emph{ripped}
+to reverse engineer a behavior model. These assumptions are not needed
+to use \migtool.
+
+Another approach for analysing user actions that has a strong
+cognitive background is SNIF-ACT \cite{pirolli2007}. However it has
+been conceived and applied to information websites, where the notion
+of \emph{information scent} can bear upon interaction. We opted for
+not applying such concepts to the interaction with a web application,
+where most of the actions are not aimed at information finding.
 
@@
 invented, back in early eighties, and why they were appealing also to
-non experts~\cite{harel09}.  For the sake of brevity, we refrain from
-describing them here in detail; the interested reader is referred to
-the UML standard~\cite{uml25} or some of the textbooks that deal with
-them, like~\cite{samek09}.
+non experts~\cite{harel09}.%   For the sake of brevity, we refrain from
+% describing them here in detail; the interested reader is referred to
+% the UML standard~\cite{uml25} or some of the textbooks that deal with
+% them, like~\cite{samek09}.
 
@@
 
-\subsection{Defining usage scenarios}
+\subsection{Grounding usage scenarios}
 
@@
 to support a fair comparison, all four models cover the same set of
-use cases: listing the content of the inbox, reading a message or
-conversation, replying to a message, composing and sending a new
-message.
+use cases at the same level of detail: listing the content of the
+inbox, reading a message or conversation, replying to a message,
+composing and sending a new message.
 
@@
 reported below. The actual model that we used consists of 
-24 states, 12 pseudo states, 12 regions, and 61 transitions.
+24 states, 12 pseudo states, 12 regions, and 61 transitions. The model
+is similar to that being discussed in examples mentioned by \cite{ifml13}.
 
@@
 particular structure is evident that was not known from the
-model. Figure~\ref{fig:gom} shows a plot using a circular layout of
-the 47 states: because of the large number of cycles that exist among
-states, edges form an intricate web of possible action sequences. We
-believe this in practice greatly reduces usefulness of typical network
-analysis metrics, such as \emph{betweenness, eccentricity, page-rank,
-  eigenvalue} centrality measures.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figures/gom.png}
-  \caption{Interaction graph obtained from the Gmail model.}
-  \label{fig:gom}
-\end{figure}
+model.
+%  Figure~\ref{fig:gom} shows a plot using a circular layout of
+% the 47 states: 
+Because of the large number of cycles that exist among states, edges
+in the interaction graph form an intricate web of possible action
+sequences. We believe this in practice greatly reduces usefulness of
+typical network analysis metrics, such as \emph{betweenness,
+  eccentricity, page-rank, eigenvalue} centrality measures.
+
+% \begin{figure}
+%   \centering
+%   \includegraphics[width=\linewidth]{figures/gom.png}
+%   \caption{Interaction graph obtained from the Gmail model.}
+%   \label{fig:gom}
+% \end{figure}
 
@@
 S\footnote{There is no need to disclose the actual brand.}, where the
-driver can engage the system, and once it is engaged, speed can be
+driver can engage the system, and once it is engaged, set speed can be
 increased or decreased with small or large steps. Of course the system
@@
 System A is more elaborate, as it includes also a memory function:
-when it is disengaged it remembers the current speed, which can be
+when it is disengaged it remembers the currently set speed, which can be
 recalled later on.  There are two ways to re-engage it: one by setting
@@
 to approximately get the same speed), the comparison produces the
-values shown in Table~\ref{tab:data}. 
-System A has 2 alternative optimal paths, their length is 4.5 (36\%
-shorter than in S); both systems have a comparable number of detour 1
-traces, but again their length in A is  28\% shorter. The probability
-of hitting the optimal trace is however twice as large in S, which
-features also a more compact design. 
-In both cases detours are caused by the possibility of disengaging the
-system at the wrong moment.
+values shown in Table~\ref{tab:data}.  System A has 2 alternative
+optimal paths, their length is 4.5 (36\% shorter than in S); both
+systems have a comparable number of detour 1 traces ($N_1$), but again
+their length in A ($L_1$) is 28\% shorter. The probability of hitting
+the optimal trace is however twice as large in S, which features also
+a more compact design.  In both cases detours are caused by the
+possibility of disengaging the system at the wrong moment.
 
@@
   because more possibilities are offered;
-\item system A features a more compact design, with fewer action types
+\item system S features a more compact design, with fewer action types
   to be performed at each moment.
@@
 are likely to produce different models for the same UI. As a
-consequence, it is possible that metrics computed by \migtool\ do depend on different
-modeling choices. It is worth mentioning, though, that because of the
-flattening process, several differences are reduced (for example,
-those dealing with using a different hierarchy of states, or with
-distributing differently concurrent regions across states), and 
-sensitivity is correspondingly reduced. At the moment, however, we do
-not have hard evidence to back this claim.
+consequence, it is possible that metrics computed by \migtool\ do
+depend on different modeling choices. It is worth mentioning, though,
+that because of the flattening process, several differences are
+reduced (for example, those dealing with using a different hierarchy
+of states, or with distributing differently concurrent regions across
+states), and sensitivity is correspondingly reduced. % At the moment,
+% however, we do not have hard evidence to back this claim.
+Future studies for addressing this issue could follow what was done in
+\cite{john10}.
 
 Differently from other model-based approaches, such as those based on
-IFML, \migtool\ uses \emph{only} a model of the behavior of the
-UI. Designers do not need to cope with data modeling, nor with
-decisions dealing with presentation. In a sense, \migtool\ uses only the
-\emph{controller} part of the Model-View-Controller paradigm that is
-adopted when developing UIs. Therefore, the designer using \migtool\ to
-perform analysis is free from other concerns that in the end affect
-usability, and the conclusions that are derived with \migtool\ can be
-combined with other results \emph{after} the analysis is
-performed.
+IFML, \migtool\ uses \emph{only} a model of the behavior of the UI. In
+IFML the modeler specifies behavior in terms of concepts that are very
+close to statecharts (IFML adopts a dual representation of
+AND/OR-states, has guards, has events, has an underlying event
+broadcasting mechanism) upon which data binding is applied, so that
+transitions can be easily bound to data instances. On the other hand,
+designers using \migtool\ do not need to cope with data modeling, nor
+with decisions dealing with presentation. In a sense, \migtool\ uses
+only the \emph{controller} part of the Model-View-Controller paradigm
+that is adopted when developing UIs. Although in principle the same
+results could be obtained from IFML models, to do so one needs to
+isolate the control part of the IFML model from the data part; this is
+difficult to do because, by design, IFML integrates very well the two
+aspects.
+
+A designer using \migtool\ is free from other concerns that in the end
+affect usability, and the conclusions that are derived with \migtool\
+can be combined with other results \emph{after} the analysis is
+performed. As mentioned above, ours is an agile modeling approach
+\cite{ambler02} that:
+\begin{inparaenum}[(a)]
+\item enables analysis of a UI well before it is developed;
+\item  requires \emph{only} a model of its behavior;
+\item  supports exploration of different tasks/scenarios;
+\item enables grounding of tasks/scenarios through a
+  platform-independent specification. As such, grounding can be viewed
+  as the specification of a \emph{reification} relation between a
+  \emph{task-and-domain-concepts} level to an
+  \emph{abstract-user-interface} level (we are referring to the
+  CAMELEON reference model \cite{calvary:iwc03:camaleon}).
+\end{inparaenum}
+It is possible to extend \migtool\ so that also task models
+(specified, for example, through CTT, UsiXML, HAMSTERS
+\cite{mori02,limbourg04:usixml,martinie15}) can be processed and used
+as specification of the scenarios to be analyzed. At that point,
+\migtool\ could be used as a tool to explore suitability of tasks with
+respect to a behavioral model of a user interface. This problem has
+been addressed in \cite{martinie15} through a mechanism that relates
+task models to source code annotations of the implementation of the
+user interface. In this way it is possible to simulate execution of
+the tasks in parallel with execution of the system and ensure
+compatibility of the system with respect to tasks. With an extended
+version of \migtool\ it could be done without having to implement the
+system.
 
@@
 more specific analyses that could be performed, for example, with
-CogTool.
+CogTool or CogTool-Helper \cite{swearngin13}.
 
@@
 to compare different designs. We opted for non doing it in light of
-the fact that lostness was designed for exploratory activities, not
-for goal oriented scenarios like the ones we considered in previous
-examples.
-
-\migtool\ is part of a suite of model-based tools for designing,
-analyzing and testing user interfaces developed by the company
-ANONYMIZED. \migtool\ is implemented partly in Java (model processing)
+the fact that lostness was designed for exploratory activities in
+information websites, not for goal oriented scenarios like the ones we
+considered in previous examples.
+
+% \migtool\ is part of a suite of model-based tools for designing,
+% analyzing and testing user interfaces developed by the company
+% ANONYMIZED. 
+\migtool\ is implemented partly in Java (model processing)
 and partly in R/iGraph. Development of a web-based UI of \migtool\ is
-underway; it will be freely available for research purposes.
+underway; it will be freely available for research purposes. Currenlty
+\migtool\ reads UML models represented as XMI files; it could be
+easily extended to handle also SCXML representations of statecharts \cite{scxml}.
 
@@
 
-% REFERENCES FORMAT
-% References must be the same font size as other body text.
 \bibliographystyle{SIGCHI-Reference-Format}

### 2016-01-18 paper.tex
submission ready version

@@
   user navigation errors, can then be computed.  For example, when
-  analyzing four mail applications, we show that Gmail compared to
-  others has 3 times more shortest routes, has 2x more routes that
-  include a single user error, has routes with 13\% fewer steps, but
-  has also optimal routes with the smallest probability to be chosen.
+  analyzing four mail applications, we show that Gmail has 3 times
+  more shortest routes, has twice more routes that include a single user
+  error, has routes with 13\% fewer steps, but has also optimal routes
+  with the smallest probability to be chosen.
 
@@
   with respect to different scenarios, or comparison of different
-  design alternatives, can be done after providing only a partial
-  specification of the user interface behavior.
+  design alternatives against a single scenario, can be done with just
+  a partial specification of the user interface behavior.
 
@@
   follow, and by computing a range of diverse metrics on these
-  results. The agile modeling approach that was followed implies that
-  relatively little work is needed upfront.
+  results.
 
@@
 We present an approach that allows a designer to assess interaction
-design (IxD) qualities such as efficiency, error-proneness and
+design qualities such as efficiency, error-proneness and
 recovery from errors. Key importance is given to the ability to
@@
 %
-Adopting Usage Centered Development (UCD) practices is effective, as is
-following established design principles~\cite{constantine99}. In
-particular, one of the most effective technique is early
-prototyping to explore part of the five-dimensional fidelity
-prototyping space~\cite{mccurdy06,buxton07}. It is particularly effective when
+Adopting Usage Centered Development (UCD) practices is effective, as
+is following established design principles~\cite{constantine99}. In
+particular, one of the most effective technique is early prototyping
+to explore part of the five-dimensional fidelity prototyping
+space~\cite{mccurdy06,buxton07}. This is particularly effective when
 paired with usability investigations based on user testing or
@@
 
-However, UCD requires prototypes which are usually developed with
+However,  prototypes are usually developed with
 certain tasks in mind, and therefore are quite restricted in terms of
@@
 moment, these straightforward questions are quite complicated to
-answer. In fact, they require development of prototypes, inspecting
+answer. In fact, they require developing  prototypes, inspecting
 them, manual tracking of which screens and widgets are used at which
-stage, and a systematic manual analysis.
+stage, followed by a systematic manual analysis.
 
 This should not be the case, however, because answers to these
-questions can be automatically computed. In this way a system can
+questions could be automatically computed. In this way a system could
 provide important insights to a designer, and support assessments of
@@
 (when users are supposed to make 0 or 1 error); when users make 2 or
-more errors the number of paths drops significantly, reducing the
-error-proneness of the UI; on the best case, Gmail features also the
-shortest paths, requiring 13\% fewer steps; however, the probability
-that a user hits an optimal path with Gmail is 10 times smaller than
-the best of the other applications, and the probability that a random
-walker hits a state that is not due to an error is 10\% smaller than
-the best of the other applications.
+more errors the number of paths drops significantly, which indicates
+that the error-proneness of the UI has reduced; in the best case,
+Gmail features also the shortest paths, requiring 13\% fewer steps;
+however, the probability that a user hits an optimal path with Gmail
+is 10 times smaller than the best of the other applications, and the
+probability that a random walker hits a state that is not due to an
+error is 10\% smaller than the best of the other applications.
 %
@@
 slightly different interaction structures adopted for uploading
-messages. If Gmail didn't exist yet, by using \migtool\ its designers
-could obtain these answers well before developing prototypes and
-performing usability studies.
-
-Other examples shown below show what is gained when a new feature is
-added to an embedded system.
+attachments. Thus, redesigning the methods with which competing
+applications deal with attachments could improve their usability. If
+Gmail didn't exist yet, by using \migtool\ its designers could obtain
+these answers well before developing prototypes and performing
+usability studies.
+
+Other examples discussed below show what is gained when a new feature
+is added to an embedded automotive system.
 
@@
 discusses approaches for measuring user interfaces in the context of
-usability and on the basis of a state transition model. This resulted
-in 78 full-text papers that were later on re-analyzed against the same
-criterion, leading to several of papers mentioned below.
+usability and whether it relies on a state transition model. This
+resulted in 78 full-text papers that were later on re-analyzed against
+the same criterion, leading to several of papers mentioned below.
 %
@@
 draw usability conclusions dates back at least to~\cite{parnas69}. In
-it, Parnas claimed that several kinds of usability problems would not
-occur if the designer adopted a design framework where states and
+it, Parnas claimed that several kinds of usability problems could be
+avoided if the designer adopted a design framework where states and
 their transitions are made explicit.
@@
 This idea was later on expanded by Thimbleby~\cite{thimbleby07};
-statecharts are seen as crucial representations that allow a designer to
+statecharts are seen as  representations that allow a designer to
 fully appreciate how devices behave. The overall stance is that ``If
 you don't understand the logic conveyed by a statechart model of a
-user interface, then you don't understand the behavior of the user
+user interface, then you don't understand the behavior of that user
 interface''.
@@
 (to display details of a item, to display or select lists of items, to
-input an item). None of these approaches focus directly on measures of
-usability.
-
-A rather different route for the problem of generating UIs is followed
+input an item). None of these approaches, however, focus directly on
+measures of usability.
+
+A  different route for the problem of generating UIs is followed
 in~\cite{heymann2007formal}. Authors assume that the UI to be generated
@@
 
-Lostness~\cite{smith1996lostness,otter2000lostness} is one of the few
-metrics for measuring the degree to which users become lost in the
+Lostness~\cite{smith1996lostness,otter2000lostness} is a
+metric for measuring the degree to which users become lost in the
 information space, and therefore considers the notion of
@@
 models can be used to specify alternative ways to achieve te same
-goal, and analyse the consequences on the reliability of the resulting
-system. Compared to MIGTool, that approach cannot take advantage of a
-model of the system and has instead to rely only on models of the
+goal, and analyse the consequences on safety and reliability of the resulting
+system. Compared to MIGTool, this approach cannot take advantage of a
+model of the system and  instead relies only on models of the
 tasks. The conseguence is that analysing alternative ways to achieve
@@
 out individual interaction steps depends on many aspects, such as the
-time to decide what goal to achieve, to understand how to do it, to
-locate the actions to do it, to perform the action, to attend to the
-feedback. Layout of the user interface, Fitt's law considerations,
-cognitive state of the user, physical and environmental aspects affect
-such times. For these reasons we opted, in \migtool, to only focus on
-required steps to complete a task, not on completion time.
+time to decide what goal to achieve, to understand how to carry it
+out, to locate the actions to do it, to perform the action, to attend
+to the feedback. Layout of the user interface, Fitt's law
+considerations, cognitive state of the user, physical and
+environmental aspects affect such times. For these reasons we opted,
+in \migtool, to only focus on required steps to complete a task, not
+on completion time.
 
 A usability analysis method capable to analytically predict task
-completion times from a storyboard of the UI is
+completion times from a storyboard of the UI is based on using
 CogTool~\cite{bellamy2011cogtool}. CogTool relies on the ACT-R
@@
 modality and which widgets are used to implement actions, the designer
-gets an estimation of how long an experienced user would spend on each
+gets an estimation of how long a skilled user would spend on each
 step. CogTool takes care of adding extra ``mental'' steps, such as
@@
 steps. Our method is less precise: it does not provide expected
-completion times that are cognitively justified. However, with our
-method, you can analyze a large part of the UI, get information about
-possible problems in some areas, and only then devolve more resources
-in building storyboards and in making assumptions regarding widgets so
-that specific execution paths previously identified can be analyzed
-with CogTool. In a sense, the output obtained with our method could be
-used to make informed decisions as to what to analyze next with
-CogTool. Approaches like \cite{swearngin13} start from the assumption
-that the user interface exists already, and that it can \emph{ripped}
-to reverse engineer a behavior model. These assumptions are not needed
-to use \migtool.
+completion times. However, with our method, a designer can analyze a
+large part of the UI, get information about possible problems in some
+areas, and only then devolve more resources in building storyboards
+and in making assumptions regarding widgets so that specific execution
+paths previously identified can be analyzed with CogTool. In a sense,
+the output obtained with our method could be used to make informed
+decisions as to what to analyze next with CogTool. Approaches like
+\cite{swearngin13} start from the assumption that the user interface
+exists already, and that it can \emph{ripped} to reverse engineer a
+behavior model. These assumptions are not needed to use \migtool.
 
@@
 cover several devices, ranging from a simple torch (with 4 states), a
-microwave cooker (6 states), a mobile phone (152 states); those are
+microwave cooker (6 states), a mobile phone (152 states). Notice that those are
 all push-down devices with a fixed set of buttons. This is obviously
@@
 In our work we automatically generate graphs from statechart models,
-and on some of them computed these metrics. We were not able to draw
+and on some of them we computed these metrics. We were not able to draw
 sensible conclusions from the values we obtained (for example from the
@@
 this is, in a sense, the inverse problem of the one we tackle in this
-paper: we want to compute a subset of the possible streams of user
+paper: we want to compute a subset of the \emph{possible} streams of user
 events given a specification of the UI, rather than trying to abstract
-general properties from streams of events.
+general properties from \emph{actual} streams of events.
 
@@
 In this section we provide a description of how \migtool\ computes
-execution traces.  They are paths (\ie, sequences of connected states
-of the model) that users can follow when performing a given scenario.
-The generation process encompasses the following steps:
+execution traces.  They are potential paths (\ie, sequences of
+connected states of the model) that users can follow when performing a
+given scenario.  The generation process encompasses the following
+steps:
 \begin{inparaenum}[(1)]
@@
 of finite state automata (FSA), with an expressive language that
-includes hierarchic levels of abstraction, concurrent regions, states
+includes hierarchical levels of abstraction, concurrent regions, states
 and pseudostates, guards, and an extended state notion based on an
@@
 
-Because in all but the most trivial interaction graphs there are
-cycles, the set of possible interaction traces is infinite. For this
-reason, scenarios need to be defined and used as constraints on the
-possible execution traces that can be generated.  Users of \migtool\
-define interaction scenarios by specifying key steps (called
+Because in all but the most trivial interaction graphs there are loops
+or cycles, the set of possible interaction traces is infinite. For
+this reason, scenarios need to be defined and used as constraints on
+the possible execution traces that can be generated.  Users of
+\migtool\ define interaction scenarios by specifying key steps (called
 \emph{bridge sets}) that users are expected to go through; we call
@@
 set is specified by selecting a subset of the edges of the interaction
-graph. In general a specification includes an initial state and a
+graph. In general a specification of a scenario includes an initial state and a
 sequence of bridge sets. For example, to specify a scenario for
@@
 
+\newpage
 \subsection{Searching traces}
@@
 To cope with this we implemented a trace searching algorithm that
-processes each of the stages sequentially, starting from the initial
+processes each of the stages sequentially, starting from the
 initial state. Given a bridge set $B_i$, the algorithm does a
@@
 paths, connecting the initial state to reachable bridges of the
-last bridge set of the scenario are called \emph{execution traces}.
+last bridge set of the scenario, are called \emph{execution traces}.
 
@@
 creates traces with detour order $k+1$ by collecting the set $D_k$ of states
-with order $0$ up to $k$, and by identifying the neighbors ($N_k$)
+with order $0$ up to $k$, and by identifying the neighbors $N_k$
 of $D_k$ ($N_k$ is the set of states  not included in $D_k$ that are
@@
 $D_k$ with $N_k$ represent deviations that users might follow, and
-geodesic paths between $N_k$ and $D_k$ constitute recovery paths that
+geodesic paths from $N_k$ to $D_k$ constitute recovery paths that
 users might follow to complete the scenario from states in
@@
 for some state in $N_k$ there is no path leading to any vertex in
-$D_k$: in such a case the deviation leads to a dead end that prevents
+$D_k$: in such a case the deviation is a dead end that prevents
 the user to complete the scenario.
@@
 edges. Therefore it scales well with the size of the
-interaction graph and/or complexity of the scenarios. In practice for
+interaction graph and/or complexity of the scenarios. In practice, for
 interaction graphs consisting of about 10000 edges and a dozen of
@@
 
-Notice that this behavior ``happens'' in one of the two concurrent
-regions specified by this model. In parallel to this, the user can
-either be reading messages (state \texttt{reading}) or may be
-composing a new message (state \texttt{composing}). In the latter case
-the user can independently add recipients, attachments, write the body
-or subject of the message; and send or cancel it.
+This behavior ``happens'' in one of the two concurrent regions
+specified by this model (left and right side of
+Figure~\ref{fig:model}). In parallel to this, the user can either be
+reading messages (state \texttt{reading}) or may be composing a new
+message (state \texttt{composing}). In the latter case the user can
+independently add recipients, attachments, write the body or subject
+of the message; and send or cancel it.
 
@@
 The model that we show here is part of what we used in the examples
-reported below. The actual model that we used consists of 
-24 states, 12 pseudo states, 12 regions, and 61 transitions. The model
-is similar to that being discussed in examples mentioned by \cite{ifml13}.
+reported below. The actual model that we used consists of 24 states,
+12 pseudo states, 12 regions, and 61 transitions. This model is
+similar to that being discussed in IFML examples mentioned by
+\cite{ifml13}.
 
@@
 in the interaction graph form an intricate web of possible action
-sequences. We believe this in practice greatly reduces usefulness of
+sequences. We believe this greatly reduces usefulness of
 typical network analysis metrics, such as \emph{betweenness,
@@
 3, consist of a graph with 474 vertices and 1753 edges. These traces
-entail 12 possible geodesic paths with order 0, 82 with order 1, 30
+entail 12  geodesic paths with order 0, 82 with order 1, 30
 with order 2 and 33 with order 3. Figure~\ref{fig:numpaths} shows the
@@
 To combine these two results, we can easily compute the frequency of
-paths having different length. Figure~\ref{fig:freqopt} shows, for
-each application, the frequency of order 0 paths, the frequency of
-paths with length less or equal to 15 (the minimum length across the
-four applications), and the frequency of optimal paths (the shortest
-ones). These values show that Gmail users have the lowest probability
-to hit an order 0 path (because of the relative large number of order
-1 paths made available by Gmail), have the lowest probability of
-hitting the shortest paths (10 times smaller than the best of the
-other applications), but have the highest probability to hit a
-path with length 15 or less.  Thus, the flexibility and efficiency
-that can be exploited with Gmail are counterbalanced by the required
-knowledge and capability of choosing an optimal path. In particular,
-Gmail offers many detours of order 1 which increase flexibility for
-some users and might decrease effectiveness for less knowledgeable
-ones.
+paths having different length within an
+application. Figure~\ref{fig:freqopt} shows, for each application, the
+frequency of order 0 paths, the frequency of paths with length less or
+equal to 15 (the minimum length across the four applications), and the
+frequency of optimal paths (the shortest ones). These values show that
+Gmail users have the lowest probability to hit an order 0 path
+(because of the relative large number of order 1 paths made available
+by Gmail), have the lowest probability of hitting the shortest paths
+(10 times smaller than the best of the other applications), but have
+the highest probability to hit a path with length 15 or less.  Thus,
+the flexibility and efficiency that can be exploited with Gmail are
+counterbalanced by the required knowledge and capability of choosing
+an optimal path. In particular, Gmail offers many detours of order 1
+which increase flexibility for some users and might decrease
+effectiveness for less skilled ones.
 
@@
 probability of making at most 1 error is approximately the
-same\footnote{These results are not very sensitive to the value chosen
+same\footnote{These results are not sensitive to the value chosen
   for the damping factor: differences across the four applications are
@@
 Manual inspection of the shortest paths indicates that one reason of
-the higher potential efficiency offered by Gmail is due to the
+the greated potential efficiency offered by Gmail is due to the
 fact that users can start composing a new message while reading a
@@
 applications, \ie\ the average number of actions per state involved
-in traces, and average number of \emph{unique} actions per state. The
-former is an overall measure of the number of options that are made
+in traces defined by the out-degree, and average number of \emph{unique} actions per state. The
+former is an overall measure of the number of actions that are made
 available by a UI, the latter can be used to analyze how many
@@
 
-Focusing now on Horde, it turns out that composing a new message
+Let's focus now on Horde. It turns out that composing a new message
 entails many more order 0 paths (154 vs 86), and a comparable number
@@
 adding a widget or feature to an existing UI.  For example, we studied
-some cruise control features of cars. One of the examples is system
+cruise control features of cars. One of the examples is system
 S\footnote{There is no need to disclose the actual brand.}, where the
@@
 increased or decreased with small or large steps. Of course the system
-can also be disengaged (by pressing the brake pedal, for example).
-System A is more elaborate, as it includes also a memory function:
-when it is disengaged it remembers the currently set speed, which can be
-recalled later on.  There are two ways to re-engage it: one by setting
-a new speed, and one by recalling the previous one. In addition, if
-the car drives for more than 5 minutes at a higher speed than the set
-one, system A automatically disengages.
+can also be disengaged (by pressing the brake pedal or by acting on a
+lever).  System A is more elaborate, as it includes also a memory
+function: when it is disengaged it remembers the currently set speed,
+which can be recalled later on.  There are two ways to re-engage it:
+one by setting a new speed, and one by recalling the previous one. In
+addition, if the car drives for more than 5 minutes at a higher speed
+than the set one, system A automatically disengages.
 
@@
 is done manually by the driver with 4 actions on ``up'' and ``down''
-to approximately get the same speed), the comparison produces the
+(to approximately set the previous speed), the comparison produces the
 values shown in Table~\ref{tab:data}.  System A has 2 alternative
@@
 \begin{inparaenum}[(1)]
-\item system A makes users more efficient (an increase of 56\%) for
-  the considered scenario;
+\item system A makes users more efficient (a saving of 36\%) for the
+  considered scenario;
 \item with A there are two possible ways to achieve the scenario, thus
@@
   because more possibilities are offered;
-\item system S features a more compact design, with fewer action types
+\item system S features a more compact design, with fewer action 
   to be performed at each moment.
@@
 industrial examples, is that models do not need to be complete
-representations of the behavior of the application under
-study. By following an agile modeling approach~\cite{ambler02}, models can
-be easily developed by one person in less than one day, using any UML
+representations of the behavior of the application under study. By
+following an agile modeling approach~\cite{ambler02}, models can be
+easily developed by one person in less than one day, using any UML
 capable design tool. Even more complex models (in our experience up to
-150 states and 350 transitions) can be developed and verified in 2-3
-days by one person.
-Experience in using statecharts to model behavior of UIs is needed
-though. Useful suggestions are given in~\cite{horrocks99,thimbleby07}.
+200 states and 450 transitions) can be developed and verified in 3
+days by one individual.  Experience in using statecharts to model
+behavior of UIs is needed though; useful suggestions are given
+in~\cite{horrocks99,thimbleby07}.
 
@@
 IFML~\cite{ifml13}. Even though there are fundamental limits
-(inability to handle undo/redo's - because that goes beyond a finite
-state representation; inability to handle customizable toolbars -
+(inability to handle undo/redo's, because that goes beyond a finite
+state representation; inability to handle customizable toolbars,
 because that requires models that change at runtime; inability to
-handle perceptive UIs - because they are not well suited to be
+handle perceptive UIs, because they are not well suited to be
 described in terms of discrete states), in many practical cases they
-can be be isolated and/or ignored.
+can be be isolated and/or ignored~\cite{gb15:icwe}.
 
@@
 reduced (for example, those dealing with using a different hierarchy
-of states, or with distributing differently concurrent regions across
+of states, or with differently distributed concurrent regions across
 states), and sensitivity is correspondingly reduced. % At the moment,
@@
 AND/OR-states, has guards, has events, has an underlying event
-broadcasting mechanism) upon which data binding is applied, so that
-transitions can be easily bound to data instances. On the other hand,
-designers using \migtool\ do not need to cope with data modeling, nor
-with decisions dealing with presentation. In a sense, \migtool\ uses
-only the \emph{controller} part of the Model-View-Controller paradigm
-that is adopted when developing UIs. Although in principle the same
-results could be obtained from IFML models, to do so one needs to
-isolate the control part of the IFML model from the data part; this is
-difficult to do because, by design, IFML integrates very well the two
-aspects.
+broadcasting mechanism) upon which components with their data binding
+are defined, so that transitions can be easily bound to data
+instances. On the other hand, designers using \migtool\ do not need to
+cope with data modeling, nor with decisions dealing with
+presentation. In a sense, \migtool\ uses only the \emph{controller}
+part of the Model-View-Controller paradigm that is adopted when
+developing UIs. Although in principle the same results could be
+obtained from IFML models, to do so one needs to isolate the control
+part of the IFML model from the data part; this is difficult to do
+because, by design, IFML integrates very well the two aspects.
 
@@
   \emph{task-and-domain-concepts} level to an
-  \emph{abstract-user-interface} level (we are referring to the
+  \emph{abstract-user-interface} level (we are using terms of the
   CAMELEON reference model \cite{calvary:iwc03:camaleon}).
@@
 the tasks in parallel with execution of the system and ensure
-compatibility of the system with respect to tasks. With an extended
-version of \migtool\ it could be done without having to implement the
-system.
-
-In terms of validity of conclusions obtained through \migtool, we can
-say that because they are devoid of user behavior assumptions (such as
-preferences, skills, interpretations, ergonomic constraints) they are
-very general and conservative. On the other hand, predictions based on
-\migtool\ metrics are also generic because they do not consider data
-and presentation aspects, nor user-related factors. For example, it is
+compatibility of the system with respect to tasks.
+
+In terms of validity of conclusions obtained through \migtool, because
+they are devoid of user behavior assumptions (such as preferences,
+skills, interpretations, ergonomic constraints) they are very general
+and conservative. On the other hand, they are also generic because
+they do not consider data and presentation aspects. For example, it is
 unfeasible to use \migtool\ to predict the time needed by a user to
@@
 weight, and therefore in such a way to perform analyses that are
-similar to the ones suggested in~\cite{thimbleby2001markov}.
+similar to the ones based on Markov chains suggested
+in~\cite{thimbleby2001markov}.
 
@@
 and partly in R/iGraph. Development of a web-based UI of \migtool\ is
-underway; it will be freely available for research purposes. Currenlty
+underway; it will be freely available for research purposes. Currently
 \migtool\ reads UML models represented as XMI files; it could be
@@
 how they could perceive that, understand that, and manipulate that.
-But, because the approach can be applied also when only a
+But, because the approach can be applied  when just a
 specification of the user interface is available, it supports
 construction and engineering of interactive systems and could help in
-iteratively simplifying or improving a design.
-
-\bibliographystyle{SIGCHI-Reference-Format}
-\bibliography{gb,database}
+iteratively improving a design before building any UI prototype.
+
+%\bibliographystyle{SIGCHI-Reference-Format}
+%\bibliography{gb,database}
+%%% -*-BibTeX-*-
+%%% Do NOT edit. File created by BibTeX with style
+%%% ACM-Reference-Format-Journals [18-Jan-2012].
+
+\begin{thebibliography}{00}
+
+%%% ====================================================================
+%%% NOTE TO THE USER: you can override these defaults by providing
+%%% customized versions of any of these macros before the \bibliography
+%%% command.  Each of them MUST provide its own final punctuation,
+%%% except for \shownote{}, \showDOI{}, and \showURL{}.  The latter two
+%%% do not use final punctuation, in order to avoid confusing it with
+%%% the Web address.
+%%%
+%%% To suppress output of a particular field, define its macro to expand
+%%% to an empty string, or better, \unskip, like this:
+%%%
+%%% \newcommand{\showDOI}[1]{\unskip}   % LaTeX syntax
+%%%
+%%% \def \showDOI #1{\unskip}           % plain TeX syntax
+%%%
+%%% ====================================================================
+
+\ifx \showCODEN    \undefined \def \showCODEN     #1{\unskip}     \fi
+\ifx \showDOI      \undefined \def \showDOI       #1{{\tt DOI:}\penalty0{#1}\ }
+  \fi
+\ifx \showISBNx    \undefined \def \showISBNx     #1{\unskip}     \fi
+\ifx \showISBNxiii \undefined \def \showISBNxiii  #1{\unskip}     \fi
+\ifx \showISSN     \undefined \def \showISSN      #1{\unskip}     \fi
+\ifx \showLCCN     \undefined \def \showLCCN      #1{\unskip}     \fi
+\ifx \shownote     \undefined \def \shownote      #1{#1}          \fi
+\ifx \showarticletitle \undefined \def \showarticletitle #1{#1}   \fi
+\ifx \showURL      \undefined \def \showURL       #1{#1}          \fi
+
+\bibitem{ambler02}
+{S. Ambler}. 2002.
+\newblock {\em Agile Modeling: Effective Practices for eXtreme Programming and
+  the Unified Process}.
+\newblock Wiley.
+\newblock
+\showISBNx{0471202827}
+
+\bibitem{appert2008swingstates}
+{C. Appert} {and} {M. Beaudouin-Lafon}. 2008.
+\newblock \showarticletitle{SwingStates: adding state machines to Java and the
+  Swing toolkit}.
+\newblock {\em Software: Practice and Experience\/} {38}, 11 (2008),
+  1149--1182.
+\newblock
+
+\bibitem{conversy07}
+{E. Barboni}, {S. Conversy}, {D. Navarre}, {and} {P. Palanque}. 2007.
+\newblock \showarticletitle{Model-based engineering of widgets, user
+  applications and servers compliant with ARINC 661 specification}.
+\newblock In {\em Interactive Systems. Design, Specification, and
+  Verification}. Springer, 25--38.
+\newblock
+
+\bibitem{scxml}
+{J. Barnett} {and} {et al}. 2015.
+\newblock {\em State Chart {XML (SCXML)}: State Machine Notation for Control
+  Abstraction}.
+\newblock {T}echnical {R}eport. W3C.
+\newblock
+\newblock
+\shownote{\url{http://www.w3.org/TR/2015/REC-scxml-20150901}.}
+
+\bibitem{bellamy2011cogtool}
+{R. Bellamy}, {B. John}, {and} {S. Kogan}. 2011.
+\newblock \showarticletitle{Deploying CogTool: integrating quantitative
+  usability assessment into real-world software development}. In {\em Software
+  Engineering (ICSE), 2011 33rd International Conference on}. IEEE, 691--700.
+\newblock
+
+\bibitem{gb15:icwe}
+{G. Brajnik} {and} {S. Harper}. 2015.
+\newblock \showarticletitle{Detaching control from data models in model-based
+  generation of user interfaces}. In {\em Proc. of Int. Conf. on Web
+  Engineering}, {F.~Frasincar} (Ed.). IEEE, Rotterdam, Netherlands.
+\newblock
+
+\bibitem{briand2005:test:uml}
+{L.~C Briand}, {Y. Labiche}, {and} {J. Cui}. 2005.
+\newblock \showarticletitle{Automated support for deriving test requirements
+  from UML statecharts}.
+\newblock {\em Software \& Systems Modeling\/} {4}, 4 (2005), 399--423.
+\newblock
+
+\bibitem{buxton07}
+{B. Buxton}. 2007.
+\newblock {\em User Experience: Getting the Design Right and the Right Design}.
+\newblock Morgan Kaufmann.
+\newblock
+
+\bibitem{calvary:iwc03:camaleon}
+{G. Calvary}, {J. Coutaz}, {D. Thevenin}, {Q. Limbourg}, {L. Bouillon}, {and}
+  {J. Vanderdonckt}. 2003.
+\newblock \showarticletitle{A Unifying Reference Framework for Multi-Target
+  User Interfaces}.
+\newblock {\em Interacting with Computers\/} {15}, 3 (2003), 289--308.
+\newblock
+
+\bibitem{ceri:00}
+{S. Ceri}, {P. Fraternali}, {and} {A. Bongio}. 2000.
+\newblock \showarticletitle{{Web Modeling Language (WebML)}: a modeling
+  language for designing web sites}.
+\newblock {\em Computer Networks\/}  {33} (2000), 137--157.
+\newblock
+
+\bibitem{constantine99}
+{L.L. Constantine} {and} {L.A.D. Lockwood}. 1999.
+\newblock {\em Software for use: a practical guide to the models and methods of
+  usage-centered design}.
+\newblock Addison-Wesley.
+\newblock
+
+\bibitem{fayollas2014}
+{C. Fayollas}, {C. Martinie}, {P. Palanque}, {Y. Deleris}, {J-C. Fabre}, {and}
+  {D. Navarre}. 2014.
+\newblock \showarticletitle{An approach for assessing the impact of
+  dependability on usability: application to interactive cockpits}. In {\em
+  Tenth European Dependable Computing Conference (EDCC), 2014}. IEEE, 198--209.
+\newblock
+
+\bibitem{pirolli2007}
+{W.-T. Fu} {and} {P. Pirolli}. 2007.
+\newblock \showarticletitle{{SNIF-ACT}: A cognitive model of user navigation on
+  the World Wide Web}.
+\newblock {\em Human--Computer Interaction\/} {22}, 4 (2007), 355--412.
+\newblock
+
+\bibitem{gomez01}
+{J. G\'{o}mez}, {C. Cachero}, {and} {O. Pastor}. 2001.
+\newblock \showarticletitle{Conceptual Modeling of Device-Independent Web
+  Applications}.
+\newblock {\em IEEE MultiMedia\/} {8}, 2 (2001), 26--39.
+\newblock
+\showISSN{1070-986X}
+\showDOI{%
+\url{http://dx.doi.org/10.1109/93.917969}}
+
+\bibitem{harel09}
+{D. Harel}. 2009.
+\newblock \showarticletitle{Statecharts in the making: a personal account}.
+\newblock {\em CACM\/} {52}, 3 (March 2009), 67--75.
+\newblock
+
+\bibitem{heymann2007formal}
+{M. Heymann} {and} {A. Degani}. 2007.
+\newblock \showarticletitle{Formal analysis and automatic generation of user
+  interfaces: approach, methodology, and an algorithm}.
+\newblock {\em Human Factors: The Journal of the Human Factors and Ergonomics
+  Society\/} {49}, 2 (2007), 311--330.
+\newblock
+
+\bibitem{hilbert2000}
+{D.~M Hilbert} {and} {D.F. Redmiles}. 2000.
+\newblock \showarticletitle{Extracting usability information from user
+  interface events}.
+\newblock {\em ACM Computing Surveys (CSUR)\/} {32}, 4 (2000), 384--421.
+\newblock
+
+\bibitem{hornbaek2006current}
+{K. Hornb{\ae}k}. 2006.
+\newblock \showarticletitle{Current practice in measuring usability: Challenges
+  to usability studies and research}.
+\newblock {\em International Journal of Human-Computer Studies\/} {64}, 2
+  (2006), 79--102.
+\newblock
+
+\bibitem{horrocks99}
+{I. Horrocks}. 1999.
+\newblock {\em Constructing the User Interface with Statecharts}.
+\newblock Addison-Wesley Longman Publishing Co., Inc., Boston, MA, USA.
+\newblock
+\showISBNx{0201342782}
+
+\bibitem{john10}
+{B.E. John}. 2010.
+\newblock \showarticletitle{Reducing the Variability between Novice Modelers:
+  Results of a Tool for Human Performance Modeling Produced through
+  Human-Centered Design}. In {\em Proc. of the 19th Conf. on Behavior
+  Representation in Modeling and Simulation}. Charleston, SC, USA, 95--102.
+\newblock
+
+\bibitem{koch02:uwe}
+{N. Koch} {and} {A. Kraus}. 2002.
+\newblock \showarticletitle{{The expressive power of {UML}-based web
+  engineering}}. In {\em Second International Workshop on Web-Oriented Software
+  Technology (IWWOST02)}, {D.~Schwabe}, {O.~Pastor}, {G.~Rossi}, {and}
+  {L.~Olsina} (Eds.). 105--199.
+\newblock
+
+\bibitem{limbourg04:usixml}
+{Q. Limbourg}, {J. Vanderdonckt}, {B. Michotte}, {L. Bouillon}, {and} {V.
+  López~Jaquero}. 2004.
+\newblock \showarticletitle{UsiXML: a Language Supporting Multi-Path
+  Development of User Interfaces}. In {\em Proc. of 9th IFIP Working Conference
+  on Engineering for Human-Computer Interaction jointly with 11th Int. Workshop
+  on Design, Specification, and Verification of Interactive Systems,
+  EHCI-DSVIS'2004} {\em (Lecture Notes in Computer Science)}, Vol. 3425.
+  Springer-Verlag, Hamburg, Germany, 200--220.
+\newblock
+
+\bibitem{martinie15}
+{C. Martinie}, {D. Navarre}, {P. Palanque}, {and} {C. Fayollas}. 2015.
+\newblock \showarticletitle{A generic tool--supported framework for coupling
+  task models and interactive applications}. In {\em Proc. of Engineering
+  Interactive Computing Systems, {EICS} '15}. ACM, ACM Press, Duisburg,
+  Germany, 244--253.
+\newblock
+
+\bibitem{mccurdy06}
+{M. McCurdy}, {C. Connors}, {G. Pyrzak}, {B. Kanefsky}, {and} {A. Vera}. 2006.
+\newblock \showarticletitle{Breaking the fidelity barrier: an examination of
+  our current characterization of prototypes and an example of a mixed-fidelity
+  success}. In {\em CHI 2006}. ACM, ACM Press, New York, NY, 1233--1242.
+\newblock
+
+\bibitem{melia08}
+{S. Meli\'a}, {J. G\'omez}, {S. P\'erez}, {and} {O. D\'iaz}. 2008.
+\newblock \showarticletitle{A Model-Driven Development for GWT- Based Rich
+  Internet Applications with OOH4RIA}. In {\em Proc. 8th Int'l Conf. Web Eng.
+  (ICWE 2008)}. IEEE CS Press, 13--23.
+\newblock
+
+\bibitem{mori02}
+{G. Mori}, {F. Patern\`o}, {and} {C. Santoro}. 2002.
+\newblock \showarticletitle{{CTTE: Support for Developing and Analysing Task
+  Models for Interactive System Design}}.
+\newblock {\em IEEE Transactions on Software Engineering\/} {28}, 8 (August
+  2002), 797--813.
+\newblock
+
+\bibitem{ifml13}
+{{OMG}}. 2013.
+\newblock {\em Interaction Flow Modeling Language ({IFML}), FTF -- Beta 1\/}
+  (omg document number: ptc/2013-03-08 ed.).
+\newblock {T}echnical {R}eport. OMG.
+\newblock
+\showURL{%
+\url{http://www.omg.org/spec/IFML/1.0}}
+
+\bibitem{otter2000lostness}
+{M. Otter} {and} {H. Johnson}. 2000.
+\newblock \showarticletitle{Lost in hyperspace: metrics and mental models}.
+\newblock {\em Interacting with Computers\/} {13}, 1 (2000), 1--40.
+\newblock
+
+\bibitem{page1999pagerank}
+{L. Page}, {S. Brin}, {R. Motwani}, {and} {T. Winograd}. 1999.
+\newblock \showarticletitle{The PageRank citation ranking: bringing order to
+  the Web.}
+\newblock  (1999).
+\newblock
+
+\bibitem{parnas69}
+{D.L. Parnas}. 1969.
+\newblock \showarticletitle{On the use of transition diagrams in the design of
+  a user interface for an interactive computer system}. In {\em ACM '69 Proc.
+  of the 1969 24th National Conference}. ACM.
+\newblock
+
+\bibitem{silva03:umli}
+{P. Pinhero~da Silva} {and} {N.W. Paton}. 2003.
+\newblock \showarticletitle{User Interface Modeling in {UMLi}}.
+\newblock {\em IEEE Software\/} (2003), 62--69.
+\newblock
+
+\bibitem{preece02}
+{J. Preece}, {Y. Rogers}, {and} {H. Sharp}. 2002.
+\newblock {\em Interaction Design}.
+\newblock John Wiley and Sons.
+\newblock
+
+\bibitem{rubin08}
+{J. Rubin} {and} {D. Chisnell}. 2008.
+\newblock {\em Handbook of Usability Testing\/} (second ed.).
+\newblock Wiley.
+\newblock
+
+\bibitem{smith1996lostness}
+{P.A. Smith}. 1996.
+\newblock \showarticletitle{Towards a practical measure of hypertext
+  usability}.
+\newblock {\em Interacting with Computers\/} {8}, 4 (1996), 365--381.
+\newblock
+
+\bibitem{swearngin13}
+{A. Swearngin}, {M.B. Cohen}, {B.E. John}, {and} {K.E. Bellamy}. 2013.
+\newblock \showarticletitle{Human Performance Regression Testing}. In {\em
+  Proc. of the 2013 International Conference on Software Engineering} {\em
+  (ICSE '13)}. IEEE Press, Piscataway, NJ, USA, 152--161.
+\newblock
+\showURL{%
+\url{http://dl.acm.org/citation.cfm?id=2486788.2486809}}
+
+\bibitem{thimbleby07}
+{H. Thimbleby}. 2007.
+\newblock {\em Press on: principles of interaction programming}.
+\newblock The MIT Press.
+\newblock
+
+\bibitem{thimbleby2001markov}
+{H. Thimbleby}, {P. Cairns}, {and} {M. Jones}. 2001.
+\newblock \showarticletitle{Usability analysis with Markov models}.
+\newblock {\em ACM Transactions on Computer-Human Interaction (TOCHI)\/} {8}, 2
+  (2001), 99--132.
+\newblock
+
+\bibitem{thimbleby09:sna:eics09}
+{H. Thimbleby} {and} {P. Oladimeji}. 2009.
+\newblock \showarticletitle{Social Network Analysis and Interactive Device
+  Design Analysis}. In {\em Proc. of Engineering Interactive Computing Systems
+  2009}. ACM Press, 91--100.
+\newblock
+
+\bibitem{winckler03}
+{M. Winckler} {and} {P.A. Palanque}. 2003.
+\newblock \showarticletitle{StateWebCharts: A Formal Description Technique
+  Dedicated to Navigation Modelling of Web Applications.}. In {\em Design,
+  Specification, and Verification: 10th International Workshop, DSV-IS'03}.
+  Funchal,Madeira Island, Portugal, 61--76.
+\newblock
+
+\end{thebibliography}
 

### 2016-04-11 paper.tex
implemented changes promised in the rebuttal

@@
 \author{%
-  \alignauthor{1st Author Name\\
-    \affaddr{Affiliation}\\
-    \affaddr{City, Country}\\
-    \email{e-mail address}}\\
-  \alignauthor{2nd Author Name\\
-    \affaddr{Affiliation}\\
-    \affaddr{City, Country}\\
-    \email{e-mail address}}\\
+  \alignauthor{Giorgio Brajnik\textsuperscript{1,2}\\
+    \affaddr{\textsuperscript{1}Dip. di Matematica, Informatica e Fisica}\\
+    \affaddr{Universit\`a di Udine, Italy}\\
+    \email{brajnik@uniud.it}}\\
+  \alignauthor{Simon Harper\textsuperscript{2}\\
+    \affaddr{\textsuperscript{2}Computer Science School}\\
+    \affaddr{University of Manchester, UK}\\
+    \email{simon.harper@manchester.ac.uk}}\\
 }
@@
 \end{inparaenum}
-All this can be done before developing prototypes of the UI.
+All this can be done before developing prototypes of the UI, similarly
+to what is claimed and done in~\cite{silva14}: ``... the analysis of solutions in
+the early phases of development is crucial as their results can then
+be iteratively injected in the development cycle''.
 
@@
 yet several questions cannot be easily answered. We applied the same
-techniques also in other domains, such as HVAC and other embedded UIs.
+techniques also in other domains, such as HVAC (Heat, Ventilation and
+Air Conditioning) and other embedded UIs.
 
@@
 %
-Adopting Usage Centered Development (UCD) practices is effective, as
+Adopting Usage Centered Development practices is effective, as
 is following established design principles~\cite{constantine99}. In
@@
 \begin{inparaenum}[(a)]
-\item the idea of using a UML state machine model to specify
-  interaction scenarios,
+\item the idea of using a UML state machine model to support
+  specification of interaction scenarios,
 \item the development of a tool (\migtool, Measuring Interaction
-  Graphs Tool) that transforms models and scenario specifications into
-  interaction paths, and
+  Graphs Tool) that, based on models, transforms scenario
+  specifications into interaction paths, and
 \item the definition of metrics that provide concise, precise and
-  objective measures of a design.
+  objective measures of the interaction structure entailed by a user
+  interface. 
 \end{inparaenum}
@@
 flight parameter. The paper in particular discusses how these task
-models can be used to specify alternative ways to achieve te same
-goal, and analyse the consequences on safety and reliability of the resulting
-system. Compared to MIGTool, this approach cannot take advantage of a
-model of the system and  instead relies only on models of the
-tasks. The conseguence is that analysing alternative ways to achieve
-the same goal requires developing different models; in addition,
-exploring suitability of a given system with respect to certain tasks
-is complex; and finally, hierarchical task models cannot take
-advantage of contexts and rich control flows that are available with
-state-transition representations, limiting therefore the richness of
-the scenarios being analysed.
+models can be used to specify alternative ways to achieve the same
+goal, and analyses the consequences on safety and reliability of the
+resulting system. Compared to MIGTool, this approach cannot take
+advantage of a model of the system and instead relies only on models
+of the tasks. The conseguence is that analysing alternative ways to
+achieve the same goal requires developing different models; in
+addition, exploring suitability of a given system with respect to
+certain tasks is complex because no model of the system is
+available. Finally, using \emph{only} hierarchical tasks as a modeling
+language limits the richness of the scenarios being analysed, because
+it is difficult to take advantage of contexts and rich control flows
+that are available with state-transition representations, especially
+with UML state diagrams.
 
@@
 
-In this section we provide a description of how \migtool\ computes
-execution traces.  They are potential paths (\ie, sequences of
+Traces are potential paths (\ie, sequences of
 connected states of the model) that users can follow when performing a
@@
 \begin{inparaenum}[(1)]
-\item automatic flattening of the statechart model;
+\item processing the model and automatic flattening of the statechart model;
 \item manual definition of the interaction scenario;
@@
 \migtool\ takes as input UML statecharts, which are a generalization
-of finite state automata (FSA), with an expressive language that
+of finite state automata (FSA) specified with an expressive language that
 includes hierarchical levels of abstraction, concurrent regions, states
 and pseudostates, guards, and an extended state notion based on an
-arbitrary underlying computational model.  Harel, the inventor of
-statecharts, gives an interesting retrospective view of how they were
-invented, back in early eighties, and why they were appealing also to
-non experts~\cite{harel09}.%   For the sake of brevity, we refrain from
-% describing them here in detail; the interested reader is referred to
+arbitrary underlying computational model.  
+% Harel, the inventor of
+% statecharts, gives an interesting retrospective view of how they were
+% invented, back in early eighties, and why they were appealing also to
+% non experts~\cite{harel09}.%   For the sake of brevity, we refrain from
+% % describing them here in detail; the interested reader is referred to
 % the UML standard~\cite{uml25} or some of the textbooks that deal with
@@
 widgets, and transitions to actions performed by users or by the
-system itself. In the classification reported in~\cite{hilbert2000},
-actions belong to the ``abstract interaction events'' category.
+system itself~\cite{horrocks99}. In the classification reported
+in~\cite{hilbert2000}, actions belong to the ``abstract interaction
+events'' category.
 
@@
 to manually inspect the resulting FSA, this aspect is in many cases
-not relevant.  \migtool\ produces an XML representation (graphML) of
-the resulting FSA, the \emph{interaction graph}. It is a directed
-multigraph\footnote{A directed multigraph is a directed graph such
-  that there are 2 or more edges that have the same end points. Cycles
-  are paths that include 2 or more occurrences of the same
-  vertex. Loops are edges that start and end on the same
-  vertex. ``Geodesic path'' is a synonymous term with ``shortest
-  path''.}, potentially with cycles and loops, with edges that are
-labeled with the name of the corresponding action. In the following we
-will be using as synonymous the terms state/vertex, and action/edge.
+not relevant. Because \migtool\ does not use executable UML,
+transition guards are not given semantics and are simply treated as
+part of the event of a transition. \migtool\ produces an XML
+representation (graphML) of the resulting FSA, the \emph{interaction
+  graph}. It is a directed multigraph\footnote{A directed multigraph
+  is a directed graph such that there are 2 or more edges that have
+  the same end points. Cycles are paths that include 2 or more
+  occurrences of the same vertex. Loops are edges that start and end
+  on the same vertex. ``Geodesic path'' is a synonymous term with
+  ``shortest path''.}, potentially with cycles and loops, with edges
+that are labeled with the name of the corresponding action. In the
+following we will be using as synonymous the terms state/vertex, and
+action/edge.
 
@@
 \migtool\ define interaction scenarios by specifying key steps (called
-\emph{bridge sets}) that users are expected to go through; we call
-this process \emph{grounding usage scenarios on models}. Each bridge
-set is specified by selecting a subset of the edges of the interaction
-graph. In general a specification of a scenario includes an initial state and a
-sequence of bridge sets. For example, to specify a scenario for
-replying to an email message, one could select all the edges
-associated to the action \texttt{reply} (bridge set 1), followed by
-edges labeled with \texttt{typeBody} (bridge set 2), followed by edges
-labeled with \texttt{send} (bridge set 3). A well formed bridge set
-consists of 1 or more edges (an empty bridge set would make the
-scenario unviable). In this way scenarios with cycles can be
-formulated (\eg, reply twice to two messages). A \emph{stage} of a
+bridge sets) that users are expected to go through; we call
+this process \emph{grounding usage scenarios on models}. A \emph{bridge set}
+is a subset of the edges of the interaction graph; a well formed
+bridge set is a non-empty set (an empty bridge set would make the
+scenario unviable). In general a specification of a scenario includes
+an initial state and a non-empty sequence bridge sets. For example, to
+specify a scenario for replying to an email message, assuming an
+initial state that corresponds to a user interface that displays all
+messaes in the inbox, one could select all the edges associated to the
+action \texttt{reply} (bridge set 1), followed by edges labeled with
+\texttt{typeBody} (bridge set 2), followed by edges labeled with
+\texttt{send} (bridge set 3).  In this way scenarios with cycles can
+be formulated (\eg, reply twice to two messages). A \emph{stage} of a
 scenario comprises two consecutive bridge sets.
@@
 
-\newpage
 \subsection{Searching traces}
@@
 opening a specific conversation (transition
-\texttt{open(conversation)}). This transition (which is assumed to
-occur when the user clicks on one of the visible conversations) leads
-to the state \texttt{viewingAConversation}, where the behavior of the
-system is defined by a more detailed state machine. By default the
-user is viewing a conversation, but by performing the \texttt{reply}
-action the UI moves to a state called \texttt{replying}, where the
-body of the reply can be typed, the subject can be changed, or another
+\texttt{open(conversation)}). This transition is assumed to occur when
+the user clicks on any one of the visible conversations - it is if
+there is a collection of transitions, each one for a different
+conversation that can be opened; the parameter \texttt{conversation}
+is bound to any occurrence of a conversation. Such transitions lead to the state
+\texttt{viewingAConversation}, where the behavior of the system is
+defined by a more detailed state machine. By default the user is
+viewing a conversation, but by performing the \texttt{reply} action
+the UI moves to a state called \texttt{replying}, where the body of
+the reply can be typed, the subject can be changed, or another
 addressee can be added.
@@
 typical network analysis metrics, such as \emph{betweenness,
-  eccentricity, page-rank, eigenvalue} centrality measures.
+  eccentricity, page-rank} and \emph{eigenvalue} centrality measures.
 
@@
 further the graph, by specifying scenarios and computing traces.  The
-following fragment of R code specifies the bridge sets of the chosen
-scenario: it comprises, in the given order, two occurrences of
-\texttt{open(conversation)}, followed by \texttt{reply}, followed by
-\texttt{typeBody}, etc.: \lstset{language=R}
+scenario, which assumes \texttt{ViewingConversations} is the initial
+state, is specified with the following 11 bridge sets:
+\newpage
+\lstset{language=R}
 \begin{lstlisting}
-edgesWithLabel(gmail,"open(conversation)"),
-edgesWithLabel(gmail,"open(conversation)"), 
-edgesWithLabel(gmail,"reply"),
-edgesWithLabel(gmail,"typeBody"),
-edgesWithLabel(gmail,"send"),
-edgesWithLabel(gmail,"compose"),
-edgesWithLabel(gmail,"addRecipient"),
-edgesWithLabel(gmail,"open(files)"),
-edgesWithLabel(gmail,"writeBody"),
-edgesWithLabel(gmail,"writeSubject"),
-edgesWithLabel(gmail,"send")
+1  open(conversation)
+2  open(conversation)
+3  reply
+4  typeBody
+5  send
+6  compose
+7  addRecipient
+8  open(files)
+9  writeBody
+10 writeSubject
+11 send
 \end{lstlisting}
-In plain language such a scenario means opening a conversation,
-opening a second one, replying to the last message of the
-second conversation by typing the body of the response, sending it, and then
-composing a new message by adding a recipient, an attachment, typing
-the body and subject, and finally sending it.
+In plain language such a scenario means opening a conversation, doing
+something and then opening a second one, replying to the last message
+of the second conversation by typing the body of the response, sending
+it, and then composing a new message by adding a recipient, an
+attachment, typing the body and subject, and finally sending
+it. Notice that because the model precludes the possibility that there
+are two \emph{consecutive} actions called \texttt{open(conversation)},
+MIGTool has to find traces such that between bridge set 1 and 2 there
+is at least one \texttt{goBack} action.
 
@@
 order.
+
+One of the order 0 geodesic paths is shown below; each line
+represents an action that the user is expected to perform; actions are
+only those entailed by the model. Actions marked with ``*'' are those not
+specified in the bridge sets and found in the interaction graph. 
+\lstset{language=Java}
+\begin{lstlisting}
+1  open(conversation) 
+2  goBack *
+3  open(conversation) 
+4  reply 
+5  typeBody 
+6  send 
+7  compose 
+8  attachFiles *
+9  addRecipient[first] 
+10 open(files) 
+11 writeBody 
+12 writeSubject 
+13 send 
+\end{lstlisting}
+% \begin{tabular}{rp{7cm}}
+%     1 & \texttt{open(conversation)} \\
+%     2 & \texttt{goBack} \\ 
+%     3 & \texttt{open(conversation)} \\
+%     4 & \texttt{reply} \\
+%     5 & \texttt{reply, addAddressee, typeBody, changeSubject} \\
+%     6 & \texttt{cancel, send, open(message)[msgIsClosed], close (message) [msgIsOpen]}\\
+%    7 & \texttt{compose} \\
+%     8 & \texttt{attachFiles} \\
+%     9 & \texttt{addRecipient[first]} \\
+%     10  & \texttt{open(files)} \\
+%     11 & \texttt{writeSubject, writeBody} \\
+%     12 & \texttt{writeSubject, close(message)[msgIsOpen], writeBody,
+%          open(message)[msgIsClosed]} \\
+%     13 & \texttt{cancel, send} \\
+%   \end{tabular}
+%   %
 
@@
 UML state diagrams provide a very expressive language, well suited to
-specify behavior of UIs based on discrete events. Evidence of this can
-also be found in recent OMG standards, such as
-IFML~\cite{ifml13}. Even though there are fundamental limits
-(inability to handle undo/redo's, because that goes beyond a finite
-state representation; inability to handle customizable toolbars,
-because that requires models that change at runtime; inability to
-handle perceptive UIs, because they are not well suited to be
-described in terms of discrete states), in many practical cases they
-can be be isolated and/or ignored~\cite{gb15:icwe}.
+specify behavior of UIs based on discrete events. in \migtool\ all the
+UML language~\cite{uml25} for state machine has been used, including sundry
+pseudostates. Evidence of this can also be found in recent OMG
+standards, such as IFML~\cite{ifml13}. Even though there are
+fundamental limits (inability to handle undo/redo's, because that goes
+beyond a finite state representation; inability to handle customizable
+toolbars, because that requires models that change at runtime;
+inability to handle perceptive UIs, because they are not well suited
+to be described in terms of discrete states), in many practical cases
+they can be be isolated and/or ignored~\cite{gb15:icwe}. One
+particular limitation of \migtool\ stems from the fact that it does
+not rely on an executable UML metamodel; as a consequence behaviors
+that rely on the fact that firing a transition broadcasts new events
+cannot be properly represented and analyzed.
 
@@
 \cite{mori02,limbourg04:usixml,martinie15}) can be processed and used
-as specification of the scenarios to be analyzed. At that point,
-\migtool\ could be used as a tool to explore suitability of tasks with
-respect to a behavioral model of a user interface. This problem has
-been addressed in \cite{martinie15} through a mechanism that relates
-task models to source code annotations of the implementation of the
-user interface. In this way it is possible to simulate execution of
-the tasks in parallel with execution of the system and ensure
+as specification of the scenarios to be analyzed. For example, a CTT
+model could be annotated by associating subtasks to bridge sets and
+then used to generate a sequence of bridge sets, from which
+\migtool\ could generate traces and metrics.  At that point, \migtool\
+could be used as a tool to explore suitability of tasks with respect
+to a behavioral model of a user interface. This problem has been
+addressed in \cite{martinie15} through a mechanism that relates task
+models to source code annotations of the implementation of the user
+interface. In this way it is possible to simulate execution of the
+tasks in parallel with execution of the system and ensure
 compatibility of the system with respect to tasks.
@@
 \migtool\ is implemented partly in Java (model processing)
-and partly in R/iGraph. Development of a web-based UI of \migtool\ is
-underway; it will be freely available for research purposes. Currently
+and partly in R/iGraph~\cite{r14}. 
+%Development of a web-based UI of \migtool\ is
+%underway; it will be freely available for research purposes. 
+Currently
 \migtool\ reads UML models represented as XMI files; it could be
@@
 it is devoid of any concerns dealing with what is presented to users,
-how they could perceive that, understand that, and manipulate that.
+how they could perceive, understand, and manipulate that.
 But, because the approach can be applied  when just a
@@
 iteratively improving a design before building any UI prototype.
-
-%\bibliographystyle{SIGCHI-Reference-Format}
-%\bibliography{gb,database}
-%%% -*-BibTeX-*-
-%%% Do NOT edit. File created by BibTeX with style
-%%% ACM-Reference-Format-Journals [18-Jan-2012].
-
-\begin{thebibliography}{00}
-
-%%% ====================================================================
-%%% NOTE TO THE USER: you can override these defaults by providing
-%%% customized versions of any of these macros before the \bibliography
-%%% command.  Each of them MUST provide its own final punctuation,
-%%% except for \shownote{}, \showDOI{}, and \showURL{}.  The latter two
-%%% do not use final punctuation, in order to avoid confusing it with
-%%% the Web address.
-%%%
-%%% To suppress output of a particular field, define its macro to expand
-%%% to an empty string, or better, \unskip, like this:
-%%%
-%%% \newcommand{\showDOI}[1]{\unskip}   % LaTeX syntax
-%%%
-%%% \def \showDOI #1{\unskip}           % plain TeX syntax
-%%%
-%%% ====================================================================
-
-\ifx \showCODEN    \undefined \def \showCODEN     #1{\unskip}     \fi
-\ifx \showDOI      \undefined \def \showDOI       #1{{\tt DOI:}\penalty0{#1}\ }
-  \fi
-\ifx \showISBNx    \undefined \def \showISBNx     #1{\unskip}     \fi
-\ifx \showISBNxiii \undefined \def \showISBNxiii  #1{\unskip}     \fi
-\ifx \showISSN     \undefined \def \showISSN      #1{\unskip}     \fi
-\ifx \showLCCN     \undefined \def \showLCCN      #1{\unskip}     \fi
-\ifx \shownote     \undefined \def \shownote      #1{#1}          \fi
-\ifx \showarticletitle \undefined \def \showarticletitle #1{#1}   \fi
-\ifx \showURL      \undefined \def \showURL       #1{#1}          \fi
-
-\bibitem{ambler02}
-{S. Ambler}. 2002.
-\newblock {\em Agile Modeling: Effective Practices for eXtreme Programming and
-  the Unified Process}.
-\newblock Wiley.
-\newblock
-\showISBNx{0471202827}
-
-\bibitem{appert2008swingstates}
-{C. Appert} {and} {M. Beaudouin-Lafon}. 2008.
-\newblock \showarticletitle{SwingStates: adding state machines to Java and the
-  Swing toolkit}.
-\newblock {\em Software: Practice and Experience\/} {38}, 11 (2008),
-  1149--1182.
-\newblock
-
-\bibitem{conversy07}
-{E. Barboni}, {S. Conversy}, {D. Navarre}, {and} {P. Palanque}. 2007.
-\newblock \showarticletitle{Model-based engineering of widgets, user
-  applications and servers compliant with ARINC 661 specification}.
-\newblock In {\em Interactive Systems. Design, Specification, and
-  Verification}. Springer, 25--38.
-\newblock
-
-\bibitem{scxml}
-{J. Barnett} {and} {et al}. 2015.
-\newblock {\em State Chart {XML (SCXML)}: State Machine Notation for Control
-  Abstraction}.
-\newblock {T}echnical {R}eport. W3C.
-\newblock
-\newblock
-\shownote{\url{http://www.w3.org/TR/2015/REC-scxml-20150901}.}
-
-\bibitem{bellamy2011cogtool}
-{R. Bellamy}, {B. John}, {and} {S. Kogan}. 2011.
-\newblock \showarticletitle{Deploying CogTool: integrating quantitative
-  usability assessment into real-world software development}. In {\em Software
-  Engineering (ICSE), 2011 33rd International Conference on}. IEEE, 691--700.
-\newblock
-
-\bibitem{gb15:icwe}
-{G. Brajnik} {and} {S. Harper}. 2015.
-\newblock \showarticletitle{Detaching control from data models in model-based
-  generation of user interfaces}. In {\em Proc. of Int. Conf. on Web
-  Engineering}, {F.~Frasincar} (Ed.). IEEE, Rotterdam, Netherlands.
-\newblock
-
-\bibitem{briand2005:test:uml}
-{L.~C Briand}, {Y. Labiche}, {and} {J. Cui}. 2005.
-\newblock \showarticletitle{Automated support for deriving test requirements
-  from UML statecharts}.
-\newblock {\em Software \& Systems Modeling\/} {4}, 4 (2005), 399--423.
-\newblock
-
-\bibitem{buxton07}
-{B. Buxton}. 2007.
-\newblock {\em User Experience: Getting the Design Right and the Right Design}.
-\newblock Morgan Kaufmann.
-\newblock
-
-\bibitem{calvary:iwc03:camaleon}
-{G. Calvary}, {J. Coutaz}, {D. Thevenin}, {Q. Limbourg}, {L. Bouillon}, {and}
-  {J. Vanderdonckt}. 2003.
-\newblock \showarticletitle{A Unifying Reference Framework for Multi-Target
-  User Interfaces}.
-\newblock {\em Interacting with Computers\/} {15}, 3 (2003), 289--308.
-\newblock
-
-\bibitem{ceri:00}
-{S. Ceri}, {P. Fraternali}, {and} {A. Bongio}. 2000.
-\newblock \showarticletitle{{Web Modeling Language (WebML)}: a modeling
-  language for designing web sites}.
-\newblock {\em Computer Networks\/}  {33} (2000), 137--157.
-\newblock
-
-\bibitem{constantine99}
-{L.L. Constantine} {and} {L.A.D. Lockwood}. 1999.
-\newblock {\em Software for use: a practical guide to the models and methods of
-  usage-centered design}.
-\newblock Addison-Wesley.
-\newblock
-
-\bibitem{fayollas2014}
-{C. Fayollas}, {C. Martinie}, {P. Palanque}, {Y. Deleris}, {J-C. Fabre}, {and}
-  {D. Navarre}. 2014.
-\newblock \showarticletitle{An approach for assessing the impact of
-  dependability on usability: application to interactive cockpits}. In {\em
-  Tenth European Dependable Computing Conference (EDCC), 2014}. IEEE, 198--209.
-\newblock
-
-\bibitem{pirolli2007}
-{W.-T. Fu} {and} {P. Pirolli}. 2007.
-\newblock \showarticletitle{{SNIF-ACT}: A cognitive model of user navigation on
-  the World Wide Web}.
-\newblock {\em Human--Computer Interaction\/} {22}, 4 (2007), 355--412.
-\newblock
-
-\bibitem{gomez01}
-{J. G\'{o}mez}, {C. Cachero}, {and} {O. Pastor}. 2001.
-\newblock \showarticletitle{Conceptual Modeling of Device-Independent Web
-  Applications}.
-\newblock {\em IEEE MultiMedia\/} {8}, 2 (2001), 26--39.
-\newblock
-\showISSN{1070-986X}
-\showDOI{%
-\url{http://dx.doi.org/10.1109/93.917969}}
-
-\bibitem{harel09}
-{D. Harel}. 2009.
-\newblock \showarticletitle{Statecharts in the making: a personal account}.
-\newblock {\em CACM\/} {52}, 3 (March 2009), 67--75.
-\newblock
-
-\bibitem{heymann2007formal}
-{M. Heymann} {and} {A. Degani}. 2007.
-\newblock \showarticletitle{Formal analysis and automatic generation of user
-  interfaces: approach, methodology, and an algorithm}.
-\newblock {\em Human Factors: The Journal of the Human Factors and Ergonomics
-  Society\/} {49}, 2 (2007), 311--330.
-\newblock
-
-\bibitem{hilbert2000}
-{D.~M Hilbert} {and} {D.F. Redmiles}. 2000.
-\newblock \showarticletitle{Extracting usability information from user
-  interface events}.
-\newblock {\em ACM Computing Surveys (CSUR)\/} {32}, 4 (2000), 384--421.
-\newblock
-
-\bibitem{hornbaek2006current}
-{K. Hornb{\ae}k}. 2006.
-\newblock \showarticletitle{Current practice in measuring usability: Challenges
-  to usability studies and research}.
-\newblock {\em International Journal of Human-Computer Studies\/} {64}, 2
-  (2006), 79--102.
-\newblock
-
-\bibitem{horrocks99}
-{I. Horrocks}. 1999.
-\newblock {\em Constructing the User Interface with Statecharts}.
-\newblock Addison-Wesley Longman Publishing Co., Inc., Boston, MA, USA.
-\newblock
-\showISBNx{0201342782}
-
-\bibitem{john10}
-{B.E. John}. 2010.
-\newblock \showarticletitle{Reducing the Variability between Novice Modelers:
-  Results of a Tool for Human Performance Modeling Produced through
-  Human-Centered Design}. In {\em Proc. of the 19th Conf. on Behavior
-  Representation in Modeling and Simulation}. Charleston, SC, USA, 95--102.
-\newblock
-
-\bibitem{koch02:uwe}
-{N. Koch} {and} {A. Kraus}. 2002.
-\newblock \showarticletitle{{The expressive power of {UML}-based web
-  engineering}}. In {\em Second International Workshop on Web-Oriented Software
-  Technology (IWWOST02)}, {D.~Schwabe}, {O.~Pastor}, {G.~Rossi}, {and}
-  {L.~Olsina} (Eds.). 105--199.
-\newblock
-
-\bibitem{limbourg04:usixml}
-{Q. Limbourg}, {J. Vanderdonckt}, {B. Michotte}, {L. Bouillon}, {and} {V.
-  López~Jaquero}. 2004.
-\newblock \showarticletitle{UsiXML: a Language Supporting Multi-Path
-  Development of User Interfaces}. In {\em Proc. of 9th IFIP Working Conference
-  on Engineering for Human-Computer Interaction jointly with 11th Int. Workshop
-  on Design, Specification, and Verification of Interactive Systems,
-  EHCI-DSVIS'2004} {\em (Lecture Notes in Computer Science)}, Vol. 3425.
-  Springer-Verlag, Hamburg, Germany, 200--220.
-\newblock
-
-\bibitem{martinie15}
-{C. Martinie}, {D. Navarre}, {P. Palanque}, {and} {C. Fayollas}. 2015.
-\newblock \showarticletitle{A generic tool--supported framework for coupling
-  task models and interactive applications}. In {\em Proc. of Engineering
-  Interactive Computing Systems, {EICS} '15}. ACM, ACM Press, Duisburg,
-  Germany, 244--253.
-\newblock
-
-\bibitem{mccurdy06}
-{M. McCurdy}, {C. Connors}, {G. Pyrzak}, {B. Kanefsky}, {and} {A. Vera}. 2006.
-\newblock \showarticletitle{Breaking the fidelity barrier: an examination of
-  our current characterization of prototypes and an example of a mixed-fidelity
-  success}. In {\em CHI 2006}. ACM, ACM Press, New York, NY, 1233--1242.
-\newblock
-
-\bibitem{melia08}
-{S. Meli\'a}, {J. G\'omez}, {S. P\'erez}, {and} {O. D\'iaz}. 2008.
-\newblock \showarticletitle{A Model-Driven Development for GWT- Based Rich
-  Internet Applications with OOH4RIA}. In {\em Proc. 8th Int'l Conf. Web Eng.
-  (ICWE 2008)}. IEEE CS Press, 13--23.
-\newblock
-
-\bibitem{mori02}
-{G. Mori}, {F. Patern\`o}, {and} {C. Santoro}. 2002.
-\newblock \showarticletitle{{CTTE: Support for Developing and Analysing Task
-  Models for Interactive System Design}}.
-\newblock {\em IEEE Transactions on Software Engineering\/} {28}, 8 (August
-  2002), 797--813.
-\newblock
-
-\bibitem{ifml13}
-{{OMG}}. 2013.
-\newblock {\em Interaction Flow Modeling Language ({IFML}), FTF -- Beta 1\/}
-  (omg document number: ptc/2013-03-08 ed.).
-\newblock {T}echnical {R}eport. OMG.
-\newblock
-\showURL{%
-\url{http://www.omg.org/spec/IFML/1.0}}
-
-\bibitem{otter2000lostness}
-{M. Otter} {and} {H. Johnson}. 2000.
-\newblock \showarticletitle{Lost in hyperspace: metrics and mental models}.
-\newblock {\em Interacting with Computers\/} {13}, 1 (2000), 1--40.
-\newblock
-
-\bibitem{page1999pagerank}
-{L. Page}, {S. Brin}, {R. Motwani}, {and} {T. Winograd}. 1999.
-\newblock \showarticletitle{The PageRank citation ranking: bringing order to
-  the Web.}
-\newblock  (1999).
-\newblock
-
-\bibitem{parnas69}
-{D.L. Parnas}. 1969.
-\newblock \showarticletitle{On the use of transition diagrams in the design of
-  a user interface for an interactive computer system}. In {\em ACM '69 Proc.
-  of the 1969 24th National Conference}. ACM.
-\newblock
-
-\bibitem{silva03:umli}
-{P. Pinhero~da Silva} {and} {N.W. Paton}. 2003.
-\newblock \showarticletitle{User Interface Modeling in {UMLi}}.
-\newblock {\em IEEE Software\/} (2003), 62--69.
-\newblock
-
-\bibitem{preece02}
-{J. Preece}, {Y. Rogers}, {and} {H. Sharp}. 2002.
-\newblock {\em Interaction Design}.
-\newblock John Wiley and Sons.
-\newblock
-
-\bibitem{rubin08}
-{J. Rubin} {and} {D. Chisnell}. 2008.
-\newblock {\em Handbook of Usability Testing\/} (second ed.).
-\newblock Wiley.
-\newblock
-
-\bibitem{smith1996lostness}
-{P.A. Smith}. 1996.
-\newblock \showarticletitle{Towards a practical measure of hypertext
-  usability}.
-\newblock {\em Interacting with Computers\/} {8}, 4 (1996), 365--381.
-\newblock
-
-\bibitem{swearngin13}
-{A. Swearngin}, {M.B. Cohen}, {B.E. John}, {and} {K.E. Bellamy}. 2013.
-\newblock \showarticletitle{Human Performance Regression Testing}. In {\em
-  Proc. of the 2013 International Conference on Software Engineering} {\em
-  (ICSE '13)}. IEEE Press, Piscataway, NJ, USA, 152--161.
-\newblock
-\showURL{%
-\url{http://dl.acm.org/citation.cfm?id=2486788.2486809}}
-
-\bibitem{thimbleby07}
-{H. Thimbleby}. 2007.
-\newblock {\em Press on: principles of interaction programming}.
-\newblock The MIT Press.
-\newblock
-
-\bibitem{thimbleby2001markov}
-{H. Thimbleby}, {P. Cairns}, {and} {M. Jones}. 2001.
-\newblock \showarticletitle{Usability analysis with Markov models}.
-\newblock {\em ACM Transactions on Computer-Human Interaction (TOCHI)\/} {8}, 2
-  (2001), 99--132.
-\newblock
-
-\bibitem{thimbleby09:sna:eics09}
-{H. Thimbleby} {and} {P. Oladimeji}. 2009.
-\newblock \showarticletitle{Social Network Analysis and Interactive Device
-  Design Analysis}. In {\em Proc. of Engineering Interactive Computing Systems
-  2009}. ACM Press, 91--100.
-\newblock
-
-\bibitem{winckler03}
-{M. Winckler} {and} {P.A. Palanque}. 2003.
-\newblock \showarticletitle{StateWebCharts: A Formal Description Technique
-  Dedicated to Navigation Modelling of Web Applications.}. In {\em Design,
-  Specification, and Verification: 10th International Workshop, DSV-IS'03}.
-  Funchal,Madeira Island, Portugal, 61--76.
-\newblock
-
-\end{thebibliography}
+\newpage
+\bibliographystyle{SIGCHI-Reference-Format}
+\bibliography{gb,database}
+
+\balance{}
 

### 2016-04-11 paper.tex
details

@@
 \newpage
-\bibliographystyle{SIGCHI-Reference-Format}
-\bibliography{gb,database}
+%\bibliographystyle{SIGCHI-Reference-Format}
+%\bibliography{gb,database}
+%%% -*-BibTeX-*-
+%%% Do NOT edit. File created by BibTeX with style
+%%% ACM-Reference-Format-Journals [18-Jan-2012].
+
+\begin{thebibliography}{00}
+
+%%% ====================================================================
+%%% NOTE TO THE USER: you can override these defaults by providing
+%%% customized versions of any of these macros before the \bibliography
+%%% command.  Each of them MUST provide its own final punctuation,
+%%% except for \shownote{}, \showDOI{}, and \showURL{}.  The latter two
+%%% do not use final punctuation, in order to avoid confusing it with
+%%% the Web address.
+%%%
+%%% To suppress output of a particular field, define its macro to expand
+%%% to an empty string, or better, \unskip, like this:
+%%%
+%%% \newcommand{\showDOI}[1]{\unskip}   % LaTeX syntax
+%%%
+%%% \def \showDOI #1{\unskip}           % plain TeX syntax
+%%%
+%%% ====================================================================
+
+\ifx \showCODEN    \undefined \def \showCODEN     #1{\unskip}     \fi
+\ifx \showDOI      \undefined \def \showDOI       #1{{\tt DOI:}\penalty0{#1}\ }
+  \fi
+\ifx \showISBNx    \undefined \def \showISBNx     #1{\unskip}     \fi
+\ifx \showISBNxiii \undefined \def \showISBNxiii  #1{\unskip}     \fi
+\ifx \showISSN     \undefined \def \showISSN      #1{\unskip}     \fi
+\ifx \showLCCN     \undefined \def \showLCCN      #1{\unskip}     \fi
+\ifx \shownote     \undefined \def \shownote      #1{#1}          \fi
+\ifx \showarticletitle \undefined \def \showarticletitle #1{#1}   \fi
+\ifx \showURL      \undefined \def \showURL       #1{#1}          \fi
+
+\bibitem{ambler02}
+{S. Ambler}. 2002.
+\newblock {\em Agile Modeling: Effective Practices for eXtreme Programming and
+  the Unified Process}.
+\newblock Wiley.
+\newblock
+\showISBNx{0471202827}
+
+\bibitem{appert2008swingstates}
+{C. Appert} {and} {M. Beaudouin-Lafon}. 2008.
+\newblock \showarticletitle{SwingStates: adding state machines to Java and the
+  Swing toolkit}.
+\newblock {\em Software: Practice and Experience\/} {38}, 11 (2008),
+  1149--1182.
+\newblock
+
+\bibitem{conversy07}
+{E. Barboni}, {S. Conversy}, {D. Navarre}, {and} {P. Palanque}. 2007.
+\newblock \showarticletitle{Model-based engineering of widgets, user
+  applications and servers compliant with ARINC 661 specification}.
+\newblock In {\em Interactive Systems. Design, Specification, and
+  Verification}. Springer, 25--38.
+\newblock
+
+\bibitem{scxml}
+{J. Barnett} {and} {et al}. 2015.
+\newblock {\em State Chart {XML (SCXML)}: State Machine Notation for Control
+  Abstraction}.
+\newblock {T}echnical {R}eport. W3C.
+\newblock
+\newblock
+\shownote{\url{http://www.w3.org/TR/2015/REC-scxml-20150901}.}
+
+\bibitem{bellamy2011cogtool}
+{R. Bellamy}, {B. John}, {and} {S. Kogan}. 2011.
+\newblock \showarticletitle{Deploying CogTool: integrating quantitative
+  usability assessment into real-world software development}. In {\em Software
+  Engineering (ICSE), 2011 33rd International Conference on}. IEEE, 691--700.
+\newblock
+
+\bibitem{gb15:icwe}
+{G. Brajnik} {and} {S. Harper}. 2015.
+\newblock \showarticletitle{Detaching control from data models in model-based
+  generation of user interfaces}. In {\em Proc. of Int. Conf. on Web
+  Engineering}, {F.~Frasincar} (Ed.). IEEE, Rotterdam, Netherlands.
+\newblock
+
+\bibitem{briand2005:test:uml}
+{L.~C Briand}, {Y. Labiche}, {and} {J. Cui}. 2005.
+\newblock \showarticletitle{Automated support for deriving test requirements
+  from UML statecharts}.
+\newblock {\em Software \& Systems Modeling\/} {4}, 4 (2005), 399--423.
+\newblock
+
+\bibitem{buxton07}
+{B. Buxton}. 2007.
+\newblock {\em User Experience: Getting the Design Right and the Right Design}.
+\newblock Morgan Kaufmann.
+\newblock
+
+\bibitem{calvary:iwc03:camaleon}
+{G. Calvary}, {J. Coutaz}, {D. Thevenin}, {Q. Limbourg}, {L. Bouillon}, {and}
+  {J. Vanderdonckt}. 2003.
+\newblock \showarticletitle{A Unifying Reference Framework for Multi-Target
+  User Interfaces}.
+\newblock {\em Interacting with Computers\/} {15}, 3 (2003), 289--308.
+\newblock
+
+\bibitem{ceri:00}
+{S. Ceri}, {P. Fraternali}, {and} {A. Bongio}. 2000.
+\newblock \showarticletitle{{Web Modeling Language (WebML)}: a modeling
+  language for designing web sites}.
+\newblock {\em Computer Networks\/}  {33} (2000), 137--157.
+\newblock
+
+\bibitem{constantine99}
+{L.L. Constantine} {and} {L.A.D. Lockwood}. 1999.
+\newblock {\em Software for use: a practical guide to the models and methods of
+  usage-centered design}.
+\newblock Addison-Wesley.
+\newblock
+
+\bibitem{fayollas2014}
+{C. Fayollas}, {C. Martinie}, {P. Palanque}, {Y. Deleris}, {J-C. Fabre}, {and}
+  {D. Navarre}. 2014.
+\newblock \showarticletitle{An approach for assessing the impact of
+  dependability on usability: application to interactive cockpits}. In {\em
+  Tenth European Dependable Computing Conference (EDCC), 2014}. IEEE, 198--209.
+\newblock
+
+\bibitem{pirolli2007}
+{W.-T. Fu} {and} {P. Pirolli}. 2007.
+\newblock \showarticletitle{{SNIF-ACT}: A cognitive model of user navigation on
+  the World Wide Web}.
+\newblock {\em Human--Computer Interaction\/} {22}, 4 (2007), 355--412.
+\newblock
+
+\bibitem{gomez01}
+{J. G\'{o}mez}, {C. Cachero}, {and} {O. Pastor}. 2001.
+\newblock \showarticletitle{Conceptual Modeling of Device-Independent Web
+  Applications}.
+\newblock {\em IEEE MultiMedia\/} {8}, 2 (2001), 26--39.
+\newblock
+\showISSN{1070-986X}
+\showDOI{%
+\url{http://dx.doi.org/10.1109/93.917969}}
+
+\bibitem{heymann2007formal}
+{M. Heymann} {and} {A. Degani}. 2007.
+\newblock \showarticletitle{Formal analysis and automatic generation of user
+  interfaces: approach, methodology, and an algorithm}.
+\newblock {\em Human Factors: The Journal of the Human Factors and Ergonomics
+  Society\/} {49}, 2 (2007), 311--330.
+\newblock
+
+\bibitem{hilbert2000}
+{D.~M Hilbert} {and} {D.F. Redmiles}. 2000.
+\newblock \showarticletitle{Extracting usability information from user
+  interface events}.
+\newblock {\em ACM Computing Surveys (CSUR)\/} {32}, 4 (2000), 384--421.
+\newblock
+
+\bibitem{hornbaek2006current}
+{K. Hornb{\ae}k}. 2006.
+\newblock \showarticletitle{Current practice in measuring usability: Challenges
+  to usability studies and research}.
+\newblock {\em International Journal of Human-Computer Studies\/} {64}, 2
+  (2006), 79--102.
+\newblock
+
+\bibitem{horrocks99}
+{I. Horrocks}. 1999.
+\newblock {\em Constructing the User Interface with Statecharts}.
+\newblock Addison-Wesley Longman Publishing Co., Inc., Boston, MA, USA.
+\newblock
+\showISBNx{0201342782}
+
+\bibitem{john10}
+{B.E. John}. 2010.
+\newblock \showarticletitle{Reducing the Variability between Novice Modelers:
+  Results of a Tool for Human Performance Modeling Produced through
+  Human-Centered Design}. In {\em Proc. of the 19th Conf. on Behavior
+  Representation in Modeling and Simulation}. Charleston, SC, USA, 95--102.
+\newblock
+
+\bibitem{koch02:uwe}
+{N. Koch} {and} {A. Kraus}. 2002.
+\newblock \showarticletitle{{The expressive power of {UML}-based web
+  engineering}}. In {\em Second International Workshop on Web-Oriented Software
+  Technology (IWWOST02)}, {D.~Schwabe}, {O.~Pastor}, {G.~Rossi}, {and}
+  {L.~Olsina} (Eds.). 105--199.
+\newblock
+
+\bibitem{limbourg04:usixml}
+{Q. Limbourg}, {J. Vanderdonckt}, {B. Michotte}, {L. Bouillon}, {and} {V.
+  López~Jaquero}. 2004.
+\newblock \showarticletitle{UsiXML: a Language Supporting Multi-Path
+  Development of User Interfaces}. In {\em Proc. of 9th IFIP Working Conference
+  on Engineering for Human-Computer Interaction jointly with 11th Int. Workshop
+  on Design, Specification, and Verification of Interactive Systems,
+  EHCI-DSVIS'2004} {\em (Lecture Notes in Computer Science)}, Vol. 3425.
+  Springer-Verlag, Hamburg, Germany, 200--220.
+\newblock
+
+\bibitem{martinie15}
+{C. Martinie}, {D. Navarre}, {P. Palanque}, {and} {C. Fayollas}. 2015.
+\newblock \showarticletitle{A generic tool--supported framework for coupling
+  task models and interactive applications}. In {\em Proc. of Engineering
+  Interactive Computing Systems, {EICS} '15}. ACM, ACM Press, Duisburg,
+  Germany, 244--253.
+\newblock
+
+\bibitem{mccurdy06}
+{M. McCurdy}, {C. Connors}, {G. Pyrzak}, {B. Kanefsky}, {and} {A. Vera}. 2006.
+\newblock \showarticletitle{Breaking the fidelity barrier: an examination of
+  our current characterization of prototypes and an example of a mixed-fidelity
+  success}. In {\em CHI 2006}. ACM, ACM Press, New York, NY, 1233--1242.
+\newblock
+
+\bibitem{melia08}
+{S. Meli\'a}, {J. G\'omez}, {S. P\'erez}, {and} {O. D\'iaz}. 2008.
+\newblock \showarticletitle{A Model-Driven Development for GWT- Based Rich
+  Internet Applications with OOH4RIA}. In {\em Proc. 8th Int'l Conf. Web Eng.
+  (ICWE 2008)}. IEEE CS Press, 13--23.
+\newblock
+
+\bibitem{mori02}
+{G. Mori}, {F. Patern\`o}, {and} {C. Santoro}. 2002.
+\newblock \showarticletitle{{CTTE: Support for Developing and Analysing Task
+  Models for Interactive System Design}}.
+\newblock {\em IEEE Transactions on Software Engineering\/} {28}, 8 (August
+  2002), 797--813.
+\newblock
+
+\bibitem{ifml13}
+{{OMG}}. 2013.
+\newblock {\em Interaction Flow Modeling Language ({IFML}), FTF -- Beta 1\/}
+  (omg document number: ptc/2013-03-08 ed.).
+\newblock {T}echnical {R}eport. OMG.
+\newblock
+\showURL{%
+\url{http://www.omg.org/spec/IFML/1.0}}
+
+\bibitem{uml25}
+{{Object Management Group} OMG}. 2015.
+\newblock {OMG Unified Modeling Language ({OMG UML})} Version 2.5.
+\newblock \url{http://www.omg.org/spec/UML/2.5}.   (March 2015).
+\newblock
+\showURL{%
+\url{http://www.omg.org/spec/UML/2.5/PDF}}
+
+\bibitem{otter2000lostness}
+{M. Otter} {and} {H. Johnson}. 2000.
+\newblock \showarticletitle{Lost in hyperspace: metrics and mental models}.
+\newblock {\em Interacting with Computers\/} {13}, 1 (2000), 1--40.
+\newblock
+
+\bibitem{page1999pagerank}
+{L. Page}, {S. Brin}, {R. Motwani}, {and} {T. Winograd}. 1999.
+\newblock \showarticletitle{The PageRank citation ranking: bringing order to
+  the Web.}
+\newblock  (1999).
+\newblock
+
+\bibitem{parnas69}
+{D.L. Parnas}. 1969.
+\newblock \showarticletitle{On the use of transition diagrams in the design of
+  a user interface for an interactive computer system}. In {\em ACM '69 Proc.
+  of the 1969 24th National Conference}. ACM.
+\newblock
+
+\bibitem{silva03:umli}
+{P. Pinhero~da Silva} {and} {N.W. Paton}. 2003.
+\newblock \showarticletitle{User Interface Modeling in {UMLi}}.
+\newblock {\em IEEE Software\/} (2003), 62--69.
+\newblock
+
+\bibitem{preece02}
+{J. Preece}, {Y. Rogers}, {and} {H. Sharp}. 2002.
+\newblock {\em Interaction Design}.
+\newblock John Wiley and Sons.
+\newblock
+
+\bibitem{r14}
+{{R Core Team}}. 2014.
+\newblock {\em R: A Language and Environment for Statistical Computing}.
+\newblock R Foundation for Statistical Computing, Vienna, Austria.
+\newblock
+\showURL{%
+\url{http://www.R-project.org/}}
+
+\bibitem{rubin08}
+{J. Rubin} {and} {D. Chisnell}. 2008.
+\newblock {\em Handbook of Usability Testing\/} (second ed.).
+\newblock Wiley.
+\newblock
+
+\bibitem{silva14}
+{Jos{\'e}-Luis Silva}, {Camille Fayollas}, {Arnaud Hamon}, {C{\'e}lia
+  Martiinie}, {Eric Barboni}, {and} {others}. 2014.
+\newblock \showarticletitle{Analysis of WIMP and Post WIMP Interactive Systems
+  based on Formal Specification}.
+\newblock {\em Electronic Communications of the EASST\/}  {69} (2014).
+\newblock
+
+\bibitem{smith1996lostness}
+{P.A. Smith}. 1996.
+\newblock \showarticletitle{Towards a practical measure of hypertext
+  usability}.
+\newblock {\em Interacting with Computers\/} {8}, 4 (1996), 365--381.
+\newblock
+
+\bibitem{swearngin13}
+{A. Swearngin}, {M.B. Cohen}, {B.E. John}, {and} {K.E. Bellamy}. 2013.
+\newblock \showarticletitle{Human Performance Regression Testing}. In {\em
+  Proc. of the 2013 International Conference on Software Engineering} {\em
+  (ICSE '13)}. IEEE Press, Piscataway, NJ, USA, 152--161.
+\newblock
+\showURL{%
+\url{http://dl.acm.org/citation.cfm?id=2486788.2486809}}
+
+\bibitem{thimbleby07}
+{H. Thimbleby}. 2007.
+\newblock {\em Press on: principles of interaction programming}.
+\newblock The MIT Press.
+\newblock
+
+\bibitem{thimbleby2001markov}
+{H. Thimbleby}, {P. Cairns}, {and} {M. Jones}. 2001.
+\newblock \showarticletitle{Usability analysis with Markov models}.
+\newblock {\em ACM Transactions on Computer-Human Interaction (TOCHI)\/} {8}, 2
+  (2001), 99--132.
+\newblock
+
+\bibitem{thimbleby09:sna:eics09}
+{H. Thimbleby} {and} {P. Oladimeji}. 2009.
+\newblock \showarticletitle{Social Network Analysis and Interactive Device
+  Design Analysis}. In {\em Proc. of Engineering Interactive Computing Systems
+  2009}. ACM Press, 91--100.
+\newblock
+
+\bibitem{winckler03}
+{M. Winckler} {and} {P.A. Palanque}. 2003.
+\newblock \showarticletitle{StateWebCharts: A Formal Description Technique
+  Dedicated to Navigation Modelling of Web Applications.}. In {\em Design,
+  Specification, and Verification: 10th International Workshop, DSV-IS'03}.
+  Funchal,Madeira Island, Portugal, 61--76.
+\newblock
+
+\end{thebibliography}
 

### 2016-09-06 paper.tex
details
added copyright piece

@@
-\documentclass{sigchi}
+\documentclass{sig-alternate-05-2015}
 
@@
 \newcommand{\eg}{\emph{e.g.}}
-\newcommand{\migtool}{MIGtool}
+\newcommand{\migtool}{MIG\-tool}
 \usepackage{soul}
@@
 % }
+\CopyrightYear{2016}
+\setcopyright{acmlicensed}
+\conferenceinfo{EICS'16,}{June 21 - 24, 2016, Brussels, Belgium}
+\isbn{978-1-4503-4322-0/16/06}\acmPrice{\$15.00}
+\doi{http://dx.doi.org/10.1145/2933242.2933246}
 
@@
 
-\keywords{Experimental; Evaluation; Statecharts: UML; Metrics; Testing.}
+\keywords{User Interfaces; Evaluation; Statecharts: UML; Metrics; Usability.}
 
@@
 
+\newpage
 \section{Introduction}
-
-% \mytodo{ Subcommittee: Technology, Systems, and Engineering
-
-% People we should pay particular attention to:
-% Caroline Appert
-% Conversy, Stéphane
-% Nebeling, Michael
-
-% Keyword we need to pay attention to:
-% This subcommittee will focus on technology, systems and engineering
-% contributions that enable, improve, or advance interaction. This will
-% include software and hardware technologies and systems that enable and
-% demonstrate novel interactive capabilities, as well as languages,
-% methods and tools for \hl{CONSTRUCTION AND ENGINEERING} of interactive
-% systems. Engineering contributions should clearly demonstrate how they
-% address interactive systems concerns such as, for example,
-% scalability, reliability, interoperability, testing, and
-% performance. Systems and technology contributions will be judged by
-% their technical innovation and/or ability to connect, SIMPLIFY or
-% enrich interactions, for example in intelligent interfaces and
-% mobile/ubiquitous computing.
-% }
-% % 
-%\end{quote}
 
@@
 \item understand how recoverable the UI is from those errors; and,
-\item perform objective quantitative comparison of different designs
+\item perform objective quantitative comparisons of different designs
   to support construction and engineering of interactive systems.
@@
 %
-Adopting Usage Centered Development practices is effective, as
-is following established design principles~\cite{constantine99}. In
-particular, one of the most effective technique is early prototyping
-to explore part of the five-dimensional fidelity prototyping
-space~\cite{mccurdy06,buxton07}. This is particularly effective when
-paired with usability investigations based on user testing or
-heuristic evaluations~\cite{preece02,rubin08}.
+Adoption of usage-centered development practices and following
+established design principles~\cite{constantine99} is a sound way to
+tackle the problem.  In particular very effective techniques are
+prototyping, to explore part of the five-dimensional fidelity
+space~\cite{mccurdy06,buxton07}, paired with usability investigations,
+based on user testing or heuristic
+evaluations~\cite{preece02,rubin08}.
 
@@
 answer. In fact, they require developing  prototypes, inspecting
-them, manual tracking of which screens and widgets are used at which
+them, manually tracking which screens and widgets are used at which
 stage, followed by a systematic manual analysis.
@@
 This should not be the case, however, because answers to these
-questions could be automatically computed. In this way a system could
+questions could be automatically computed. Such a system could
 provide important insights to a designer, and support assessments of
@@
 
-Our approach is based on UML statechart models of UIs which are
+Our approach is based on UML state machine models of UIs which are
 automatically processed to produce \emph{interaction graphs}. These
@@
 are compatible with the specific scenarios being considered. Traces
-are fed to graph-theoretic computations which produce a
-dashboard with different results. Except for development of models and
-specification of the desired scenarios, which have to be done
-manually, the other steps are totally automatic; models of a design
-(such as the ones shown below) can be developed in a matter of a
-couple of hours.
+are processed to produce a dashboard with different results. Except
+for development of models and specification of the desired scenarios,
+which have to be done manually, the other steps are totally automatic;
+models of a design (such as the ones shown below) can be developed in
+a matter of a couple of hours.
 
@@
 that the error-proneness of the UI has reduced; in the best case,
-Gmail features also the shortest paths, requiring 13\% fewer steps;
-however, the probability that a user hits an optimal path with Gmail
-is 10 times smaller than the best of the other applications, and the
-probability that a random walker hits a state that is not due to an
-error is 10\% smaller than the best of the other applications.
+Gmail features also the shortest paths, requiring 13\% fewer steps
+than other applications; however, the probability that a user hits an
+optimal path with Gmail is 10 times smaller than the best of the other
+applications, and the probability that a random walker hits a state
+that is not due to an error is 10\% smaller than the best of the other
+applications.
 %
@@
 
-Lostness~\cite{smith1996lostness,otter2000lostness} is a
-metric for measuring the degree to which users become lost in the
-information space, and therefore considers the notion of
-\emph{deviation}. Defined specifically for hypertext systems, lostness
-is a user performance measure  which is a function of the number of
-visited nodes, the number of different nodes that were visited nodes,
-and the number of required nodes. This measure of efficiency is
-usually applied to traces of actual users, and is argued to be
-suitable for hypertext systems because the predominant task is
-browsing information, rather than trying to achieve specific goals. It
-is suitable therefore when some of the following three assumptions can
-be relaxed: that there is a task to complete, that there is a correct
-way to carry it out, that the purpose of the system is to support
-users to carry out their tasks.
+On the other hand the literature on metrics is also large.
+Lostness~\cite{smith1996lostness,otter2000lostness} is a metric for
+measuring the degree to which users become lost in the information
+space, and it considers the notion of \emph{deviation}. Defined
+specifically for hypertext systems, lostness is a user performance
+measure which is a function of the number of visited nodes, the number
+of different nodes that were visited, and the number of required
+nodes. This measure of efficiency is usually applied to traces of
+actual users, and is argued to be suitable for hypertext systems
+because the predominant task is browsing information, rather than
+trying to achieve specific goals. It is suitable therefore when some
+of the following three assumptions can be relaxed: that there is a
+task to complete, that there is a correct way to carry it out, that
+the purpose of the system is to support users in carrying out their
+tasks.
 
@@
 advantage of a model of the system and instead relies only on models
-of the tasks. The conseguence is that analysing alternative ways to
+of the tasks. The consequence is that analyzing alternative ways to
 achieve the same goal requires developing different models; in
@@
 available. Finally, using \emph{only} hierarchical tasks as a modeling
-language limits the richness of the scenarios being analysed, because
+language limits the richness of the scenarios being analyzed, because
 it is difficult to take advantage of contexts and rich control flows
@@
   patterns}, number of keystrokes, mouse clicks, and visited objects
-as possible metrics. Notice that in general the time needed to carry
-out individual interaction steps depends on many aspects, such as the
+as possible metrics. Of course also the time needed to carry
+out individual interaction steps counts and it depends on many aspects, such as the
 time to decide what goal to achieve, to understand how to carry it
@@
 gets an estimation of how long a skilled user would spend on each
-step. CogTool takes care of adding extra ``mental'' steps, such as
-``think'' steps, before certain patterns of provided steps, according
-to the cognitive theory underlying ACT-R. As a result, users of
-CogTool obtain the breakdown of the times required by each of the
-steps. Our method is less precise: it does not provide expected
-completion times. However, with our method, a designer can analyze a
-large part of the UI, get information about possible problems in some
-areas, and only then devolve more resources in building storyboards
-and in making assumptions regarding widgets so that specific execution
-paths previously identified can be analyzed with CogTool. In a sense,
-the output obtained with our method could be used to make informed
-decisions as to what to analyze next with CogTool. Approaches like
-\cite{swearngin13} start from the assumption that the user interface
-exists already, and that it can \emph{ripped} to reverse engineer a
-behavior model. These assumptions are not needed to use \migtool.
-
-Another approach for analysing user actions that has a strong
+step. CogTool takes care of adding extra ``mental'' steps before
+certain patterns of provided steps, according to the cognitive theory
+underlying ACT-R. As a result, users of CogTool obtain the breakdown
+of the times required by the task. Our method is less precise: it does
+not provide expected completion times. However, with our method, a
+designer can analyze a large part of the UI, get information about
+possible problems in some areas, and only then devolve more resources
+in building storyboards and in making assumptions regarding widgets so
+that specific execution paths previously identified can be analyzed
+with CogTool. In a sense, the output obtained with our method could be
+used to make informed decisions as to what to analyze next with
+CogTool. Other approaches, like \cite{swearngin13}, start from the assumption
+that the user interface exists already, and that it can \emph{ripped}
+to reverse engineer a behavior model. These assumptions are not needed
+when using \migtool.
+
+Another approach for analyzing user actions that has a strong
 cognitive background is SNIF-ACT \cite{pirolli2007}. However it has
 been conceived and applied to information websites, where the notion
-of \emph{information scent} can bear upon interaction. We opted for
+of \emph{information scent}  bears upon interaction. We opted for
 not applying such concepts to the interaction with a web application,
@@
 associated to a probability distribution, were used
-in~\cite{thimbleby2001markov}, as a means to perform usability
-analysis as early as possible, even before building prototypes of the
+in~\cite{thimbleby2001markov} as a means to perform usability analysis
+as early as possible, even before building prototypes of the
 UI. Vertices represent states of the UI and edges correspond to user
-actions (such as pressing a push-down button).  Probabilities can be
-used as a model of user knowledge: equiprobable actions correspond to
-a knowledge-free user, whereas when some actions have a very low
+actions (such as pressing a button).  Probabilities can be used as a
+model of user knowledge: equiprobable actions correspond to a
+knowledge-free user, whereas when some actions have a very low
 probability it means that for that user the action is unlikely to be
@@
 cover several devices, ranging from a simple torch (with 4 states), a
-microwave cooker (6 states), a mobile phone (152 states). Notice that those are
-all push-down devices with a fixed set of buttons. This is obviously
-not the case for UIs of information systems, where buttons may change
-screen by screen. This makes it more difficult to specify the
+microwave cooker (6 states), a mobile phone (152 states). Notice that
+those are all push-down devices with a fixed set of buttons. This is
+obviously not the case for UIs of information systems, where buttons
+may change screen by screen and there could also be an arbitrary
+number of links. This makes it more difficult to specify the
 transition matrix.
@@
 In our work we automatically generate graphs from statechart models,
-and on some of them we computed these metrics. We were not able to draw
-sensible conclusions from the values we obtained (for example from the
-models presented below). One possible explanation rests on the
-different types of models: in our case they reflect the variety and
-flexibility with which ``buttons'' can be used in modern web
-applications. For infusion pumps the UI is more constrained in how a
-task can be carried on, and this difference might reflect on the
-usefulness of those metrics.
+and on some of them we computed these metrics. In case studies
+presented below we were not able to draw sensible conclusions from the
+values we obtained (for example from the models presented below). One
+possible explanation rests on the different types of models: in our
+case they reflect the variety and flexibility with which ``buttons''
+can be used in modern web applications. For infusion pumps the UI is
+more constrained in how a task can be carried on, and this difference
+might reflect on the usefulness of those metrics.
 
@@
 
-\section{Generation of interaction traces}
+\section{Generation of traces}
 \label{sec:traces}
@@
 transition guards are not given semantics and are simply treated as
-part of the event of a transition. \migtool\ produces an XML
+part of the event label of a transition. \migtool\ produces an XML
 representation (graphML) of the resulting FSA, the \emph{interaction
@@
   ``shortest path''.}, potentially with cycles and loops, with edges
-that are labeled with the name of the corresponding action. In the
-following we will be using as synonymous the terms state/vertex, and
-action/edge.
+labeled with the name of the corresponding action. In the following we
+will be using as synonymous the terms state/vertex, action/edge and
+trace/path.
 
@@
 scenario unviable). In general a specification of a scenario includes
-an initial state and a non-empty sequence bridge sets. For example, to
+an initial state and a non-empty sequence of bridge sets. For example, to
 specify a scenario for replying to an email message, assuming an
 initial state that corresponds to a user interface that displays all
-messaes in the inbox, one could select all the edges associated to the
+messages in the inbox, one could select all the edges associated to the
 action \texttt{reply} (bridge set 1), followed by edges labeled with
@@
 Quite expressive languages can be conceived for grounding scenarios
-(\eg, regular expressions on sequences of action names). But such
+(\eg, regular expressions on sequences of action labels). But such
 expressivity bonus needs to be balanced with computational
@@
 To cope with this we implemented a trace searching algorithm that
-processes each of the stages sequentially, starting from the
-initial state. Given a bridge set $B_i$, the algorithm does a
+processes each of the stages sequentially, starting from the initial
+state. Given a stage $i$ and a bridge set $B_i$, the algorithm does a
 breadth-first search of all the geodesic paths that connect
@@
 found for stage $i$ are joined with those of ${i+1}$. These global
-paths, connecting the initial state to reachable bridges of the
-last bridge set of the scenario, are called \emph{execution traces}.
+paths, connecting the initial state to reachable bridges of the last
+bridge set of the scenario, are called \emph{execution traces}.
 
@@
 actions, not all the necessary ones. For example, if the model
-prescribes that in order to \texttt{view(message)} while reading
-another one, one has to \texttt{goBack} to the inbox first, a scenario
-specifying two consecutive \texttt{view(message)} would lead to traces
-that include also the \texttt{goBack} action, even if that step is not
-explicitly specified in the scenario. It is the task of \migtool\ to
-unfold paths in the graph and search all the geodesic paths that
-connect the desired user actions.
+prescribes that in order to perform action \texttt{view(message)}
+while reading another message one has to \texttt{goBack} to the inbox
+first, a scenario specifying two consecutive \texttt{view(message)}
+would lead to traces that include also the \texttt{goBack} action,
+even if that step is not explicitly specified in the scenario. It is
+the task of \migtool\ to unfold paths in the graph and search all the
+geodesic paths that connect the desired user actions.
 
@@
 
-\migtool\ is used to process a model and a scenario
-and to generate execution traces of detour order = 0,
-..., H. 
-For each stage, up to a maximum detour order $H$, the search algorithm
-creates traces with detour order $k+1$ by collecting the set $D_k$ of states
-with order $0$ up to $k$, and by identifying the neighbors $N_k$
-of $D_k$ ($N_k$ is the set of states  not included in $D_k$ that are
-connected through an edge to some state in $D_k$). Edges connecting
-$D_k$ with $N_k$ represent deviations that users might follow, and
-geodesic paths from $N_k$ to $D_k$ constitute recovery paths that
-users might follow to complete the scenario from states in
-$N_k$. These deviations and recovery paths are joined and constitute
-the traces of order $k+1$.  It could happen that
-for some state in $N_k$ there is no path leading to any vertex in
-$D_k$: in such a case the deviation is a dead end that prevents
-the user to complete the scenario.
+\migtool\ is used to process a model and a scenario and to generate
+execution traces of detour order = 0, ..., H.  For each stage, up to a
+maximum detour order $H$, the search algorithm creates traces with
+detour order $k+1$ by collecting the set $D_k$ of states with order
+$0$ up to $k$, and by identifying the neighbors $N_k$ of $D_k$ ($N_k$
+is the set of states not included in $D_k$ that can be reached through
+an edge from some state in $D_k$). Edges connecting $D_k$ with $N_k$
+represent deviations that users might follow, and geodesic paths from
+$N_k$ to $D_k$ constitute recovery paths that users might follow to
+complete the scenario from states in $N_k$. These deviations and
+recovery paths are joined and constitute the traces of order $k+1$.
+It could happen that for some state in $N_k$ there is no path leading
+to any vertex in $D_k$: in such a case the deviation is a dead end
+that prevents the user to complete the scenario.
 
@@
 point, a user is viewing all the conversations of the inbox (state
-\texttt{viewingConversations}); available actions include moving to
+\texttt{viewing\-Conversations}); available actions include moving to
 the next or previous block of conversations, refreshing the list, or
@@
 \texttt{open(conversation)}). This transition is assumed to occur when
-the user clicks on any one of the visible conversations - it is if
+the user clicks on any one of the visible conversations - it is as if
 there is a collection of transitions, each one for a different
 conversation that can be opened; the parameter \texttt{conversation}
-is bound to any occurrence of a conversation. Such transitions lead to the state
-\texttt{viewingAConversation}, where the behavior of the system is
-defined by a more detailed state machine. By default the user is
-viewing a conversation, but by performing the \texttt{reply} action
-the UI moves to a state called \texttt{replying}, where the body of
-the reply can be typed, the subject can be changed, or another
-addressee can be added.
+is conceptually bound to any occurrence of a conversation. Such
+transitions lead to the state \texttt{viewingAConversation}, where the
+behavior of the system is defined by a more detailed state machine. By
+default the user is viewing a conversation, but by performing the
+\texttt{reply} action the UI moves to a state called
+\texttt{replying}, where the body of the reply can be typed, the
+subject can be changed, or another addressee can be added.
 
@@
 state, is specified with the following 11 bridge sets:
-\newpage
 \lstset{language=R}
@@
 it. Notice that because the model precludes the possibility that there
-are two \emph{consecutive} actions called \texttt{open(conversation)},
+are two \emph{consecutive} actions called \texttt{open(conver\-sation)},
 MIGTool has to find traces such that between bridge set 1 and 2 there
@@
 
-One of the order 0 geodesic paths is shown below; each line
-represents an action that the user is expected to perform; actions are
-only those entailed by the model. Actions marked with ``*'' are those not
-specified in the bridge sets and found in the interaction graph. 
+One of the order 0 geodesic paths is shown below; each line represents
+an action that the user is expected to perform, which is entailed
+by the model. Actions marked with ``*'' are those not specified in the
+bridge sets and found in the interaction graph.
 \lstset{language=Java}
@@
 \end{lstlisting}
-% \begin{tabular}{rp{7cm}}
-%     1 & \texttt{open(conversation)} \\
-%     2 & \texttt{goBack} \\ 
-%     3 & \texttt{open(conversation)} \\
-%     4 & \texttt{reply} \\
-%     5 & \texttt{reply, addAddressee, typeBody, changeSubject} \\
-%     6 & \texttt{cancel, send, open(message)[msgIsClosed], close (message) [msgIsOpen]}\\
-%    7 & \texttt{compose} \\
-%     8 & \texttt{attachFiles} \\
-%     9 & \texttt{addRecipient[first]} \\
-%     10  & \texttt{open(files)} \\
-%     11 & \texttt{writeSubject, writeBody} \\
-%     12 & \texttt{writeSubject, close(message)[msgIsOpen], writeBody,
-%          open(message)[msgIsClosed]} \\
-%     13 & \texttt{cancel, send} \\
-%   \end{tabular}
-%   %
 
@@
 
-Gmail has the highest number of order 0 and 1 traces; it has the
-highest difference between the number of order 0 and 1, and between
-order 1 and 2 (at least 3 times as many order 0 paths than any of the
-other applications, and at least twice as many order 1 paths as the
-any of the other applications).  In other words, Gmail offers 3 times
-as many error-free alternative paths to accomplish the scenario, which
-indicates that users might more easily follow one of those paths, than
-when using other applications.  Notice that Roundcube has the smallest number
-of order 0 paths (2 of them), which means that users are not given
-much flexibility and freedom in carrying out correctly the scenario.
-However, Gmail provides also almost twice as many order 1 paths, which
-means that users could be more easily induced into an erroneous path
-than when using another system (or given more flexibility).
-Because the number of order 2 or 3 paths decreases, Gmail reduces
-therefore the ``error proneness'' of this UI, for executions that
-involve 2 or 3 errors.
+Compared to the other applications, Gmail has the highest number of
+order 0 and 1 traces; it has the highest difference between the number
+of order 0 and 1, and between order 1 and 2 traces (at least 3 times
+as many order 0 paths than any of the other applications, and at least
+twice as many order 1 paths as the any of the other applications).  In
+other words, Gmail offers 3 times as many error-free alternative paths
+to accomplish the scenario, which indicates that users might more
+easily follow one of those paths, than when using other applications.
+Notice that Roundcube has the smallest number of order 0 paths (2 of
+them), which means that users are not given much flexibility and
+freedom in carrying out correctly the scenario.  However, Gmail
+provides also almost twice as many order 1 paths, which means that
+users could be more easily induced into an erroneous path than when
+using another system (or, as an alternative interpretation, users
+could be given more flexibility).  Because the number of order 2 or 3
+paths decreases, Gmail reduces therefore the ``error proneness'' of
+this UI, for executions that involve 2 or 3 errors.
 
@@
 
-Manual inspection of the shortest paths indicates that one reason of
-the greated potential efficiency offered by Gmail is due to the
+Manual inspection of the shortest paths indicates that one reason for
+the greater potential efficiency offered by Gmail is due to the
 fact that users can start composing a new message while reading a
@@
 Figure~\ref{fig:ad} shows the \emph{action density} of the four
-applications, \ie\ the average number of actions per state involved
-in traces defined by the out-degree, and average number of \emph{unique} actions per state. The
-former is an overall measure of the number of actions that are made
-available by a UI, the latter can be used to analyze how many
-\emph{new} options the user is presented with in any state. For our
-examples, the values are all in the range between 5.9 actions/state in
-the case of Gmail and 7.1 for Horde, and 1.3 unique actions for
-SquirrelMail and 2 for Horde. This suggests that Gmail features a more
-compact design (fewer actions to do the same things), and SquirrelMail
-is even more compact when it comes to the different types of actions;
-thus it could be easier to learn.
+applications, \ie\ the average number of actions per state involved in
+traces (defined by the out-degree), and average number of
+\emph{unique} actions per state. The former is an overall measure of
+the number of actions that are made available by a UI, the latter can
+be used to analyze how many \emph{new} options the user is presented
+with in any state. For our examples, the values are all in the range
+between 5.9 actions/state in the case of Gmail and 7.1 for Horde, and
+1.3 unique actions for SquirrelMail and 2 for Horde. This suggests
+that Gmail features a more compact design (fewer actions to do the
+same things), and SquirrelMail is even more compact when it comes to
+the different types of actions; thus it could be easier to learn.
 
@@
 scenarios. For example, a designer might be interested in a comparison
-of replying to a message as opposed to composing a new one.
-
-Let's focus now on Horde. It turns out that composing a new message
-entails many more order 0 paths (154 vs 86), and a comparable number
+of replying to a message (scenario ``reply'') as opposed to composing
+a new one (``compose''), given a single UI.
+
+Let's focus now on Horde. It turns out that ``compose'' entails many
+more order 0 paths than ``reply'' (154 vs 86), and a comparable number
 of higher order paths. The path length in the best case is the same
-across the two scenarios, but in the average case composing has a
-length of 7.75 steps compared to 8.5 for reply. Probability of hitting
-an optimal path is 4 times higher for compose.
-
-This means that users will have twice as much many choices between 
+across the two scenarios, but in the average case ``compose'' has a
+length of 7.75 steps compared to 8.5 for ``reply''. Probability of hitting
+an optimal path is 4 times higher for ``compose''.
+
+This means that users will have twice as many choices between 
 correct paths when composing a new message rather than simply
@@
 optimal paths, their length is 4.5 (36\% shorter than in S); both
-systems have a comparable number of detour 1 traces ($N_1$), but again
-their length in A ($L_1$) is 28\% shorter. The probability of hitting
-the optimal trace is however twice as large in S, which features also
-a more compact design.  In both cases detours are caused by the
+systems have a comparable number of detour 1 traces ($N_1$), but their
+length in A ($L_1$) is 28\% shorter. The probability of hitting the
+optimal trace is however twice as large in S (p=$11\%$), which features also a
+more compact design (AD=$1.75$).  In both cases detours are caused by the
 possibility of disengaging the system at the wrong moment.
@@
   because more possibilities are offered;
-\item system S features a more compact design, with fewer action 
+\item system S features a more compact design, with fewer actions
   to be performed at each moment.
@@
 specify behavior of UIs based on discrete events. in \migtool\ all the
-UML language~\cite{uml25} for state machine has been used, including sundry
-pseudostates. Evidence of this can also be found in recent OMG
-standards, such as IFML~\cite{ifml13}. Even though there are
-fundamental limits (inability to handle undo/redo's, because that goes
-beyond a finite state representation; inability to handle customizable
-toolbars, because that requires models that change at runtime;
-inability to handle perceptive UIs, because they are not well suited
-to be described in terms of discrete states), in many practical cases
-they can be be isolated and/or ignored~\cite{gb15:icwe}. One
-particular limitation of \migtool\ stems from the fact that it does
-not rely on an executable UML metamodel; as a consequence behaviors
-that rely on the fact that firing a transition broadcasts new events
-cannot be properly represented and analyzed.
+UML language~\cite{uml25} for state machine has been used, including
+sundry pseudostates. Even though there are fundamental limits
+(inability to handle undo/redo's, because this goes beyond a finite
+state representation; inability to handle customizable toolbars,
+because this requires models that change at runtime; inability to
+handle perceptive UIs, because they are not well suited to be
+described in terms of discrete states), in many practical cases they
+can be be isolated and/or ignored~\cite{gb15:icwe}. One particular
+limitation of \migtool\ stems from the fact that it does not rely on
+an executable UML metamodel; as a consequence behaviors that rely on
+the fact that firing a transition broadcasts new events cannot be
+properly represented and analyzed. For the same reason the meaning of
+guards and actions is not considered by \migtool, and they are treated
+simply as parts of the labels.
 
@@
 \bibitem{silva14}
-{Jos{\'e}-Luis Silva}, {Camille Fayollas}, {Arnaud Hamon}, {C{\'e}lia
-  Martiinie}, {Eric Barboni}, {and} {others}. 2014.
+{J.L. Silva}, {C. Fayollas}, {A. Hamon}, {P. Palanque}, {C. Martinie}, {and}
+  {E. Barboni}. 2014.
 \newblock \showarticletitle{Analysis of WIMP and Post WIMP Interactive Systems
